---
title: "Numerical experiments: from simulated data to model comparisons"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: hide
params:
  profile: "quick"
  study: "both"
  action: "run"
  cores: 4
  artifact_root: ""
---

This notebook demonstrates the complete numerical workflow: define the study,
generate and verify frozen data, inspect the data, fit the models, and examine
predictions and diagnostics. The same notebook handles both studies. After
installing the dependencies listed in the repository README, select a profile
above and **Knit**, or run its chunks in order.

## 1. Choose the computation you want

| `profile` | Purpose | Datasets per setting | Chains × iterations (warmup) |
|:--|:--|--:|:--|
| `quick` | Worked example, default | 1 | 1 × 120 (40) |
| `development` | All paper settings with reduced Monte Carlo effort | 3 | 4 × 1,750 (500) |
| `paper` | Configuration used in the paper | 50 | 4 × 20,000 (5,000) |

`study` is `study1`, `study2`, or `both`. The paper has two Study I settings
(within-category heterogeneity η = 0 or 1), and three Study II settings
(two or four ordinal proxies, plus measurement misspecification). Both paper
studies use 100 training and 100 test observations. The quick profile also
reduces test sizes and calibration grids; it demonstrates execution and does
not estimate the paper's Monte Carlo results. Development retains the full
scientific grid. Paper execution across both studies entails **600 EIV-GP fits**
plus ablations and 750 standalone competitor fits, before reuse of saved work.

Use `action: "run"` for the complete workflow. `plan` prints the resolved
settings without fitting; `data` stops after data generation; `fit` saves fits
without reporting; `report` regenerates summaries from saved fits. The quick
profile shows EIV-GP and its internal baselines. Development and paper runs also
fit UC-GP, LVGP, and EzGP on the same frozen datasets, then combine the results.
The checked-in HTML is a **quick worked example**, not the paper's result table.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE)
# These defaults also support running chunks interactively in a clean session.
defaults <- list(profile = "quick", study = "both", action = "run", cores = 4L,
                 artifact_root = "")
settings <- if (exists("params")) utils::modifyList(defaults, params) else defaults
profile <- match.arg(settings$profile, c("quick", "development", "paper"))
action <- match.arg(settings$action, c("plan", "data", "fit", "run", "report"))
studies <- if (settings$study == "both") c("study1", "study2") else
  match.arg(settings$study, c("study1", "study2"))
input <- knitr::current_input(dir = TRUE)
repository_dir <- if (length(input) == 1L && nzchar(input)) {
  normalizePath(file.path(dirname(input), ".."))
} else if (file.exists("codes/simulation_helpers.R")) normalizePath(".") else normalizePath("..")
artifact_root <- if (nzchar(settings$artifact_root)) settings$artifact_root else
  file.path(repository_dir, "reproduction")
source(file.path(repository_dir, "codes", "reporting", "notebook_helpers.R"), local = TRUE)
source(file.path(repository_dir, "codes", "core", "00_diagnostics.R"), local = TRUE)
source(file.path(repository_dir, "codes", "simulation_helpers.R"), local = TRUE)
source(file.path(repository_dir, "codes", "simulations", "notebook_workflow.R"), local = TRUE)
source(file.path(repository_dir, "codes", "simulations", "competitor_workflow.R"), local = TRUE)
source(file.path(repository_dir, "codes", "reporting", "experiment_reporting.R"), local = TRUE)
source(file.path(repository_dir, "codes", "reporting", "combine_experiment_results.R"), local = TRUE)
configs <- setNames(lapply(studies, replication_config, profile = profile,
  core_budget = as.numeric(settings$cores), repository_dir = repository_dir,
  artifact_root = artifact_root), studies)
```

```{r selected-profile, echo=FALSE, results="asis"}
for (config in configs) {
  print(knitr::kable(data.frame(study = config$study, profile = profile, action = action,
    chains = config$mcmc$n_chains, iterations = config$mcmc$n_iter, warmup = config$mcmc$burn)))
  print(knitr::kable(mixedgp_cell_summary(config)[, c("cell_id", "n", "n_test", "n_rep", "calibration_grid")]))
}
```

## 2. Freeze and inspect the datasets

A seed identifies each dataset. Calibration subsets are nested and shared across
methods. Existing files are verified against their manifests, so rerunning does
not silently replace a dataset. Quick, development, and paper artifacts use
separate locations. The figures below use the prespecified first replication
and show the observed data, rather than fitted GP estimates.

```{r data, fig.width=10, fig.height=7}
for (config in configs) {
  if (action == "plan") {
    preview <- config; preview$mode <- "dry_run"; preview$stages <- character()
    invisible(notebook_logged(file.path(artifact_root, "logs", paste0(config$study, "-plan.log")), mixedgp_run_simulation(preview)))
  } else if (action != "report") {
    data_config <- config; data_config$stages <- "data"
    frozen <- notebook_logged(file.path(artifact_root, "logs", paste0(config$study, "-data.log")),
      mixedgp_run_simulation(data_config))
    example <- mixedgp_read_cell_replication(config, config$cells[[1L]], 1L)$data$train
    observed <- data.frame(y = example$y,
      x = if (config$study == "study1") example$x else example$X[, 1L],
      category = factor(if (config$study == "study1") example$c else example$C[, 1L]))
    print(ggplot2::ggplot(observed, ggplot2::aes(x, y, color = category)) +
      ggplot2::geom_point(alpha = .65) + ggplot2::theme_minimal() +
      ggplot2::labs(title = paste(config$study, "— first frozen training dataset"),
        x = "Observed covariate", y = "Response", color = "Ordinal category"))
    print(ggplot2::ggplot(observed, ggplot2::aes(category, y, fill = category)) +
      ggplot2::geom_boxplot() + ggplot2::theme_minimal() +
      ggplot2::theme(legend.position = "none") +
      ggplot2::labs(x = "Ordinal category", y = "Response"))
  }
}
```

## 3. Fit the models

EIV-GP learns from responses, observed covariates, ordinal proxies, and the
selected exact latent-input measurements. Every method uses the same train/test
split. Saved fits allow an interrupted run to resume; compatible completed work
is reused. Short chains can flag poor mixing: those warnings remain visible and
are not evidence for or against a method.

```{r fit, results="asis"}
results <- list()
if (action %in% c("run", "fit")) for (study in studies) {
  fit_config <- configs[[study]]; fit_config$stages <- "fit"
  results[[study]] <- notebook_logged(file.path(artifact_root, "logs", paste0(study, "-fit.log")),
    mixedgp_run_simulation(fit_config))
  gates <- results[[study]]$gates
  assessment <- aggregate(gates$pass, list(setting = gates$cell_id), function(x) paste(sum(x), "/", length(x)))
  names(assessment)[2L] <- "Passed checks"
  print(knitr::kable(assessment))
  if (any(!gates$pass)) cat("\nShort-run MCMC checks are flagged; inspect full diagnostics before interpretation.\n")
}
```

## 4. Report and compare

Reporting reads saved fits rather than rerunning MCMC. Prediction errors,
interval coverage, proper scores, imputation, and diagnostics answer different
questions and should be read together. The full profiles also run the three
published competitors independently, reuse their caches, and check dataset
identity before combining results. Outputs remain local under `reproduction/`.

```{r report, results="asis", fig.width=9, fig.height=4}
if (action %in% c("run", "report")) for (study in studies) {
  config <- configs[[study]]
  run_dir <- if (!is.null(results[[study]])) results[[study]]$run_dir else mixedgp_run_directory(config)
  reporting <- notebook_logged(file.path(artifact_root, "logs", paste0(study, "-report.log")),
    mixedgp_report_run(run_dir, "report", code_dir = config$code_dir))
  if (any(reporting$status != "success")) stop("Reporting failed; inspect the saved status table.")
  if (profile != "quick") {
    competitor_dir <- file.path(artifact_root, "competitor-reports", study, config$mode)
    competitors <- notebook_logged(file.path(artifact_root, "logs", paste0(study, "-competitors.log")),
      mixedgp_run_competitor_workflow(config, competitor_dir,
        workers = config$parallel$core_budget, action = if (action == "report") "export" else "run"))
    comparison <- mixedgp_combine_saved_runs(run_dir, competitor_dir,
      file.path(run_dir, "reporting", "combined-comparisons"))
    print(knitr::kable(comparison$summary, digits = 3))
    print(knitr::kable(as.data.frame(table(competitors$status$method, competitors$status$status))))
  }
  predictive <- read.csv(file.path(run_dir, "reporting", "combined", "predictive_metrics.csv"))
  if ("evaluation_stratum" %in% names(predictive)) predictive <- predictive[predictive$evaluation_stratum == "overall", ]
  selected <- aggregate(predictive[, c("RMSE", "MAE", "Coverage95")],
    predictive[, c("cell_id", "n_calib", "method")], mean)
  print(knitr::kable(selected, digits = 3,
    caption = paste(study, "— held-out predictive summaries")))
  print(ggplot2::ggplot(selected, ggplot2::aes(factor(n_calib), RMSE, color = method, group = method)) +
    ggplot2::geom_point(size = 2) + ggplot2::geom_line() + ggplot2::facet_wrap(~cell_id) +
    ggplot2::theme_minimal() + ggplot2::labs(x = "Exact training inputs", y = "Prediction RMSE", color = "Method"))
}
```

## 5. Reproduce the paper configuration

Select `profile: "paper"`, `study: "both"`, and `action: "run"` and rerun this
same notebook. The displayed configuration is the paper configuration confirmed
by the authors: 50 datasets per setting, four chains, 20,000 iterations per
chain, including 5,000 warmup. No automatic chain extension changes that budget.
Inspect failures, convergence diagnostics, and successful-replication counts
before comparing tables. Runtime depends on the machine and cannot be inferred
from the short HTML example. `action: "report"` avoids fitting once the required
fits and competitor caches exist.

```{r session, echo=FALSE}
knitr::kable(data.frame(R = as.character(getRversion()), eivGP = as.character(packageVersion("eivGP")), profile = profile))
```
