---
title: "Ocean silicate: learning from oxygen classes and partial measurements"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: hide
params:
  profile: "quick"
  action: "run"
  data_dir: ""
  output_dir: ""
---

Can oxygen categories and a small number of exact oxygen measurements help
predict silicate? This notebook follows that question from exploratory plots
to held-out predictions using the frozen GLODAPv2.2023 North Pacific subset.
The 200 observations represent distinct station profiles at 300–1,500 dbar,
with oxygen below 180 µmol kg⁻¹. The response is silicate; fully observed inputs
are salinity and log pressure. Oxygen is treated as partially observed, with
classes bounded at 50, 100, and 150 µmol kg⁻¹.

## 1. Select the computation

| Profile | Folds | Exact training oxygen | Chains × iterations (warmup) |
|:--|:--|:--|:--|
| `quick` (this HTML example) | first fold | 0%, 50% | 1 × 80 (30) |
| `paper` | all five | 0%, 20%, 50% | 4 × 20,000 (5,000) |

The paper workflow also runs UC-GP, LVGP, and EzGP and includes the recorded
20,000-iteration continuation step for the selected fits (fold 1/O0,
fold 3/O20, fold 4/O0). Selection was based on training-chain diagnostics.
The quick example demonstrates EIV-GP prediction and latent-input uncertainty;
its short chain cannot establish convergence or comparative performance.

The frozen public subset and train/test files are included in
`data/ocean_silicate/`. No separate preprocessing notebook is needed.
`action: "run"` generates the analysis below; `report` rereads saved results;
`plan` only displays settings. Supporting scripts are in
`codes/applications/ocean_silicate/`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE)
defaults <- list(profile = "quick", action = "run", data_dir = "", output_dir = "")
settings <- if (exists("params")) utils::modifyList(defaults, params) else defaults
profile <- match.arg(settings$profile, c("quick", "paper"))
action <- match.arg(settings$action, c("plan", "run", "report"))
input <- knitr::current_input(dir = TRUE)
repository_dir <- if (length(input) == 1L && nzchar(input)) normalizePath(file.path(dirname(input), "..")) else
  if (file.exists("codes/load_mixedgp.R")) normalizePath(".") else normalizePath("..")
source(file.path(repository_dir, "codes", "reporting", "notebook_helpers.R"), local = TRUE)
source(file.path(repository_dir, "codes", "applications", "ocean_silicate", "workflow.R"), local = TRUE)
source(file.path(repository_dir, "codes", "applications", "ocean_silicate", "reporting.R"), local = TRUE)
data_dir <- if (nzchar(settings$data_dir)) settings$data_dir else file.path(repository_dir, "data", "ocean_silicate")
output_dir <- if (nzchar(settings$output_dir)) settings$output_dir else
  file.path(repository_dir, "reproduction", "ocean", profile)
```

```{r selected-profile, echo=FALSE}
knitr::kable(data.frame(profile = profile, action = action,
  folds = if (profile == "quick") 1L else 5L,
  chains = if (profile == "quick") 1L else 4L,
  initial_iterations = if (profile == "quick") 80L else 20000L))
```

## 2. Explore the observed data

All source oxygen measurements are retained for checking imputation; exact values
are withheld according to the fixed calibration masks during fitting. Each fold
has 160 training and 40 test observations. Calibration masks are nested, and no
station profile appears in both training and test data in the same fold.

```{r data, eval=action != "plan"}
cohort <- ocean_read_inputs(data_dir)
knitr::kable(as.data.frame(table(cohort$oxygen_class)),
  col.names = c("Oxygen class (µmol/kg)", "Observations"))
```

```{r exploratory-plots, eval=action != "plan", fig.width=10, fig.height=3.8}
library(ggplot2)
library(patchwork)
p1 <- ggplot(cohort, aes(U, Y, color = oxygen_class)) + geom_point(alpha = .65) +
  labs(x = "Measured oxygen (µmol/kg)", y = "Silicate (µmol/kg)", color = "Oxygen class") +
  theme_minimal() + theme(legend.position = "none")
p2 <- ggplot(cohort, aes(oxygen_class, Y, fill = oxygen_class)) + geom_boxplot(outlier.alpha = .4) +
  labs(x = "Oxygen class (µmol/kg)", y = "Silicate (µmol/kg)") +
  theme_minimal() + theme(legend.position = "none")
p3 <- ggplot(cohort, aes(G2salinity, Y, color = log_pressure)) + geom_point(alpha = .65) +
  labs(x = "Salinity", y = "Silicate (µmol/kg)", color = "Log pressure") + theme_minimal()
p1 + p2 + p3
```

These plots describe the frozen subset. They do not select calibration levels,
folds, or model settings based on held-out performance.

## 3. Fit and make held-out predictions

The class-only prediction task uses salinity, log pressure, and oxygen class.
A second, mixed-information task additionally reveals the prespecified fraction
of exact test oxygen values to EIV-GP. These tasks must be distinguished when
interpreting scores. At zero calibration, latent oxygen is identified on the
model scale; it is not scored as raw-concentration recovery.

```{r fit, eval=action == "run"}
run <- notebook_logged(paste0(output_dir, "-execution.log"),
  ocean_run_analysis(repository_dir, data_dir, output_dir, profile))
```

## 4. Inspect predictions and uncertainty

```{r results, eval=action != "plan", echo=FALSE, results="asis"}
result <- if (profile == "paper") ocean_read_paper_results(output_dir) else ocean_read_results(output_dir, profile)
shown <- result$metrics[, c("cell", "method", "RMSE", "MAE", "CRPS", "Coverage95")]
summary <- aggregate(shown[, c("RMSE", "MAE", "CRPS", "Coverage95")],
  shown[, c("cell", "method")], mean)
print(knitr::kable(summary, digits = 3))
```

```{r predictions, eval=action != "plan", fig.width=8, fig.height=3.5}
ggplot(result$predictions, aes(observed_Y, predicted_Y)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey50") +
  geom_point(alpha = .65, color = "#2c7fb8") + facet_wrap(~cell) +
  labs(x = "Observed held-out silicate", y = "Predicted silicate",
       title = "Class-only held-out predictions") + theme_minimal()
```

```{r diagnostic-note, eval=action != "plan", echo=FALSE, results="asis"}
if (profile == "quick") cat("**Worked example only. One short chain cannot establish convergence; these scores are not paper results.**") else
  print(knitr::kable(result$diagnostics, digits = 3))
```

## 5. Reproduce and cite

Choose `profile: "paper"` and rerun this notebook to execute the complete
five-fold workflow. Saved compatible fits are reused; changed inputs/settings
require a separate output directory. Full diagnostic and execution logs stay
in the local result directory, keeping this HTML focused on the analysis.

Data: GLODAPv2.2023 merged and adjusted product, DOI **10.25921/zyrq-ht66**.
Product reference: Lauvset et al. (2024), *Earth System Science Data* 16,
2047–2072, DOI **10.5194/essd-16-2047-2024**. See the bundled data README for
subset provenance and fair-data-use attribution.

```{r reproducibility, echo=FALSE}
knitr::kable(data.frame(R = as.character(getRversion()), eivGP = as.character(packageVersion("eivGP")), profile = profile))
```
