# Publication simulation API

The current publication workflow has one configuration constructor and one
runner per study. Configurations freeze design cells, seeds, eligible methods,
MCMC controls, checksums, and diagnostic gates. Both runners default to a
read-only dry run. Smoke and publication fits automatically run the exact
published-competitor and paired-design validators before data generation or
fitting; publication mode requires every competitor validation to succeed.

```{r copy-publication-api, include=FALSE}
repository_dir <- dirname(knitr::current_input(dir = TRUE))
source_dir <- file.path(repository_dir, "script", "R-scripts")
source_file <- file.path(source_dir, "simulation_helpers.R")
target_file <- file.path(usethis::proj_get(), "R", basename(source_file))
stopifnot(file.copy(source_file, target_file, overwrite = TRUE))
stopifnot(identical(
  unname(tools::md5sum(source_file)), unname(tools::md5sum(target_file))
))
```

The exact scripts hashed and sourced by the publication engine are installed
with the package. This preserves the same source-identity checks when `eivGP`
is called outside a development checkout. Validator outputs and their status
are written under the fingerprinted run's `config/` directory.

```{r install-publication-resources, include=FALSE}
simulation_resources <- c(
  "00_parallel_utils.R",
  "00_public_api.R",
  "00_study1_functions.R",
  "00_study2_functions.R",
  "03_study2_published_competitors.R",
  "00_synthetic_data.R",
  "04_study1_ablations.R",
  "04_study2_ablations.R",
  "08_published_competitor_validation.R",
  "09_experiment_design_validation.R",
  "02_study1_monte_carlo.R",
  "02_study2_monte_carlo.R",
  "simulation_helpers.R",
  "run_study1_simulation.R",
  "run_study2_simulation.R",
  "run_study1_all.R",
  "run_study2_all.R",
  "README.md",
  "PUBLICATION_SIMULATIONS.md"
)
simulation_dir <- file.path(
  usethis::proj_get(), "inst", "simulation-source"
)
dir.create(simulation_dir, recursive = TRUE, showWarnings = FALSE)
copied <- file.copy(
  file.path(source_dir, simulation_resources),
  file.path(simulation_dir, simulation_resources),
  overwrite = TRUE
)
stopifnot(all(copied))
stopifnot(identical(
  unname(tools::md5sum(file.path(source_dir, simulation_resources))),
  unname(tools::md5sum(file.path(simulation_dir, simulation_resources)))
))
```

```{r publication-api-example, eval=FALSE}
study1 <- eivGP::run_study1_simulation(
  eivGP::study1_simulation_config(mode = "publication", workers = 8L)
)
study2 <- eivGP::run_study2_simulation(
  eivGP::study2_simulation_config(mode = "publication", workers = 8L)
)
```
