# Reproduction workflows

The current publication workflow has one configuration constructor and one
runner per study. Configurations freeze design cells, seeds, eligible methods,
MCMC controls, checksums, and diagnostic gates. Both runners default to a
read-only dry run. Smoke and publication fits automatically run the exact
published-competitor and paired-design validators before data generation or
fitting. External competitor failures are retained as auditable status records
without preventing EIV-GP MCMC. The EIV-GP convergence gate is always computed
and archived, but a fit that reaches its finite iteration cap below target is a
warning rather than a reason to discard all later frozen replications or cells.
Actual execution errors still stop the affected cell immediately.

```{r copy-publication-api, include=FALSE}
repository_dir <- dirname(knitr::current_input(dir = TRUE))
source_dir <- file.path(
  repository_dir, "script", "newly-written-and-modified-scripts"
)
source_file <- file.path(source_dir, "reproduction_workflows.R")
target_file <- file.path(usethis::proj_get(), "R", basename(source_file))
stopifnot(file.copy(source_file, target_file, overwrite = TRUE))
stopifnot(identical(
  unname(tools::md5sum(source_file)), unname(tools::md5sum(target_file))
))
```

The exact scripts hashed and sourced by the publication engine are installed
with the package. This preserves the same source-identity checks when `eivGP`
is called outside a development checkout. Validator outputs and their status
are written under the fingerprinted run's `config/` directory.

```{r install-publication-resources, include=FALSE}
simulation_resources <- c(
  "core_parallel.R",
  "core_numerics.R",
  "model_api.R",
  "model_univariate.R",
  "model_multivariate.R",
  "competitors.R",
  "reproduction_data.R",
  "04_study1_ablations.R",
  "04_study2_ablations.R",
  "08_published_competitor_validation.R",
  "09_experiment_design_validation.R",
  "02_study1_monte_carlo.R",
  "02_study2_monte_carlo.R",
  "reproduction_workflows.R",
  "run_study1_simulation.R",
  "run_study2_simulation.R",
  "run_study1_all.R",
  "run_study2_all.R",
  "README.md",
  "PUBLICATION_SIMULATIONS.md"
)
simulation_dir <- file.path(
  usethis::proj_get(), "inst", "simulation-source"
)
dir.create(simulation_dir, recursive = TRUE, showWarnings = FALSE)
copied <- file.copy(
  file.path(source_dir, simulation_resources),
  file.path(simulation_dir, simulation_resources),
  overwrite = TRUE
)
stopifnot(all(copied))
stopifnot(identical(
  unname(tools::md5sum(file.path(source_dir, simulation_resources))),
  unname(tools::md5sum(file.path(simulation_dir, simulation_resources)))
))
```

```{r publication-api-example, eval=FALSE}
study1 <- eivGP::run_study1_simulation(
  eivGP::study1_simulation_config(mode = "publication", workers = 8L)
)
study2 <- eivGP::run_study2_simulation(
  eivGP::study2_simulation_config(mode = "publication", workers = 8L)
)
```
