---
title: "Real-data application"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  execute: false
  data_file: ""
  engine: "multivariate"
  kernel: "matern"
  seed: 20260902
---

# Status and data contract

This top-to-bottom application document is ready for the final real dataset,
which is not committed yet. Its default render records the required contract
without fitting. When the audited data bundle is ready, set `execute: true`
and provide `data_file` as an RDS file containing at least `X`, `y`, and `C`.

The bundle may also contain `U_obs`, `latent_dim`, `ordinal_levels`,
`fit_args`, `new_X`, `new_C`, and `new_U`. This keeps the final analysis data
and its explicit model configuration together without embedding a private
filesystem path in the package.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE,
  error = FALSE, cache = FALSE
)
document_file <- knitr::current_input(dir = TRUE)
repository_dir <- normalizePath(
  file.path(dirname(document_file), ".."), winslash = "/", mustWork = TRUE
)
package_dir <- file.path(repository_dir, "eivGP")
if (!requireNamespace("pkgload", quietly = TRUE)) {
  stop("Install pkgload to run the development version of eivGP.")
}
pkgload::load_all(package_dir, quiet = TRUE)
EXECUTE <- isTRUE(params$execute)
```

# Load and validate

```{r load-data}
APPLICATION_DATA <- NULL
if (!EXECUTE) {
  message("Dry application render: no real-data fit was requested.")
} else {
  if (!nzchar(params$data_file)) {
    stop("Set data_file when execute is true.")
  }
  data_path <- normalizePath(params$data_file, winslash = "/", mustWork = TRUE)
  APPLICATION_DATA <- readRDS(data_path)
  required <- c("X", "y", "C")
  missing <- setdiff(required, names(APPLICATION_DATA))
  if (length(missing) > 0L) {
    stop("The data bundle is missing: ", paste(missing, collapse = ", "))
  }
}
```

# Fit

All additional sampler settings must be named in `fit_args` inside the audited
data bundle. The visible call below is the only model-fitting entry point.

```{r fit-model}
APPLICATION_FIT <- NULL
if (EXECUTE) {
  fit_args <- APPLICATION_DATA$fit_args
  if (is.null(fit_args)) fit_args <- list()
  if (!is.list(fit_args) || (length(fit_args) > 0L &&
      (is.null(names(fit_args)) || any(!nzchar(names(fit_args)))))) {
    stop("fit_args must be a named list.")
  }
  common <- list(
    X = APPLICATION_DATA$X,
    y = APPLICATION_DATA$y,
    C = APPLICATION_DATA$C,
    U_obs = APPLICATION_DATA$U_obs,
    latent_dim = APPLICATION_DATA$latent_dim,
    ordinal_levels = APPLICATION_DATA$ordinal_levels,
    engine = as.character(params$engine),
    kernel = as.character(params$kernel),
    seed = as.integer(params$seed)
  )
  common <- common[!vapply(common, is.null, logical(1L))]
  duplicate <- intersect(names(common), names(fit_args))
  if (length(duplicate) > 0L) {
    stop("fit_args duplicates: ", paste(duplicate, collapse = ", "))
  }
  APPLICATION_FIT <- do.call(eivGP::fit_eivgp, c(common, fit_args))
  print(summary(APPLICATION_FIT))
}
```

# Prediction targets

```{r predictions}
APPLICATION_PREDICTIONS <- list()
if (EXECUTE && !is.null(APPLICATION_DATA$new_X) &&
    !is.null(APPLICATION_DATA$new_C)) {
  APPLICATION_PREDICTIONS$mean <- eivGP::predict_eivgp(
    APPLICATION_FIT,
    new_X = APPLICATION_DATA$new_X,
    new_C = APPLICATION_DATA$new_C,
    target = "mean",
    seed = as.integer(params$seed) + 1L
  )
  APPLICATION_PREDICTIONS$response <- eivGP::predict_eivgp(
    APPLICATION_FIT,
    new_X = APPLICATION_DATA$new_X,
    new_C = APPLICATION_DATA$new_C,
    target = "response",
    seed = as.integer(params$seed) + 2L
  )
}
if (EXECUTE && !is.null(APPLICATION_DATA$new_X) &&
    !is.null(APPLICATION_DATA$new_U)) {
  APPLICATION_PREDICTIONS$surface <- eivGP::predict_eivgp(
    APPLICATION_FIT,
    new_X = APPLICATION_DATA$new_X,
    new_U = APPLICATION_DATA$new_U,
    target = "surface",
    seed = as.integer(params$seed) + 3L
  )
}
lapply(APPLICATION_PREDICTIONS, dim)
```

# Session information

```{r session-information}
sessionInfo()
```
