---
title: "Creating the eivmixgp R package"
author: "EIV-GP authors"
date: "Package-development draft"
knit: litr::render
output: litr::litr_html_document
params:
  package_name: "eivmixgp"
  package_parent_dir: "../package-build"
---

<!--
This is the literate package-build source. During the manuscript revision the
canonical implementations remain in ../codes. Rendering copies those audited
modules into the generated package, so fixes are made once rather than in a
separate package fork.
-->

## Package setup

```{r package-setup, message=FALSE, results='hide'}
usethis::create_package(
  path = ".",
  fields = list(
    Package = params$package_name,
    Version = "0.2.1",
    Title = "Gaussian Process Regression with Ordinal Mixed Inputs",
    Description = paste(
      "Fits errors-in-variables Gaussian process regressions for ordinal",
      "mixed inputs, with sparse calibration measurements and audited",
      "wrappers for published mixed-input GP competitors."
    ),
    `Authors@R` = c(
      person(given = "Penghui", family = "Fu", role = "aut"),
      person(given = "Xinye", family = "Chen", role = "aut"),
      person(given = "Jianhua Z.", family = "Huang", role = "aut"),
      person(
        given = "Sheng", family = "Jiang",
        email = "jiangsheng@cuhk.edu.cn", role = c("aut", "cre")
      )
    ),
    License = "MIT + file LICENSE",
    Encoding = "UTF-8",
    Depends = "R (>= 4.1.0)",
    Imports = paste(
      "methods, parallel, posterior, stats, tools, TruncatedNormal, utils"
    ),
    Suggests = paste(
      "dplyr, EzGP, ggplot2, kergp, knitr, LVGP, patchwork,",
      "testthat, tidyr"
    )
  )
)
writeLines(
  c(
    "YEAR: 2026",
    paste(
      "COPYRIGHT HOLDER: Penghui Fu, Xinye Chen, Jianhua Z. Huang,",
      "and Sheng Jiang"
    )
  ),
  file.path(usethis::proj_get(), "LICENSE")
)

writeLines(
  c(
    "# eivmixgp",
    "",
    "`eivmixgp` fits Gaussian process regressions with numeric",
    "predictors and ordinal proxies for latent continuous inputs. Sparse",
    "calibration measurements of the latent inputs can be supplied without",
    "discarding the larger sample observed only through the ordinal proxies.",
    "",
    "## Installation",
    "",
    "Build and install the package from the generated source directory:",
    "",
    "```sh",
    "R CMD build eivmixgp",
    "R CMD INSTALL eivmixgp_0.2.1.tar.gz",
    "```",
    "",
    "## Statistical targets",
    "",
    "The fitted posterior supports four related tasks:",
    "",
    "- inference on the observed-input mean `m(x,c) = E{f(x,U) | C=c}`;",
    "- inference on the latent response surface `f(x,u)`;",
    "- prediction of a future response given `(x,c)` or an exact new `u`; and",
    "- posterior imputation of latent inputs for training or prospective units.",
    "",
    "## Stable interface",
    "",
    "```r",
    "fit <- fit_eivgp(",
    "  X, y, C, U_obs,",
    "  engine = \"multivariate\",",
    "  latent_dim = 2,",
    "  ident = \"none\", # unrestricted loadings; requires some calibration",
    "  kernel = \"matern\",",
    "  n_chains = 4",
    ")",
    "",
    "summary(fit)",
    "",
    "m_draws <- predict(",
    "  fit, new_X = X_new, new_C = C_new, target = \"mean\"",
    ")",
    "y_draws <- predict(",
    "  fit, new_X = X_new, new_C = C_new, target = \"response\"",
    ")",
    "f_draws <- predict(",
    "  fit, new_X = X_grid, new_U = U_grid, target = \"surface\"",
    ")",
    "u_draws <- impute_eivgp(fit, new_C = C_new)",
    "```",
    "",
    "A small self-contained univariate smoke test is below. Its one short chain",
    "is for checking installation only, not for scientific inference:",
    "",
    "```r",
    "set.seed(1)",
    "n <- 24L",
    "u <- seq(-2, 2, length.out = n)",
    "X <- cbind(exposure = stats::runif(n, -1, 1))",
    "C <- ordered(",
    "  cut(u, c(-Inf, -0.5, 0.6, Inf), labels = c('low', 'mid', 'high')),",
    "  levels = c('low', 'mid', 'high')",
    ")",
    "y <- sin(X[, 1]) + 0.7 * tanh(u) + stats::rnorm(n, sd = 0.1)",
    "U_obs <- rep(NA_real_, n)",
    "U_obs[c(2L, 7L, 13L, 19L, 23L)] <- u[c(2L, 7L, 13L, 19L, 23L)]",
    "fit <- fit_eivgp(",
    "  X, y, C, U_obs, engine = 'univariate', parallel = FALSE,",
    "  n_iter = 300L, burn = 150L, n_chains = 1L, # smoke test only",
    "  preset = 'fast', seed = 11L, verbose = FALSE",
    ")",
    "new_X <- cbind(exposure = c(-0.5, 0.5))",
    "new_C <- ordered(c('low', 'high'), levels = levels(C))",
    "m_draws <- predict(",
    "  fit, new_X = new_X, new_C = new_C, target = 'mean',",
    "  n_latent = 64L, seed = 12L",
    ")",
    "dim(m_draws)",
    "```",
    "",
    "Select `engine = \"univariate\"` for one deterministically thresholded",
    "latent input and `engine = \"multivariate\"` for the noisy ordinal-probit",
    "measurement model. The multivariate latent dimension must be specified",
    "when no calibration measurements are available; it is never guessed.",
    "",
    "## Diagnose, extend, and recheck",
    "",
    "Use at least four independently initialized chains for scientific work.",
    "Every post-warm-up draw is kept; the public API rejects `thin > 1`.",
    "Choose panel locations covering the scientific comparisons you will report:",
    "",
    "```r",
    "report <- diagnose_eivgp(fit, X = X_new, C = C_new)",
    "report",
    "report$table # R-hat, bulk/tail ESS, mean MCSE, and per-functional status",
    "saveRDS(fit, 'fit-checkpoint.rds')",
    "# After reviewing diagnostics, if additional sampling is appropriate:",
    "fit <- continue_eivgp(readRDS('fit-checkpoint.rds'), n_iter = 2000L)",
    "report <- diagnose_eivgp(fit, X = X_new, C = C_new)",
    "```",
    "",
    "`n_iter` in `continue_eivgp()` means additional transitions per chain.",
    "It resumes terminal states and RNG streams, freezes warm-up tuning,",
    "and appends draws within each chain. It does not restart or repeat warm-up.",
    "Keep the same package version and numerical environment when resuming.",
    "Old fits without checkpoints must be refitted. This is not periodic crash",
    "recovery while a fitting call is still running.",
    "",
    "Poor exploration requires investigation, not an automatic longer run.",
    "Agreement with low ESS or excessive MCSE suggests extending and rechecking.",
    "A diagnostic pass is only a screen for the monitored functionals, not proof",
    "of convergence, identification, or model adequacy. Supply `U` with the panel",
    "to additionally check calibrated `f(x,u)` summaries, and `additional_series`",
    "for scientific contrasts aligned with every retained draw in every chain.",
    "Check sensitivity to `n_latent` and the integration seed: integration error",
    "is not included in reported MCMC error. Quantiles need separate MCSE checks.",
    "",
    "## Calibration and latent scale",
    "",
    "Raw or physical latent coordinates are reported only when calibrated",
    "latent inputs have affine rank `d + 1`. This is a conservative software",
    "reporting rule, not a necessary-condition theorem for identification.",
    "Otherwise outputs use model coordinates. Without calibration, inference on",
    "`m(x,c)` and response prediction given `(x,c)` remain model-defined, while",
    "`f(x,u)` and latent-input imputations must be interpreted on the working",
    "scale. `summary(fit)` reports the calibration count, affine rank, and active",
    "scale; requesting `scale = \"raw\"` when the scale is unanchored is an error.",
    "",
    "Both engines currently fix the latent population law to `N_d(0,I_d)`,",
    "independent of X in model coordinates. Calibration standardization is",
    "on by default in the public interface and uses calibrated training rows;",
    "these plug-in centers/scales have no posterior uncertainty, and scaling",
    "does not remove correlation. Inspect `fit$model_specification`.",
    "",
    "With sparse/rank-deficient calibration, select the loading structure",
    "explicitly: `ident = \"none\"` leaves loadings unrestricted, whereas",
    "`ident = \"lower_triangular\"` imposes structural zeros and positive",
    "diagonal entries. Triangular loadings with diagonal ARD are not merely",
    "a coordinate convention. A richer `U | X` population model is not yet",
    "implemented; assess these assumptions before physical-scale inference.",
    "",
    "## Prospective latent simulation",
    "",
    "For the multivariate engine, a minimax-tilted accept--reject sampler for",
    "`U | C` is the default (`latent_sampler = \"minimax_tilting\"`) and is exact",
    "on successful completion. Exact prior rejection",
    "is available as `\"rejection\"`; finite Gibbs sampling is retained only as",
    "an explicitly requested diagnostic approximation. The default uses",
    "`TruncatedNormal::rtmvnorm`.",
    "",
    "## Output contracts",
    "",
    "`predict()` and `predict_eivgp()` return a numeric matrix whose columns",
    "follow the evaluation rows and whose rows follow posterior `draw_ids`, with",
    "the `n_per_draw` replicates for each posterior state contiguous. The `joint`",
    "attribute distinguishes coherent multivariate draws from pointwise draws;",
    "use `joint = TRUE` for simultaneous functionals. `impute_eivgp()` returns a",
    "posterior-draw by unit by latent-coordinate array and labels the source,",
    "coordinate scale, and calibration anchoring in attributes.",
    "",
    "## Reproducible numerical experiments",
    "",
    "Both simulation studies use the same validated runner:",
    "",
    "```r",
    "study1 <- run_mixedgp_experiment(",
    "  \"study1\", config = \"quick\", parallel_level = \"chains\"",
    ")",
    "study2 <- run_mixedgp_experiment(",
    "  \"study2\", config = \"quick\", parallel_level = \"chains\"",
    ")",
    "```",
    "",
    "Use `config = \"balanced\"` or `\"thorough\"` for publication-scale runs.",
    "Frozen synthetic datasets are collision-checked, and seeded serial and fork",
    "backends preserve the caller's random-number stream. Previously archived",
    "Study I fits are reused only through an explicit `study_options` opt-in.",
    "Published competitor packages and reporting packages are optional",
    "dependencies; the experiment preflight records availability and versions.",
    "",
    "## Citation",
    "",
    "Run `citation('eivmixgp')` for the manuscript citation bundled with the",
    "package."
  ),
  file.path(usethis::proj_get(), "README.md")
)

writeLines(
  c(
    "# eivmixgp 0.2.1",
    "",
    "- Added eivgp_run_settings(core_budget, pending_datasets) for explicit",
    "  dataset/chain allocation within a CPU budget on macOS/Linux.",
    "- Development defaults: four chains, 500 warmup + 1250 sampling",
    "  transitions per chain; retain all 5000 posterior draws.",
    "- Both study drivers support core-budgeted nested parallelism.",
    "- Continuation remains explicit via continue_eivgp(), not automatic.",
    "- The complete development orchestration brief is not yet implemented;",
    "  publication gate policy and existing cache granularity are unchanged.",
    "",
    "# eivmixgp 0.2.0",
    "",
    "- Added diagnose_eivgp(): full-window, target-aware R-hat, bulk/tail ESS,",
    "  and estimator-specific mean MCSE with explicit failure classifications.",
    "- Added continue_eivgp(): serializable terminal-state checkpoints, per-chain",
    "  RNG restoration, frozen warm-up tuning, cumulative iteration schedules,",
    "  and recomputed diagnostics after appending every new draw.",
    "- Public fits now require no thinning (thin = 1). Multivariate defaults",
    "  and preset retention settings also keep every post-warm-up draw.",
    "- Diagnostic computation is shared by the package and simulation helpers.",
    "- Added split/uninterrupted, serial/parallel, save/reload, repeated-resume,",
    "  invalid-checkpoint and scientific-functional diagnostic regression tests.",
    "",
    "# eivmixgp 0.1.0",
    "",
    "- Added the stable `fit_eivgp()`, `predict_eivgp()`/`predict()`,",
    "  `impute_eivgp()`, and `run_mixedgp_experiment()` interfaces.",
    "- Added concise `print()` and calibration-aware `summary()` methods for",
    "  fitted EIV-GP models.",
    "- Made minimax-tilted accept--reject simulation, exact on successful",
    "  completion, the default for prospective multivariate latent inputs; exact",
    "  prior rejection and diagnostic finite Gibbs remain available explicitly.",
    "- Required explicit multivariate latent dimension when calibration cannot",
    "  supply it, and restricted raw-scale output to full-affine-rank calibration.",
    "- Added squared-exponential and Matern kernels, multivariate numeric inputs,",
    "  labeled ordinal-factor mappings, and exact-new-latent response prediction.",
    "- Unified both numerical studies behind one runner with frozen-data collision",
    "  checks, reproducible serial/fork RNG, audited competitor wrappers, and",
    "  explicit sampler safety failures.",
    "- Added generator/design fingerprints, deep truth validation, and atomic",
    "  incremental manifests for frozen synthetic datasets.",
    "- Added strict public count, logical, name, identification, and argument",
    "  validation while preserving caller random-number streams.",
    "- Kept plotting, table, and published-competitor packages optional so loading",
    "  `eivmixgp` does not attach packages, modify the ggplot theme, or set thread",
    "  environment variables."
  ),
  file.path(usethis::proj_get(), "NEWS.md")
)

dir.create(file.path(usethis::proj_get(), "inst"), showWarnings = FALSE)
writeLines(
  c(
    "bibentry(",
    "  bibtype = 'Unpublished',",
    "  title = 'Ordinal Mixed-Input Gaussian Process Regression',",
    "  author = c(",
    "    person('Penghui', 'Fu'), person('Xinye', 'Chen'),",
    "    person('Jianhua Z.', 'Huang'), person('Sheng', 'Jiang')",
    "  ),",
    "  year = '2026',",
    "  note = 'Manuscript in preparation'",
    ")"
  ),
  file.path(usethis::proj_get(), "inst", "CITATION")
)
```

```{package_doc package-documentation}
#' eivmixgp: Gaussian process regression with ordinal mixed inputs
#'
#' @import stats
#' @importFrom utils combn flush.console head tail
```

```{r global-variables, send_to="R/global-variables.R", eval=FALSE}
utils::globalVariables(c(
  "block", "chain", "ess", "global_u_accept_total", "global_u_eval_total",
  "global_u_total", "ref", "rhat", "theta_eval_total",
  "theta_update_total", "u_ess_accept_total", "u_ess_eval_total",
  "u_ess_total", "value", "xintercept"
))
```

## Canonical computational modules

The manuscript code is already organized into modules with a narrow public
interface. We copy those exact files into the package rather than maintaining a
second implementation.

```{r copy-canonical-modules, message=FALSE}
source_document <- knitr::current_input(dir = TRUE)
canonical_dir <- normalizePath(
  file.path(dirname(source_document), "..", "codes"),
  winslash = "/",
  mustWork = TRUE
)
package_dir <- usethis::proj_get()
build_ignore <- file.path(package_dir, ".Rbuildignore")
if (!file.exists(build_ignore)) file.create(build_ignore)
canonical_modules <- c(
  "00_parallel_utils.R",
  "00_study1_functions.R",
  "00_study2_functions.R",
  "03_study2_published_competitors.R",
  "00_synthetic_data.R",
  "00_public_api.R",
  "00_diagnostics.R",
  "00_mcmc_workflow.R",
  "00_experiment_runner.R"
)
dir.create(file.path(package_dir, "R"), showWarnings = FALSE)
copied <- file.copy(
  file.path(canonical_dir, canonical_modules),
  file.path(package_dir, "R", canonical_modules),
  overwrite = TRUE
)
stopifnot(all(copied))

experiment_scripts <- c(
  "simulation_helpers.R",
  "01_study1_representative_figures.R",
  "02_study1_monte_carlo.R",
  "02_study1_targeted_controls.R",
  "04_study1_ablations.R",
  "01_study2_representative_figures.R",
  "02_study2_monte_carlo.R",
  "04_study2_ablations.R"
)
experiment_dir <- file.path(package_dir, "inst", "experiments")
dir.create(experiment_dir, recursive = TRUE, showWarnings = FALSE)
copied_experiments <- file.copy(
  file.path(canonical_dir, experiment_scripts),
  file.path(experiment_dir, experiment_scripts),
  overwrite = TRUE
)
stopifnot(all(copied_experiments))
dir.create(file.path(package_dir, "tests", "testthat"), recursive = TRUE, showWarnings = FALSE)
writeLines(c("library(testthat)", "library(eivmixgp)", "test_check('eivmixgp')"),
           file.path(package_dir, "tests", "testthat.R"))
stopifnot(file.copy(file.path(canonical_dir, "tests", "test_public_mcmc_workflow.R"),
                   file.path(package_dir, "tests", "testthat", "test-mcmc-workflow.R"),
                   overwrite = TRUE))
devtools::load_all(package_dir, quiet = TRUE)
```

## Public interface

The package-facing calls are:

```{r, eval=FALSE}
fit <- fit_eivgp(
  X, y, C, U_obs,
  engine = "multivariate",
  ident = "none", # choose the loading structure explicitly with calibration
  kernel = "matern",
  parallel = TRUE,
  n_cores = 4,
  n_chains = 4
)

y_draws <- predict_eivgp(
  fit, new_X, new_C,
  target = "response",
  latent_sampler = "minimax_tilting",
  seed = 20260831
)

f_draws <- predict_eivgp(
  fit, grid_X, new_U = grid_U,
  target = "surface",
  seed = 20260831
)

uc <- fit_ucgp(X, y, C, new_X, new_C)
lv <- fit_lvgp(X, y, C, new_X, new_C)
ez <- fit_ezgp(X, y, C, new_X, new_C)

study1 <- run_mixedgp_experiment(
  "study1", config = "quick", parallel_level = "chains"
)
study2 <- run_mixedgp_experiment(
  "study2", config = "quick", parallel_level = "chains"
)
```

## Minimal package tests

```{r}
mixedgp_test_univariate_fit <- function() {
  x <- matrix(c(-1, 0, 1), ncol = 1L, dimnames = list(NULL, "x1"))
  level_maps <- list(severity = c("low", "medium", "high"))
  structure(
    list(
      data = list(
        x = x, x_raw = x, y = c(-0.5, 0, 0.6),
        x_center = 0, x_scale = 1, predictor_names = "x1",
        y_center = 0, y_scale = 1, m = 3L, C_names = "severity",
        C_level_maps = level_maps,
        U_center = 0, U_scale = 1, U_names = "latent_severity",
        calib_idx = integer(0), miss_idx = 1:3,
        latent_scale_anchored = FALSE,
        latent_anchor_rank = 0L, latent_anchor_required_rank = 2L
      ),
      mcmc = list(
        samples_u = rbind(c(-0.8, 0, 0.8), c(-0.7, 0.1, 0.9)),
        samples_tau = rbind(c(-0.25, 0.4), c(-0.2, 0.45)),
        samples_logtheta = rbind(
          log(c(1.2, 0.7, 0.9)), log(c(1.1, 0.8, 1))
        ),
        samples_sigma2 = c(0.2, 0.25)
      ),
      kernel = list(name = "se", matern_nu = NA_real_),
      diagnostics = list(summary = data.frame(
        kernel = "se", matern_nu = NA_real_, n_chains = 2L,
        parallel_backend = "serial", parallel_cores = 1L,
        n_iter = 20L, burn = 10L, thin = 5L,
        saved_per_chain = 1L, total_saved_draws = 2L,
        max_rhat_hyper = 1.01, max_rhat_tau = 1.02,
        median_rhat_missing_u = 1.00, max_rhat_missing_u = 1.01,
        min_ess_key = 75, time_seconds = 0.1
      )),
      interface = list(
        engine = "univariate", ordinal_level_maps = level_maps
      )
    ),
    class = c("eivgp_fit", "list")
  )
}

mixedgp_test_multivariate_fit <- function() {
  x <- matrix(c(-1, 0, 1), ncol = 1L, dimnames = list(NULL, "x1"))
  samples_U <- array(NA_real_, dim = c(2L, 3L, 1L))
  samples_U[1L, , 1L] <- c(-0.7, 0, 0.7)
  samples_U[2L, , 1L] <- c(-0.6, 0.1, 0.8)
  samples_A <- array(NA_real_, dim = c(2L, 2L, 1L))
  samples_A[1L, , 1L] <- c(1, 0.7)
  samples_A[2L, , 1L] <- c(1.1, 0.8)
  level_maps <- list(
    proxy_a = c("low", "medium", "high"),
    proxy_b = c("none", "mild", "severe")
  )
  structure(
    list(
      data = list(
        X = x, X_raw = x, y = c(-0.5, 0, 0.6),
        X_center = 0, X_scale = 1, y_center = 0, y_scale = 1,
        p = 1L, d = 1L, q = 2L, m_vec = c(3L, 3L),
        C_names = names(level_maps), C_level_maps = level_maps,
        C_ord = cbind(
          proxy_a = c(1L, 2L, 3L), proxy_b = c(1L, 2L, 3L)
        ),
        U_center = 0, U_scale = 1, U_names = "latent1",
        calib_idx = integer(0), miss_idx = 1:3, ident = "lower_triangular",
        latent_scale_anchored = FALSE,
        latent_anchor_rank = 0L, latent_anchor_required_rank = 2L
      ),
      mcmc = list(
        samples_U = samples_U, samples_A = samples_A,
        samples_tau = rbind(
          c(-0.5, 0.5, -0.5, 0.5),
          c(-0.4, 0.6, -0.45, 0.55)
        ),
        samples_logtheta = rbind(
          log(c(1.2, 0.7, 0.9)), log(c(1.1, 0.8, 1))
        ),
        samples_sigma2 = c(0.2, 0.25)
      ),
      kernel = list(name = "se", matern_nu = NA_real_),
      diagnostics = list(summary = data.frame(
        sampler_strategy = "blocked", kernel = "se", matern_nu = NA_real_,
        n_chains = 2L, parallel_backend = "serial", parallel_cores = 1L,
        n_iter = 20L, burn = 10L, thin = 5L,
        saved_per_chain = 1L, total_saved_draws = 2L,
        max_rhat_hyper = 1.01, max_rhat_A = 1.02,
        max_rhat_tau = 1.01, median_rhat_missing_U = 1.00,
        max_rhat_missing_U = 1.02, min_ess_key = 80,
        mean_u_ess_accept = 1, mean_global_u_accept = 1,
        time_seconds = 0.2
      )),
      interface = list(
        engine = "multivariate", ordinal_level_maps = level_maps
      )
    ),
    class = c("eivgp_fit", "list")
  )
}

testthat::test_that("parallel maps are reproducible", {
  f <- function(i) c(i = i, u = stats::runif(1))
  serial <- mixedgp_parallel_lapply(
    as.list(1:3), f, n_cores = 1, seeds = 101:103
  )
  parallel <- mixedgp_parallel_lapply(
    as.list(1:3), f, n_cores = 2, seeds = 101:103
  )
  testthat::expect_identical(serial, parallel)

  serial_m <- mixedgp_parallel_mapply(
    function(a, b) c(sum = a + b, u = stats::runif(1)),
    a = 1:3, b = 3:1, n_cores = 1, seeds = 201:203
  )
  parallel_m <- mixedgp_parallel_mapply(
    function(a, b) c(sum = a + b, u = stats::runif(1)),
    a = 1:3, b = 3:1, n_cores = 2, seeds = 201:203
  )
  testthat::expect_identical(serial_m, parallel_m)
})

testthat::test_that("seeded parallel maps preserve the caller RNG stream", {
  set.seed(4101)
  expected_lapply_continuation <- stats::runif(4)
  set.seed(4101)
  invisible(mixedgp_parallel_lapply(
    as.list(1:3), function(i) stats::runif(i),
    n_cores = 2, seeds = 501:503
  ))
  testthat::expect_identical(stats::runif(4), expected_lapply_continuation)

  set.seed(4102)
  expected_mapply_continuation <- stats::runif(4)
  set.seed(4102)
  invisible(mixedgp_parallel_mapply(
    function(a, b) c(a + b, stats::runif(1)),
    a = 1:3, b = 3:1, n_cores = 2, seeds = 601:603
  ))
  testthat::expect_identical(stats::runif(4), expected_mapply_continuation)
})

testthat::test_that("seeded public operations preserve the caller RNG stream", {
  univariate <- mixedgp_test_univariate_fit()
  new_X <- matrix(
    c(-0.25, 0.25), ncol = 1L, dimnames = list(NULL, "x1")
  )
  new_C <- data.frame(severity = ordered(
    c("high", "low"), levels = c("low", "high")
  ))

  set.seed(4110)
  expected_predict <- stats::runif(4)
  set.seed(4110)
  invisible(predict_eivgp(
    univariate, new_X, new_C, target = "mean", draw_ids = 1L,
    n_latent = 4L, include_process_uncertainty = FALSE, seed = 9001L
  ))
  testthat::expect_identical(stats::runif(4), expected_predict)

  set.seed(4111)
  expected_impute <- stats::runif(4)
  set.seed(4111)
  invisible(impute_eivgp(
    univariate, new_C = new_C, draw_ids = 1L, seed = 9002L
  ))
  testthat::expect_identical(stats::runif(4), expected_impute)

  X <- matrix(seq(-1, 1, length.out = 8L), ncol = 1L)
  u <- seq(-1.2, 1.2, length.out = 8L)
  C <- ordered(
    ifelse(u < 0, "low", "high"), levels = c("low", "high")
  )
  U_obs <- rep(NA_real_, 8L)
  U_obs[c(1L, 4L, 8L)] <- u[c(1L, 4L, 8L)]
  set.seed(4112)
  expected_fit <- stats::runif(4)
  set.seed(4112)
  invisible(fit_eivgp(
    X, sin(u), C, U_obs, engine = "univariate", parallel = FALSE,
    n_iter = 20L, burn = 10L, thin = 1L, n_chains = 1L,
    preset = "fast", seed = 9003L, verbose = FALSE
  ))
  testthat::expect_identical(stats::runif(4), expected_fit)

  adapter <- function(..., seed) {
    set.seed(seed)
    stats::runif(2L)
  }
  testthat::local_mocked_bindings(
    mixedgp_adapter_ucgp = adapter,
    mixedgp_adapter_lvgp = adapter,
    mixedgp_adapter_ezgp = adapter,
    .package = "eivmixgp"
  )
  set.seed(4113)
  expected_competitors <- stats::runif(4)
  set.seed(4113)
  invisible(fit_ucgp(1, 1, 1, 1, 1, seed = 9004L))
  invisible(fit_lvgp(1, 1, 1, 1, 1, seed = 9005L))
  invisible(fit_ezgp(1, 1, 1, 1, 1, seed = 9006L))
  testthat::expect_identical(stats::runif(4), expected_competitors)
})

testthat::test_that("public controls and input schemas fail closed", {
  X <- matrix(c(-1, 0, 1), ncol = 1L)
  y <- c(-0.5, 0, 0.5)
  C1 <- ordered(c("low", "middle", "high"))
  testthat::expect_error(
    fit_eivgp(X, y, C1, U_obs = factor(c(10, 20, 30)),
              engine = "univariate"),
    "must be numeric"
  )
  testthat::expect_error(
    fit_eivgp(X, y, C1, engine = "univariate", parallel = 1),
    "parallel must be TRUE or FALSE"
  )
  testthat::expect_error(
    fit_eivgp(X, y, C1, engine = "univariate", m_vec = 3.5),
    "integer-valued"
  )

  C2 <- data.frame(
    proxy_a = ordered(c("low", "middle", "high")),
    proxy_b = ordered(c("none", "mild", "severe"))
  )
  U_rank_deficient <- matrix(
    c(0, 0, 1, 1, NA, NA), nrow = 3L, byrow = TRUE
  )
  testthat::expect_error(
    fit_eivgp(
      X, y, C2, U_obs = U_rank_deficient, latent_dim = 2L,
      engine = "multivariate"
    ),
    "explicit ident choice"
  )
  testthat::expect_identical(
    mixedgp_resolve_loading_rule("none", 2L, FALSE), "none"
  )
  testthat::expect_error(
    fit_eivgp(X, y, C2, latent_dim = 1.5, engine = "multivariate"),
    "integer-valued"
  )

  duplicate_X <- matrix(1:6, nrow = 3L)
  colnames(duplicate_X) <- c("x", "x")
  testthat::expect_error(
    as_numeric_matrix_strict(duplicate_X, "X"), "complete and unique"
  )
  duplicate_C <- data.frame(
    ordered(c("a", "b", "a")), ordered(c("c", "d", "c")),
    check.names = FALSE
  )
  names(duplicate_C) <- c("proxy", "proxy")
  testthat::expect_error(
    prepare_ordinal_matrix(duplicate_C, name = "C"), "complete and unique"
  )
  testthat::expect_error(
    mixedgp_do_call(identity, list(x = 1), list(2)), "must be named"
  )
})

testthat::test_that("package loading has no reporting or thread side effects", {
  package_path <- getNamespaceInfo(asNamespace("eivmixgp"), "path")
  check_script <- tempfile("eivmixgp-load-side-effects-", fileext = ".R")
  on.exit(unlink(check_script), add = TRUE)
  writeLines(
    c(
      "args <- commandArgs(trailingOnly = TRUE)",
      "package_path <- normalizePath(args[[1L]], winslash = '/', mustWork = TRUE)",
      "thread_vars <- c('OMP_NUM_THREADS', 'OPENBLAS_NUM_THREADS',",
      "  'MKL_NUM_THREADS', 'VECLIB_MAXIMUM_THREADS', 'NUMEXPR_NUM_THREADS')",
      "reporting <- c('ggplot2', 'dplyr', 'tidyr')",
      "have_ggplot <- requireNamespace('ggplot2', quietly = TRUE)",
      "theme_before <- if (have_ggplot) ggplot2::theme_get() else NULL",
      "loaded_before <- reporting %in% loadedNamespaces()",
      "names(loaded_before) <- reporting",
      "search_before <- search()",
      "env_before <- Sys.getenv(thread_vars, unset = NA_character_)",
      "source_modules <- c(",
      "  '00_parallel_utils.R', '00_study1_functions.R',",
      "  '00_study2_functions.R', '03_study2_published_competitors.R',",
      "  '00_synthetic_data.R', '00_public_api.R',",
      "  '00_experiment_runner.R')",
      "source_paths <- file.path(package_path, 'R', source_modules)",
      "if (all(file.exists(source_paths))) {",
      "  package_env <- new.env(parent = baseenv())",
      "  for (path in source_paths) sys.source(path, envir = package_env)",
      "} else {",
      "  suppressPackageStartupMessages(library(",
      "    'eivmixgp', character.only = TRUE, lib.loc = dirname(package_path)",
      "  ))",
      "}",
      "search_after <- search()",
      "allowed <- unique(c(search_before, 'package:eivmixgp'))",
      "stopifnot(length(setdiff(search_after, allowed)) == 0L)",
      "stopifnot(identical(",
      "  Sys.getenv(thread_vars, unset = NA_character_), env_before",
      "))",
      "loaded_after <- reporting %in% loadedNamespaces()",
      "names(loaded_after) <- reporting",
      "stopifnot(identical(loaded_after, loaded_before))",
      "if (have_ggplot) stopifnot(identical(ggplot2::theme_get(), theme_before))"
    ),
    check_script
  )
  output <- system2(
    file.path(R.home("bin"), "Rscript"),
    c("--vanilla", shQuote(check_script), shQuote(package_path)),
    stdout = TRUE,
    stderr = TRUE
  )
  status <- attr(output, "status")
  testthat::expect_true(
    is.null(status) || identical(as.integer(status), 0L),
    info = paste(output, collapse = "\n")
  )
})

testthat::test_that("tail-stable truncated normals respect extreme support", {
  lower <- rep(c(8, -Inf, 40, -41), each = 64L)
  upper <- rep(c(Inf, -8, 41, -40), each = 64L)
  set.seed(4103)
  z <- rtruncnorm_vec(0, 1, lower, upper)
  testthat::expect_true(all(is.finite(z)))
  testthat::expect_true(all(z >= lower & z <= upper))
})

testthat::test_that("stored synthetic data round trips", {
  dat <- make_study1_synthetic_dataset(
    rep_id = 1, n = 30, n_test = 10, m = 6, min_class_count = 1,
    calib_grid = c(0L, 5L, 10L, 20L, 30L)
  )
  testthat::expect_s3_class(dat, "mixedgp_synthetic_dataset")
  testthat::expect_equal(nrow(dat$data$train), 30)
})

testthat::test_that("frozen-data filename collisions fail closed", {
  data_dir <- tempfile("eivmixgp-frozen-")
  dir.create(data_dir)
  on.exit(unlink(data_dir, recursive = TRUE), add = TRUE)

  original <- make_study1_synthetic_dataset(
    rep_id = 2, n = 30, n_test = 10, m = 6, min_class_count = 1,
    heterogeneity_eta = 0, calib_grid = c(0L, 5L, 10L, 20L, 30L)
  )
  collision <- make_study1_synthetic_dataset(
    rep_id = 2, n = 30, n_test = 10, m = 6, min_class_count = 1,
    heterogeneity_eta = 1, calib_grid = c(0L, 5L, 10L, 20L, 30L)
  )
  manifest <- store_mixedgp_synthetic_dataset(original, data_dir)
  saveRDS(manifest, file.path(data_dir, "manifest.rds"), version = 3L)
  testthat::expect_error(
    store_mixedgp_synthetic_dataset(collision, data_dir),
    "different synthetic dataset"
  )
  reloaded <- load_mixedgp_synthetic_dataset(
    file.path(data_dir, manifest$file)
  )
  testthat::expect_identical(reloaded, original)
  strict <- load_mixedgp_synthetic_dataset_strict(
    file.path(data_dir, manifest$file),
    expected = list(
      study = "study1", scenario = "active", rep_id = 2L,
      n = 30L, n_test = 10L, m = 6L,
      calib_grid = c(0L, 5L, 10L, 20L, 30L)
    )
  )
  testthat::expect_identical(strict$data, original$data)
  testthat::expect_identical(strict$design, original$design)
  testthat::expect_identical(attr(strict, "manifest_md5"), manifest$md5)
  testthat::expect_error(
    load_mixedgp_synthetic_dataset_strict(
      file.path(data_dir, manifest$file),
      expected = list(study = "study1", n = 31L)
    ),
    "expected 31"
  )
  testthat::expect_error(
    make_nested_calibration_sets(30L, c(0L, 31L), seed = 1L),
    "between zero and n"
  )
})

testthat::test_that("synthetic manifests merge atomically and fail on worker errors", {
  data_dir <- tempfile("eivmixgp-incremental-")
  dir.create(data_dir)
  on.exit(unlink(data_dir, recursive = TRUE), add = TRUE)

  first <- generate_study2_synthetic_datasets(
    n_rep = 1L, scenarios = "primary", directory = data_dir,
    n_cores = 1L, n = 12L, n_test = 5L, q = 2L, m = 3L,
    calib_grid = c(0L, 3L)
  )
  merged <- generate_study2_synthetic_datasets(
    n_rep = 1L, scenarios = "high_uncertainty", directory = data_dir,
    n_cores = 1L, n = 12L, n_test = 5L, q = 2L, m = 3L,
    calib_grid = c(0L, 3L)
  )
  testthat::expect_equal(nrow(first), 1L)
  testthat::expect_equal(nrow(merged), 2L)
  testthat::expect_identical(readRDS(file.path(data_dir, "manifest.rds")), merged)
  testthat::expect_setequal(
    utils::read.csv(file.path(data_dir, "manifest.csv"))$file, merged$file
  )
  testthat::expect_true(all(nzchar(merged$generator_tag)))
  testthat::expect_true(all(grepl("^[0-9a-f]{32}$", merged$design_fingerprint)))

  failed <- structure("worker boom", class = "try-error")
  testthat::expect_error(
    mixedgp_stop_on_worker_errors(list(failed), "generation"),
    "no manifest was written"
  )
  inconsistent <- merged[1L, , drop = FALSE]
  inconsistent$md5 <- paste(rep("0", 32L), collapse = "")
  testthat::expect_error(
    mixedgp_merge_manifests(merged, inconsistent), "metadata disagree"
  )
  testthat::expect_error(
    generate_study2_synthetic_datasets(
      n_rep = 0L, scenarios = "primary", directory = data_dir
    ),
    "n_rep"
  )
  testthat::expect_error(
    generate_study2_synthetic_datasets(
      n_rep = 1L, scenarios = character(), directory = data_dir
    ),
    "one or more"
  )
})

testthat::test_that("Study II frozen truth and fingerprints are validated", {
  artifact <- make_study2_synthetic_dataset(
    rep_id = 1L, scenario = "primary", n = 12L, n_test = 5L,
    q = 2L, m = 3L, calib_grid = c(0L, 3L)
  )
  testthat::expect_invisible(validate_mixedgp_synthetic_dataset(artifact))

  bad_U <- artifact
  bad_U$data$train$U[1L, 1L] <- NA_real_
  testthat::expect_error(
    validate_mixedgp_synthetic_dataset(bad_U), "Study II data"
  )
  bad_truth <- artifact
  bad_truth$data$test$f <- bad_truth$data$test$f[-1L]
  testthat::expect_error(
    validate_mixedgp_synthetic_dataset(bad_truth), "Study II data"
  )
  stale <- artifact
  stale$design$generator_tag <- "stale-generator"
  testthat::expect_error(
    validate_mixedgp_synthetic_dataset(stale), "generator tag"
  )
})

testthat::test_that("Study II oracle pools are tied to their generating model", {
  pars <- make_study2_true_params("primary", q = 4L, m = 4L)
  pool <- make_oracle_pool_2d(pars, n_pool = 200L, seed = 41025L)
  testthat::expect_invisible(validate_oracle_pool_2d(pool, pars))
  changed <- pars
  changed$A[1L, 1L] <- changed$A[1L, 1L] + 0.01
  testthat::expect_error(
    validate_oracle_pool_2d(pool, changed),
    "provenance"
  )
})

testthat::test_that("ordinal factor maps survive dropped prediction levels", {
  training <- data.frame(severity = ordered(
    c("low", "high", "medium"),
    levels = c("low", "medium", "high")
  ))
  encoded <- prepare_ordinal_matrix(training, name = "C")
  prediction <- data.frame(severity = ordered(
    c("high", "low"), levels = c("low", "high")
  ))
  encoded_new <- prepare_ordinal_matrix(
    prediction,
    m_vec = encoded$m_vec,
    level_maps = encoded$level_maps,
    expected_names = encoded$column_names,
    name = "new_C"
  )
  testthat::expect_identical(unname(encoded_new$C[, 1L]), c(3L, 1L))
  testthat::expect_error(
    prepare_ordinal_matrix(
      data.frame(severity = ordered("unseen")),
      m_vec = encoded$m_vec,
      level_maps = encoded$level_maps,
      expected_names = encoded$column_names,
      name = "new_C"
    ),
    "unknown level"
  )
})

testthat::test_that("multivariate latent dimension is never silently guessed", {
  X <- matrix(c(-1, 0, 1), ncol = 1L)
  y <- c(-0.5, 0, 0.5)
  C <- data.frame(
    proxy_a = ordered(c("low", "medium", "high")),
    proxy_b = ordered(c("none", "mild", "severe"))
  )
  testthat::expect_error(
    fit_eivgp(X, y, C, engine = "multivariate", U_obs = NULL),
    "latent_dim must be supplied explicitly"
  )
  testthat::expect_error(
    fit_eivgp(
      X, y, C, engine = "multivariate",
      U_obs = matrix(NA_real_, nrow = 3L, ncol = 2L)
    ),
    "U_obs contains no calibration values"
  )

  rank_deficient <- mixedgp_latent_anchor_status(
    matrix(c(0, 0, 1, 1), ncol = 2L, byrow = TRUE), 1:2, d = 2L
  )
  anchored <- mixedgp_latent_anchor_status(
    matrix(c(0, 0, 1, 0, 0, 1), ncol = 2L, byrow = TRUE), 1:3, d = 2L
  )
  testthat::expect_false(rank_deficient$anchored)
  testthat::expect_identical(rank_deficient$affine_rank, 2L)
  testthat::expect_true(anchored$anchored)
  testthat::expect_identical(anchored$affine_rank, 3L)
})

testthat::test_that("duplicate-point GP predictions remain positive semidefinite", {
  prediction <- gp_predict_draw_general(
    X_train = matrix(c(-1, 0, 1), ncol = 1L),
    U_train = matrix(c(-0.5, 0, 0.5), ncol = 1L),
    y_train = c(-0.7, 0, 0.8),
    X_star = matrix(c(0.25, 0.25), ncol = 1L),
    U_star = matrix(c(0.1, 0.1), ncol = 1L),
    logtheta = log(c(1, 1, 1)),
    sigma2_eps = 0.1,
    return_cov = TRUE
  )
  testthat::expect_equal(prediction$mean[1L], prediction$mean[2L])
  testthat::expect_true(
    min(eigen(prediction$cov, symmetric = TRUE, only.values = TRUE)$values) >=
      -1e-10
  )
  set.seed(4104)
  draws <- rmvnorm_psd(32L, prediction$mean, prediction$cov)
  testthat::expect_true(all(is.finite(draws)))
  testthat::expect_equal(draws[, 1L], draws[, 2L], tolerance = 1e-7)
})

testthat::test_that("optional exact no-HMC controls and prior maps are available", {
  testthat::expect_identical(formals(fit_eivgp_1d)$noise_strategy, "conditional")
  control <- make_default_control_ordprobit(24L, 18L, "balanced", d = 2L)
  testthat::expect_identical(control$joint_measurement_every, 0L)
  testthat::expect_identical(control$loading_transport_every, 0L)
  bad <- control
  bad$joint_theta_every <- bad$joint_global_every <- bad$joint_collapsed_every <- 0L
  testthat::expect_error(validate_control_ordprobit(bad, "interwoven", 18L, 100L),
                         "freezes all GP")
  z <- c(.3, -.2, .1, -.4, .7)
  mapped <- measurement_row_from_gaussian(z, j = 2L, d = 2L, K = 3L,
                                         ident = "lower_triangular")
  testthat::expect_true(mapped$A[2L] > 0)
  testthat::expect_true(all(diff(c(-6, mapped$tau, 6)) > 0))
  testthat::expect_equal(measurement_row_to_gaussian(mapped$A, mapped$tau,
                         j = 2L, ident = "lower_triangular"), z, tolerance = 1e-12)
  state <- list(logdetA = 1.2, quad = 4.3)
  expected <- -.5 * (3 * log(2 * base::pi) + state$logdetA) +
    2 * log(.05) - lgamma(2) + lgamma(3.5) - 3.5 * log(.05 + state$quad / 2)
  testthat::expect_equal(gp_update_loglik_1d(state, 3L, TRUE), expected)
})

testthat::test_that("sampler safety caps fail explicitly", {
  testthat::expect_error(
    bounded_slice_update_1d(
      x0 = 0, logf = function(x) stats::dnorm(x, log = TRUE),
      max_steps_out = 0L, max_iter = 0L, fail_on_limit = TRUE
    ),
    "Slice width and search budgets must be positive"
  )

  X <- matrix(c(-0.5, 0.5), ncol = 1L)
  U <- matrix(c(-0.5, 0.5), ncol = 1L)
  C <- matrix(c(1L, 2L), ncol = 1L)
  prior <- make_gp_prior(p = 1L, d = 1L)
  testthat::expect_error(
    update_U_theta_ess_integrated_general(
      y = c(-0.2, 0.2), X = X, C = C, U_curr = U,
      logtheta_curr = prior$mean, gp_prior = prior,
      block_idx = 1:2, reference = "prior",
      A = matrix(1, nrow = 1L), tau = list(0), max_try = 0L
    ),
    "exceeded max_try"
  )
})

testthat::test_that("both engines expose mean inference and labeled imputation", {
  univariate <- mixedgp_test_univariate_fit()
  multivariate <- mixedgp_test_multivariate_fit()
  new_X <- matrix(
    c(-0.25, 0.25), ncol = 1L, dimnames = list(NULL, "x1")
  )
  new_C_1d <- data.frame(severity = ordered(
    c("high", "low"), levels = c("low", "high")
  ))
  new_C_multi <- data.frame(
    proxy_a = ordered(c("medium", "low"), levels = c("low", "medium")),
    proxy_b = ordered(c("mild", "none"), levels = c("none", "mild"))
  )

  mean_1d <- predict_eivgp(
    univariate, new_X, new_C_1d, target = "mean", draw_ids = 1:2,
    n_latent = 4L, include_process_uncertainty = FALSE, seed = 4201
  )
  mean_multi <- predict_eivgp(
    multivariate, new_X, new_C_multi, target = "mean", draw_ids = 1:2,
    n_latent = 4L, include_process_uncertainty = FALSE,
    seed = 4202
  )
  testthat::expect_identical(dim(mean_1d), c(2L, 2L))
  testthat::expect_identical(dim(mean_multi), c(2L, 2L))
  testthat::expect_true(all(is.finite(mean_1d)))
  testthat::expect_true(all(is.finite(mean_multi)))
  testthat::expect_identical(attr(mean_1d, "target"), "m(x,c)")
  testthat::expect_identical(attr(mean_multi, "target"), "m(x,c)")

  train_1d <- impute_eivgp(univariate, rows = 1:2, draw_ids = 1:2)
  train_multi <- impute_eivgp(multivariate, rows = 1:2, draw_ids = 1:2)
  prospective_1d <- impute_eivgp(
    univariate, new_C = new_C_1d, draw_ids = 1L,
    n_per_draw = 2L, seed = 4203
  )
  prospective_multi <- impute_eivgp(
    multivariate, new_C = new_C_multi, draw_ids = 1L,
    n_per_draw = 2L, seed = 4204
  )
  testthat::expect_identical(dim(train_1d), c(2L, 2L, 1L))
  testthat::expect_identical(dim(train_multi), c(2L, 2L, 1L))
  testthat::expect_identical(attr(train_1d, "source"), "training")
  testthat::expect_identical(attr(train_multi, "source"), "training")
  testthat::expect_identical(attr(prospective_1d, "source"), "prospective")
  testthat::expect_identical(attr(prospective_multi, "source"), "prospective")
  testthat::expect_identical(attr(prospective_1d, "scale"), "working")
  testthat::expect_identical(attr(prospective_multi, "scale"), "working")
  testthat::expect_false(attr(prospective_1d, "latent_scale_anchored"))
  testthat::expect_false(attr(prospective_multi, "latent_scale_anchored"))
  testthat::expect_identical(
    dimnames(prospective_1d)[[3L]], "latent_severity"
  )
  testthat::expect_identical(dimnames(prospective_multi)[[3L]], "latent1")
  testthat::expect_true(all(prospective_1d[, 1L, 1L] > 0.4))
  testthat::expect_true(all(prospective_1d[, 2L, 1L] < -0.25))
  testthat::expect_identical(
    attr(prospective_multi, "latent_sampler"), "minimax_tilting"
  )
})

testthat::test_that("fit S3 methods report identification and delegate prediction", {
  univariate <- mixedgp_test_univariate_fit()
  multivariate <- mixedgp_test_multivariate_fit()
  summary_1d <- summary.eivgp_fit(univariate)
  summary_multi <- summary.eivgp_fit(multivariate)

  testthat::expect_s3_class(summary_1d, "summary_eivgp_fit")
  testthat::expect_s3_class(summary_multi, "summary_eivgp_fit")
  testthat::expect_identical(summary_1d$engine, "univariate")
  testthat::expect_identical(summary_multi$engine, "multivariate")
  testthat::expect_identical(
    summary_1d$dimensions, c(n = 3L, p = 1L, q = 1L, d = 1L)
  )
  testthat::expect_identical(
    summary_multi$dimensions, c(n = 3L, p = 1L, q = 2L, d = 1L)
  )
  testthat::expect_identical(summary_1d$calibration$status, "none")
  testthat::expect_identical(summary_multi$calibration$status, "none")
  testthat::expect_false(summary_1d$calibration$anchored)
  testthat::expect_false(summary_multi$calibration$anchored)
  testthat::expect_match(
    summary_1d$calibration$interpretation, "No calibration data"
  )
  testthat::expect_equal(summary_1d$convergence$max_rhat_tau, 1.02)
  testthat::expect_equal(summary_multi$convergence$min_ess_key, 80)
  testthat::expect_false("mean_u_ess_accept" %in% names(summary_multi$convergence))
  printed_fit <- paste(
    utils::capture.output(print.eivgp_fit(univariate)), collapse = "\n"
  )
  printed_summary <- paste(
    utils::capture.output(print.summary_eivgp_fit(summary_multi)),
    collapse = "\n"
  )
  if (!grepl("working latent scale", printed_fit, fixed = TRUE)) {
    stop("The fitted-model print method omitted its latent-scale label.")
  }
  if (!grepl("No calibration data", printed_summary, fixed = TRUE)) {
    stop("The summary print method omitted its calibration interpretation.")
  }

  anchored_1d <- univariate
  anchored_1d$data$calib_idx <- 1:2
  anchored_1d$data$latent_scale_anchored <- TRUE
  anchored_1d$data$latent_anchor_rank <- 2L
  anchored_summary <- summary.eivgp_fit(anchored_1d)
  testthat::expect_identical(
    anchored_summary$calibration$status, "affine_anchored"
  )
  testthat::expect_identical(anchored_summary$calibration$output_scale, "raw")

  rank_deficient_multi <- multivariate
  rank_deficient_multi$data$calib_idx <- 1L
  rank_deficient_multi$data$latent_anchor_rank <- 1L
  rank_deficient_summary <- summary.eivgp_fit(rank_deficient_multi)
  testthat::expect_identical(
    rank_deficient_summary$calibration$status, "rank_deficient"
  )
  testthat::expect_match(
    rank_deficient_summary$calibration$interpretation,
    "affine rank 1 of 2"
  )

  new_X <- matrix(
    c(-0.25, 0.25), ncol = 1L, dimnames = list(NULL, "x1")
  )
  new_C <- data.frame(severity = ordered(
    c("high", "low"), levels = c("low", "high")
  ))
  via_s3 <- predict.eivgp_fit(
    univariate, new_X = new_X, new_C = new_C, target = "mean",
    draw_ids = 1L, n_latent = 4L,
    include_process_uncertainty = FALSE, seed = 4210
  )
  via_public <- predict_eivgp(
    univariate, new_X = new_X, new_C = new_C, target = "mean",
    draw_ids = 1L, n_latent = 4L,
    include_process_uncertainty = FALSE, seed = 4210
  )
  testthat::expect_identical(via_s3, via_public)
})

testthat::test_that("exact new U and unanchored scale semantics are explicit", {
  univariate <- mixedgp_test_univariate_fit()
  multivariate <- mixedgp_test_multivariate_fit()
  new_X <- matrix(
    c(-0.25, 0.25), ncol = 1L, dimnames = list(NULL, "x1")
  )
  new_U_multi <- matrix(
    c(-0.1, 0.2), ncol = 1L,
    dimnames = list(NULL, "latent1")
  )
  response_1d <- predict_eivgp(
    univariate, new_X, new_U = c(-0.1, 0.2),
    target = "response", draw_ids = 1:2, seed = 4205
  )
  response_multi <- predict_eivgp(
    multivariate, new_X, new_U = new_U_multi,
    target = "response", draw_ids = 1:2, seed = 4206
  )
  testthat::expect_true(all(is.finite(response_1d)))
  testthat::expect_true(all(is.finite(response_multi)))
  testthat::expect_identical(dim(response_1d), c(2L, 2L))
  testthat::expect_identical(dim(response_multi), c(2L, 2L))
  testthat::expect_identical(attr(response_1d, "latent_input_scale"), "working")
  testthat::expect_identical(
    attr(response_multi, "latent_input_scale"), "working"
  )
  testthat::expect_true(all(is.na(
    attr(response_multi, "rejection_acceptance")
  )))
  testthat::expect_false(attr(response_1d, "latent_scale_anchored"))
  testthat::expect_false(attr(response_multi, "latent_scale_anchored"))

  imputed_1d <- impute_eivgp(univariate, rows = 1:2, scale = "auto")
  imputed_multi <- impute_eivgp(multivariate, rows = 1:2, scale = "auto")
  testthat::expect_identical(attr(imputed_1d, "scale"), "working")
  testthat::expect_identical(attr(imputed_multi, "scale"), "working")
  testthat::expect_false(attr(imputed_1d, "latent_scale_anchored"))
  testthat::expect_false(attr(imputed_multi, "latent_scale_anchored"))
  testthat::expect_error(
    impute_eivgp(univariate, scale = "raw"), "requires calibration data"
  )
  testthat::expect_error(
    impute_eivgp(multivariate, scale = "raw"), "requires calibration data"
  )
  testthat::expect_error(
    predict_eivgp(
      univariate, new_X, new_U = c(-0.1, 0.2),
      new_U_scale = "raw", target = "response"
    ),
    "requires calibration data"
  )
  testthat::expect_error(
    predict_eivgp(
      multivariate, new_X, new_U = new_U_multi,
      new_U_scale = "raw", target = "response"
    ),
    "requires calibration data"
  )
})

testthat::test_that("irrelevant prediction and imputation controls are rejected", {
  univariate <- mixedgp_test_univariate_fit()
  multivariate <- mixedgp_test_multivariate_fit()
  new_X <- matrix(
    c(-0.25, 0.25), ncol = 1L, dimnames = list(NULL, "x1")
  )
  new_C_1d <- data.frame(severity = ordered(
    c("high", "low"), levels = c("low", "high")
  ))
  new_C_multi <- data.frame(
    proxy_a = ordered(c("medium", "low"), levels = c("low", "medium")),
    proxy_b = ordered(c("mild", "none"), levels = c("none", "mild"))
  )

  testthat::expect_error(
    predict_eivgp(
      univariate, new_X, new_C_1d, new_U = c(0, 0), target = "mean"
    ),
    "new_U is not used"
  )
  testthat::expect_error(
    predict_eivgp(
      univariate, new_X, new_C_1d, new_U = c(0, 0), target = "surface"
    ),
    "new_C is not used"
  )
  testthat::expect_error(
    predict_eivgp(
      univariate, new_X, new_C_1d, target = "mean", new_U_scale = "model"
    ),
    "only when new_U"
  )
  testthat::expect_error(
    predict_eivgp(
      univariate, new_X, new_C_1d, target = "response",
      include_process_uncertainty = FALSE
    ),
    "applies only"
  )
  testthat::expect_error(
    predict_eivgp(
      univariate, new_X, new_C_1d, target = "response", n_latent = 4L
    ),
    "only to target='mean'"
  )
  testthat::expect_error(
    predict_eivgp(
      univariate, new_X, new_C_1d, target = "mean",
      latent_sampler = "rejection"
    ),
    "univariate engine"
  )
  testthat::expect_error(
    predict_eivgp(
      multivariate, new_X, new_C_multi, target = "mean",
      rejection_batch_size = 64L
    ),
    "only for latent_sampler='rejection'"
  )
  testthat::expect_error(
    impute_eivgp(univariate, n_per_draw = NA_integer_), "integer-valued"
  )
  testthat::expect_error(
    impute_eivgp(univariate, seed = 1L), "not used"
  )
  testthat::expect_error(
    impute_eivgp(
      univariate, new_C = new_C_1d, latent_sampler = "rejection"
    ),
    "univariate engine"
  )
  testthat::expect_error(
    impute_eivgp(
      multivariate, new_C = new_C_multi, rejection_max_batches = 2L
    ),
    "only for latent_sampler='rejection'"
  )
})

testthat::test_that("joint and pointwise prediction modes are labeled", {
  univariate <- mixedgp_test_univariate_fit()
  multivariate <- mixedgp_test_multivariate_fit()
  new_X <- matrix(
    c(-0.25, 0.25), ncol = 1L, dimnames = list(NULL, "x1")
  )
  new_C_1d <- data.frame(severity = ordered(
    c("high", "low"), levels = c("low", "high")
  ))
  new_C_multi <- data.frame(
    proxy_a = ordered(c("medium", "low"), levels = c("low", "medium")),
    proxy_b = ordered(c("mild", "none"), levels = c("none", "mild"))
  )

  pointwise_1d <- predict_eivgp(
    univariate, new_X, new_C_1d, target = "mean", draw_ids = 1L,
    n_latent = 4L, joint = FALSE, seed = 4207
  )
  joint_1d <- predict_eivgp(
    univariate, new_X, new_C_1d, target = "mean", draw_ids = 1L,
    n_latent = 4L, joint = TRUE, seed = 4207
  )
  pointwise_multi <- predict_eivgp(
    multivariate, new_X, new_C_multi, target = "mean", draw_ids = 1L,
    n_latent = 4L, joint = FALSE, seed = 4208
  )
  joint_multi <- predict_eivgp(
    multivariate, new_X, new_C_multi, target = "mean", draw_ids = 1L,
    n_latent = 4L, joint = TRUE, seed = 4208
  )
  testthat::expect_false(attr(pointwise_1d, "joint"))
  testthat::expect_true(attr(joint_1d, "joint"))
  testthat::expect_false(attr(pointwise_multi, "joint"))
  testthat::expect_true(attr(joint_multi, "joint"))
  testthat::expect_true(attr(joint_1d, "include_process_uncertainty"))
  testthat::expect_true(attr(joint_multi, "include_process_uncertainty"))
})

testthat::test_that("empirical-Q GP integration matches brute-force aggregation", {
  X_train <- matrix(c(-1, 0, 1), ncol = 1L)
  U_train <- matrix(c(-0.6, 0.1, 0.8), ncol = 1L)
  y_train <- c(-0.5, 0.1, 0.7)
  X_star <- matrix(c(-0.2, 0.4), ncol = 1L)
  U_mc <- array(NA_real_, dim = c(3L, 2L, 1L))
  U_mc[, 1L, 1L] <- c(-0.8, -0.1, 0.5)
  U_mc[, 2L, 1L] <- c(-0.3, 0.4, 1)
  logtheta <- log(c(1.2, 0.7, 0.9))
  sigma2_eps <- 0.2

  integrated <- gp_integrated_mean_state_general(
    X_train, U_train, y_train, X_star, U_mc,
    logtheta = logtheta, sigma2_eps = sigma2_eps, return_cov = TRUE
  )
  M <- dim(U_mc)[1L]
  N <- nrow(X_star)
  brute_mean <- vapply(seq_len(N), function(i) {
    mean(vapply(seq_len(M), function(a) {
      gp_predict_draw_general(
        X_train, U_train, y_train,
        X_star = X_star[i, , drop = FALSE],
        U_star = matrix(U_mc[a, i, ], nrow = 1L),
        logtheta = logtheta, sigma2_eps = sigma2_eps
      )$mean
    }, numeric(1L)))
  }, numeric(1L))
  brute_cov <- matrix(NA_real_, N, N)
  for (i in seq_len(N)) {
    for (j in seq_len(N)) {
      values <- numeric(M * M)
      pos <- 0L
      for (a in seq_len(M)) {
        for (b in seq_len(M)) {
          pos <- pos + 1L
          pair <- gp_predict_draw_general(
            X_train, U_train, y_train,
            X_star = rbind(
              X_star[i, , drop = FALSE], X_star[j, , drop = FALSE]
            ),
            U_star = matrix(
              c(U_mc[a, i, ], U_mc[b, j, ]), ncol = 1L
            ),
            logtheta = logtheta, sigma2_eps = sigma2_eps,
            return_cov = TRUE
          )
          values[pos] <- pair$cov[1L, 2L]
        }
      }
      brute_cov[i, j] <- mean(values)
    }
  }
  testthat::expect_equal(integrated$mean, brute_mean, tolerance = 1e-10)
  testthat::expect_equal(integrated$cov, brute_cov, tolerance = 1e-10)

  integrated_1d <- gp_integrated_mean_state_1d(
    X_train, as.numeric(U_train), y_train, X_star, U_mc[, , 1L],
    logtheta = logtheta, sigma2_eps = sigma2_eps, return_cov = TRUE
  )
  brute_mean_1d <- vapply(seq_len(N), function(i) {
    mean(vapply(seq_len(M), function(a) {
      gp_predict_draw(
        X_train, as.numeric(U_train), y_train,
        x_star = X_star[i, , drop = FALSE],
        u_star = U_mc[a, i, 1L],
        logtheta = logtheta, sigma2_eps = sigma2_eps
      )$mean
    }, numeric(1L)))
  }, numeric(1L))
  brute_cov_1d <- matrix(NA_real_, N, N)
  for (i in seq_len(N)) {
    for (j in seq_len(N)) {
      values <- numeric(M * M)
      pos <- 0L
      for (a in seq_len(M)) {
        for (b in seq_len(M)) {
          pos <- pos + 1L
          pair <- gp_predict_draw(
            X_train, as.numeric(U_train), y_train,
            x_star = rbind(
              X_star[i, , drop = FALSE], X_star[j, , drop = FALSE]
            ),
            u_star = c(U_mc[a, i, 1L], U_mc[b, j, 1L]),
            logtheta = logtheta, sigma2_eps = sigma2_eps,
            return_cov = TRUE
          )
          values[pos] <- pair$cov[1L, 2L]
        }
      }
      brute_cov_1d[i, j] <- mean(values)
    }
  }
  testthat::expect_equal(
    integrated_1d$mean, brute_mean_1d, tolerance = 1e-10
  )
  testthat::expect_equal(
    integrated_1d$cov, brute_cov_1d, tolerance = 1e-10
  )
})

testthat::test_that("both numerical studies use one validated interface", {
  s1 <- run_mixedgp_experiment("study1", dry_run = TRUE)
  s2 <- run_mixedgp_experiment("study2", dry_run = TRUE)
  testthat::expect_s3_class(s1, "mixedgp_experiment_spec")
  testthat::expect_s3_class(s2, "mixedgp_experiment_spec")
  testthat::expect_identical(s1$config, s2$config)
  testthat::expect_identical(s1$stages, s2$stages)

  thorough <- run_mixedgp_experiment(
    "study1", config = "thorough", stages = "data", dry_run = TRUE
  )
  run_env <- mixedgp_experiment_environment(thorough)
  mixedgp_assign_experiment_controls(thorough, run_env)
  testthat::expect_false(run_env$STUDY1_REUSE_LOCKED_EIV)

  abbreviated <- run_mixedgp_experiment(
    "study1", config = "t", stages = "data", dry_run = TRUE
  )
  quick <- run_mixedgp_experiment(
    "study1", config = "q", stages = "data", dry_run = TRUE
  )
  explicit <- run_mixedgp_experiment(
    "study1", config = "thorough", stages = "data",
    strict_competitors = FALSE, dry_run = TRUE
  )
  testthat::expect_identical(abbreviated$config, "thorough")
  testthat::expect_true(abbreviated$strict_competitors)
  testthat::expect_false(quick$strict_competitors)
  testthat::expect_false(explicit$strict_competitors)

  serial_spec <- run_mixedgp_experiment(
    "study1", stages = "data", parallel_level = "none", n_cores = 4L,
    dry_run = TRUE
  )
  testthat::expect_identical(mixedgp_experiment_data_cores(serial_spec), 1L)
  testthat::expect_false(mixedgp_parallel_chains_enabled("none", 4L))
  chain_spec <- run_mixedgp_experiment(
    "study1", stages = "data", parallel_level = "chains", n_cores = 4L,
    dry_run = TRUE
  )
  replication_spec <- run_mixedgp_experiment(
    "study1", stages = "data", parallel_level = "replications", n_cores = 2L,
    dry_run = TRUE
  )
  testthat::expect_identical(mixedgp_experiment_data_cores(chain_spec), 1L)
  testthat::expect_identical(
    mixedgp_experiment_data_cores(replication_spec),
    replication_spec$execution$data$cores
  )
  testthat::expect_identical(chain_spec$execution$data$backend, "serial")
  policy_scripts <- file.path(
    s1$code_dir,
    c(
      "01_study1_representative_figures.R",
      "02_study1_targeted_controls.R",
      "01_study2_representative_figures.R"
    )
  )
  for (path in policy_scripts) {
    text <- readLines(path, warn = FALSE)
    testthat::expect_true(any(grepl("mixedgp_parallel_chains_enabled", text)))
    testthat::expect_false(any(grepl("parallel::detectCores", text)))
  }

  testthat::expect_error(
    run_mixedgp_experiment(
      "study2", stages = "data", dry_run = TRUE,
      study_options = list(STUDY2_MC_N_REPP = 2L)
    ),
    "Unknown study_options"
  )
  testthat::expect_error(
    run_mixedgp_experiment(
      "study2", stages = "data", dry_run = TRUE,
      study_options = list(STUDY2_MC_N_REP = 2.5)
    ),
    "integer-valued"
  )
  testthat::expect_error(
    run_mixedgp_experiment(
      "study1", stages = "data", use_cache = 1, dry_run = TRUE
    ),
    "use_cache must be TRUE or FALSE"
  )
})

testthat::test_that("experiment execution preserves RNG and package search state", {
  testthat::skip_if(
    isTRUE(getOption("knitr.in.progress")),
    "requires the installed-style package namespace; exercised by R CMD check"
  )
  output_dir <- tempfile("eivmixgp-runner-")
  dir.create(output_dir)
  on.exit(unlink(output_dir, recursive = TRUE), add = TRUE)
  testthat::local_mocked_bindings(
    mixedgp_generate_experiment_data = function(spec) {
      set.seed(9010L)
      stats::runif(2L)
      data.frame()
    },
    .package = "eivmixgp"
  )
  set.seed(4114)
  expected <- stats::runif(4)
  set.seed(4114)
  invisible(run_mixedgp_experiment(
    "study1", stages = "data", output_dir = output_dir,
    parallel_level = "none"
  ))
  testthat::expect_identical(stats::runif(4), expected)

  testthat::skip_if_not_installed("ggplot2")
  spec <- run_mixedgp_experiment(
    "study1", stages = "data", output_dir = output_dir, dry_run = TRUE
  )
  run_env <- mixedgp_experiment_environment(spec)
  search_before <- search()
  eval(quote(suppressPackageStartupMessages(library(ggplot2))), run_env)
  testthat::expect_identical(search(), search_before)
  testthat::expect_true(exists("ggplot", envir = run_env, inherits = FALSE))
})

testthat::test_that("core-count environment variables are parsed strictly", {
  old <- Sys.getenv("MIXEDGP_CORES", unset = NA_character_)
  on.exit({
    if (is.na(old)) Sys.unsetenv("MIXEDGP_CORES") else
      Sys.setenv(MIXEDGP_CORES = old)
  }, add = TRUE)
  Sys.setenv(MIXEDGP_CORES = "2")
  testthat::expect_identical(mixedgp_resolve_cores(), 2L)
  Sys.setenv(MIXEDGP_CORES = "2.5")
  testthat::expect_error(mixedgp_resolve_cores(), "base-10 integer")
})

testthat::test_that("experiment defaults do not leak from the global workspace", {
  sentinel <- "MIXEDGP_GLOBAL_SENTINEL_FOR_TEST"
  assign(sentinel, TRUE, envir = .GlobalEnv)
  on.exit(rm(list = sentinel, envir = .GlobalEnv), add = TRUE)

  spec <- run_mixedgp_experiment(
    "study1", stages = "data", dry_run = TRUE
  )
  run_env <- mixedgp_experiment_environment(spec)
  testthat::expect_false(eval(substitute(exists(NAME), list(NAME = sentinel)), run_env))
  testthat::expect_error(eval(as.name(sentinel), run_env), "not found")
  assign(sentinel, TRUE, envir = run_env)
  testthat::expect_true(eval(substitute(exists(NAME), list(NAME = sentinel)), run_env))
  testthat::expect_true(eval(as.name(sentinel), run_env))
})

testthat::test_that("the installed experiment runner ignores cwd script copies", {
  testthat::skip_if(
    isTRUE(getOption("knitr.in.progress")),
    "requires installed system.file() semantics; exercised by R CMD check"
  )
  fake_dir <- tempfile("eivmixgp-fake-cwd-")
  dir.create(fake_dir)
  invisible(file.create(
    file.path(fake_dir, "01_study1_representative_figures.R")
  ))
  old_wd <- setwd(fake_dir)
  on.exit({
    setwd(old_wd)
    unlink(fake_dir, recursive = TRUE)
  }, add = TRUE)

  spec <- run_mixedgp_experiment("study1", stages = "data", dry_run = TRUE)
  installed_scripts <- normalizePath(
    system.file("experiments", package = "eivmixgp"),
    winslash = "/", mustWork = TRUE
  )
  testthat::expect_identical(spec$code_dir, installed_scripts)
})
```

## Documentation and checking

```{r}
litr::document()
# devtools::check(document = FALSE)
```
