---
title: "Creating the eivGP R package"
author: "EIV-GP authors"
date: "Version 0.3.1"
knit: litr::render
output: litr::litr_html_document
params:
  package_name: "eivGP"
  package_parent_dir: ".."
---

This is the sole active package build. The implementation is maintained in
`../codes`; the generated installable package is `../eivGP`. Numerical-study
designs, frozen data, and launchers remain in the repository experiment layer.

## Package setup

```{r package-setup, message=FALSE, results='hide'}
usethis::create_package(
  path = ".",
  fields = list(
    Package = params$package_name,
    Version = "0.3.1",
    Title = "Gaussian Process Regression with Ordinal Mixed Inputs",
    Description = paste(
      "Fits errors-in-variables Gaussian process regressions with numeric",
      "predictors, ordinal proxies, and optional calibration measurements.",
      "Uses collapsed variance inference, a finite latent-length dictionary,",
      "and no-thinning elliptical slice sampling. Supports posterior",
      "prediction, latent-input imputation, diagnostics, and continuation."
    ),
    `Authors@R` = c(
      person(given = "Penghui", family = "Fu", role = "aut"),
      person(given = "Xinye", family = "Chen", role = "aut"),
      person(given = "Jianhua Z.", family = "Huang", role = "aut"),
      person(given = "Sheng", family = "Jiang",
             email = "jiangsheng@cuhk.edu.cn", role = c("aut", "cre"))
    ),
    License = "MIT + file LICENSE",
    Encoding = "UTF-8",
    Depends = "R (>= 4.1.0)",
    Imports = "methods, parallel, posterior, stats, tools, TruncatedNormal, utils",
    Suggests = "dplyr, EzGP, ggplot2, kergp, knitr, LVGP, patchwork, testthat, tidyr"
  )
)
source_document <- knitr::current_input(dir = TRUE)
repository_dir <- normalizePath(file.path(dirname(source_document), ".."),
                                winslash = "/", mustWork = TRUE)
canonical_dir <- file.path(repository_dir, "codes")
package_dir <- usethis::proj_get()
if (!file.exists(file.path(package_dir, ".Rbuildignore"))) {
  writeLines(character(), file.path(package_dir, ".Rbuildignore"))
}
writeLines(c("YEAR: 2026", paste(
  "COPYRIGHT HOLDER: Penghui Fu, Xinye Chen, Jianhua Z. Huang, and Sheng Jiang"
)), file.path(package_dir, "LICENSE"))
stopifnot(file.copy(file.path(repository_dir, "README.md"),
                   file.path(package_dir, "README.md"), overwrite = TRUE))
stopifnot(file.copy(file.path(repository_dir, "PACKAGE-UPDATE-0.3.1.md"),
                   file.path(package_dir, "NEWS.md"), overwrite = TRUE))
dir.create(file.path(package_dir, "inst"), showWarnings = FALSE)
writeLines(c(
  "bibentry(", "  bibtype = 'Unpublished',",
  "  title = 'Ordinal Mixed-Input Gaussian Process Regression',",
  "  author = c(person('Penghui', 'Fu'), person('Xinye', 'Chen'),",
  "    person('Jianhua Z.', 'Huang'), person('Sheng', 'Jiang')),",
  "  year = '2026',", "  note = 'Manuscript in preparation'", ")"
), file.path(package_dir, "inst", "CITATION"))
```

```{package_doc package-documentation}
#' eivGP: Gaussian process regression with ordinal mixed inputs
#'
#' @import stats
#' @importFrom utils combn flush.console head tail
```

```{r global-variables, send_to="R/global-variables.R", eval=FALSE}
utils::globalVariables(c(
  "block", "chain", "ess", "global_u_accept_total", "global_u_eval_total",
  "global_u_total", "ref", "rhat", "theta_eval_total",
  "theta_update_total", "u_ess_accept_total", "u_ess_eval_total",
  "u_ess_total", "value", "xintercept"
))
```

## Canonical model modules and tests

The two historical helper filenames contain reusable model and numerical
functions. The current sampler modules follow them in a fixed load order.
Synthetic-data and experiment-runner modules are intentionally excluded.

```{r copy-canonical-modules, message=FALSE}
canonical_modules <- c(
  "00_parallel_utils.R",
  "00_study1_functions.R",
  "00_study2_functions.R",
  "00_sampler_v030.R",
  "00_sampler_v030_api.R",
  "03_study2_published_competitors.R",
  "00_public_api.R",
  "00_diagnostics.R",
  "00_mcmc_workflow.R"
)
dir.create(file.path(package_dir, "R"), showWarnings = FALSE)
stopifnot(all(file.copy(file.path(canonical_dir, canonical_modules),
  file.path(package_dir, "R", canonical_modules), overwrite = TRUE)))
write_module_order <- function() {
  description_path <- file.path(package_dir, "DESCRIPTION")
  description <- as.list(read.dcf(description_path)[1L, , drop = TRUE])
  other_modules <- setdiff(list.files(file.path(package_dir, "R"),
                                     pattern = "\\.[Rr]$"), canonical_modules)
  description$Collate <- paste(sprintf("'%s'", c(canonical_modules, other_modules)),
                               collapse = "\n")
  write.dcf(as.data.frame(description, check.names = FALSE), description_path)
}
write_module_order()
test_dir <- file.path(package_dir, "tests", "testthat")
dir.create(test_dir, recursive = TRUE, showWarnings = FALSE)
writeLines(c("library(testthat)", "library(eivGP)", "test_check('eivGP')"),
           file.path(package_dir, "tests", "testthat.R"))
stopifnot(file.copy(file.path(repository_dir, "litr", "tests", "test-utilities.R"),
                   file.path(test_dir, "test-utilities.R"), overwrite = TRUE))
stopifnot(file.copy(file.path(canonical_dir, "tests", "test_v030_core.R"),
                   file.path(test_dir, "test-v030-core.R"), overwrite = TRUE))
stopifnot(file.copy(file.path(canonical_dir, "tests", "test_v030_workflow.R"),
                   file.path(test_dir, "test-v030-workflow.R"), overwrite = TRUE))
stopifnot(file.copy(file.path(canonical_dir, "tests", "test_v031_updates.R"),
                   file.path(test_dir, "test-v031-updates.R"), overwrite = TRUE))
stopifnot(file.copy(file.path(canonical_dir, "tests", "testthat", "test-recovery-helpers.R"),
                   file.path(test_dir, "test-recovery-helpers.R"), overwrite = TRUE))
devtools::load_all(package_dir, quiet = TRUE)
```

## Documentation and checking

Rendering creates the source package; it does not launch numerical studies.
Run the package checks separately after rendering.

```{r document-package}
litr::document()
write_module_order()
```
