---
title: "Ocean validation with real observations"
output:
  html_document:
    toc: true
---

This validation uses controlled hiding of measured phosphate. It uses the
univariate threshold model in **eivGP 0.3.1**. ADNI is the separate primary
observational application with naturally missing inputs.

## Configuration

Set `OCEAN_DATA_DIR` to the folder containing `study1_data.csv` and `meta.json`.
`OCEAN_QUICK=true` is the default pipeline check; use `false` for production.
`MIXEDGP_CORES` controls the worker budget. Kernel, cache, and output options
are described in the README.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
if (!exists("OCEAN_REALDATA_DIR")) {
  input <- knitr::current_input(dir = TRUE)
  if (length(input) != 1L || !nzchar(input)) stop("Use run_application.R or knit this notebook.")
  OCEAN_REALDATA_DIR <- dirname(normalizePath(input))
}
```

## Fit, predict, and generate figures

The scripts fit representative calibration subsets and repeated nested
calibration subsets. All EIV fitting and prediction use the installed package.
Baseline comparisons use the same kernel family and test information target.
The full-training-U oracle also integrates test U given its ordinal interval.

```{r analysis}
source(file.path(OCEAN_REALDATA_DIR, "run_ocean_all.R"), local = TRUE)
```

## Reproducibility

```{r session}
writeLines(capture.output(sessionInfo()), file.path(OCEAN_OUT_PREFIX, "sessionInfo.txt"))
sessionInfo()
```
