---
title: "Raw-data provenance, preprocessing, and prepared-data checks"
output:
  html_document:
    toc: true
---

## Reproducibility status

**The original raw-to-prepared scripts are not yet available.** This document
records the recoverable processing specification and validates prepared inputs;
it does not reconstruct the frozen datasets. No cohort selection, cutoff,
record matching, or random split is invented to fill a gap.

Run the checks from the repository root with:

```sh
Rscript codes/real-data/run_application.R data
```

This executes the R chunks without fitting a model. It prints aggregate checks
only. Missing datasets are reported as unavailable; existing invalid datasets
stop with an error. Successful checks establish prepared-data consistency,
not reproduction from raw source data. Use `ADNI_DATA_DIR` and `OCEAN_DATA_DIR`
for data stored outside the repository. Raw participant-level data should remain
in authorized local storage; the shareable artifacts are processing code and
non-participant-level documentation.

```{r setup}
if (!exists("DATA_WORKFLOW_DIR")) {
  input <- knitr::current_input(dir = TRUE)
  if (length(input) != 1L || !nzchar(input)) stop("Use run_application.R data or knit this notebook.")
  DATA_WORKFLOW_DIR <- dirname(normalizePath(input))
}
```

## ADNI: raw exports to the frozen observational cohort

The current analysis and `ADNI-toledo/DATA_MANIFEST.txt` establish the output
contract below. They do not establish how raw exports were joined or filtered.

| Model role | Prepared variable | Analysis treatment |
|---|---|---|
| Response | `y_centiloid` | PET outcome supplied to the GP |
| Continuous input U | `u_csf_abeta42` | Observed when R=1; masked to NA when R=0 |
| X | `x_age_years`, `x_female`, `x_apoe4_dose` | Three predictors |
| Ordinal C | `c_ab42_ab40_3`, `c_gfap_3` | Ordered codes with levels 1, 2, 3 |
| Participant key | `RID` | Joins frozen folds to the cohort |

The final cohort has 495 participants, 133 with observed U and 362 with
naturally missing U. Three repeats each assign every participant to one of
three folds; each fold has 165 test and 330 training participants. The supplied
analysis runs repeats 2 and 3. Preserve the frozen assignments until their
original generator is recovered.

Within each fold, the current notebook sorts training and test rows by RID,
sets U to NA for R=0, constructs ordered C, and fits eivGP 0.3.1 with
`standardize_U=TRUE`. This model-time standardization uses training information;
it is separate from upstream cohort construction. At prediction, observed test
CSF is supplied for R=1 and missing CSF is integrated for R=0. Do not replace
raw CSF with the available log-CSF column: the frozen model selects the raw
`u_csf_abeta42` field.

### Original processing details still required

1. Names, release dates, checksums, and dictionaries of each raw ADNI export.
2. Participant/visit join keys; anchor-visit choice; PET, plasma, and CSF date
   windows; duplicate and tie resolution; reasons for exclusions at each step.
3. PET tracer handling and exact Centiloid conversion, including whether values
   were taken from a supplied field or computed; assay units, batches, quality
   flags, detection limits, and any biomarker transformations.
4. Coding rules for sex, APOE4 dose, and diagnosis; R definition; treatment of
   missing predictors and outcome.
5. Exact plasma ratio and GFAP cutoffs, interval boundary conventions, and the
   population used to compute any data-derived cutoffs. In particular, determine
   whether any thresholds used held-out data.
6. Frozen fold-generation code: balancing variables, algorithm, row order,
   random seed, and RNG settings. File names alone do not establish these rules.

Once recovered, the script should write the two frozen CSVs and an aggregate
exclusion flow. Reproduction should match the existing hashes when using the
same source snapshots and serialization. A deliberate revised dataset needs a
new version and separate results, rather than silently replacing the freeze.

```{r check-adni}
adni_dir <- Sys.getenv("ADNI_DATA_DIR", file.path(DATA_WORKFLOW_DIR, "ADNI-toledo", "data"))
adni_paths <- file.path(adni_dir, c("toledo_adni_cohort_n495.csv", "toledo_adni_balanced_repeated_3fold.csv"))
if (!all(file.exists(adni_paths))) {
  cat("ADNI unavailable: supply both frozen CSVs through ADNI_DATA_DIR.\n")
} else {
  expected <- c("d10b980253304620efea9f2a013166a9", "895a5759c7baf22e1771516c0aa84d70")
  stopifnot(identical(unname(tools::md5sum(adni_paths)), expected))
  cohort <- read.csv(adni_paths[1L], check.names = FALSE)
  folds <- read.csv(adni_paths[2L], check.names = FALSE)
  required <- c("RID", "R", "y_centiloid", "u_csf_abeta42", "x_age_years",
                "x_female", "x_apoe4_dose", "c_ab42_ab40_3", "c_gfap_3")
  stopifnot(all(required %in% names(cohort)), nrow(cohort) == 495L,
            ncol(cohort) == 29L, !anyNA(cohort$RID), !anyDuplicated(cohort$RID),
            all(cohort$R %in% 0:1), sum(cohort$R == 1L) == 133L,
            all(is.finite(cohort$u_csf_abeta42[cohort$R == 1L])))
  for (v in c("y_centiloid", "x_age_years", "x_female", "x_apoe4_dose"))
    stopifnot(is.numeric(cohort[[v]]), all(is.finite(cohort[[v]])))
  for (v in c("c_ab42_ab40_3", "c_gfap_3")) stopifnot(all(cohort[[v]] %in% 1:3))
  stopifnot(nrow(folds) == 1485L, all(c("RID", "repeat_id", "fold") %in% names(folds)),
            setequal(folds$repeat_id, 1:3))
  for (r in 1:3) {
    f <- folds[folds$repeat_id == r, ]
    stopifnot(nrow(f) == 495L, !anyDuplicated(f$RID), setequal(f$RID, cohort$RID),
              setequal(f$fold, 1:3), all(table(f$fold) == 165L))
  }
  cat("PASS: ADNI frozen hashes, 495 participants, 133 observed U, all repeated folds.\n")
  rm(cohort, folds)
}
```

## Ocean: source records to controlled-hiding validation

The following describes the **existing manuscript**, not an independently
reproduced pipeline. Evidence: `appendix-real-ocean.tex`, sections "Eligibility
and variable roles" and "Ordinal coarsening and calibration design", inspected
in the local manuscript snapshot `study2-design-prose/project/paper-ordinal/`
on 2026-09-10. Current loader and allocation code corroborate the input schema
and training allocation, but the raw and prepared ocean files are missing.

1. Start from BCO-DMO's AllBottle data for ECOHAB-PNW. The manuscript identifies
   `NO3_NO2` as combined nitrate plus nitrite (response) and `H2PO4` as reported
   orthophosphate (latent continuous input), with temperature and salinity as X.
2. Restrict to depth <=10 m, valid temperature/salinity, and strictly positive
   concentrations of both nutrients. The manuscript reports deduplication by
   cruise and depth identifier and 1,393 eligible records. Exact keys and tie
   handling still require the source code.
3. Compute log phosphate, centered and scaled over the eligible pool according
   to the manuscript. Preserve its center and scale in metadata. This is a
   documented pool-based construction, not a claim of training-only preprocessing.
4. Use type-1 empirical phosphate quintiles, then split the lowest quintile at
   standardized log phosphate -1.5 to produce six ordered classes. The recorded
   raw cutoffs are approximately 0.287, 0.44, 0.72, 1.22, and 1.70. Class r uses
   `(tau[r-1], tau[r]]`. Rounded values are descriptive, not regeneration inputs.
5. Freeze 100 training and 300 test rows. Recover the original selection seed,
   row identifiers/order, and sampling algorithm before regenerating this split.
6. In training, reveal exact U in nested calibration subsets of sizes 10, 25,
   and 50. The included allocation CSV has training counts (6,20,13,24,15,22).
   The existing `make_ocean_nested_calibration()` implements this downstream
   selection. Original construction thresholds do not enter the current fit.
7. Retain true U for validation, but pass only calibrated training U to fitting;
   predict held-out response given X and ordinal C without supplying test U.

### Original processing details still required

Recover the exact source accession/version and download recipe; raw-file hash,
units, missing-value sentinels and QC rules; deduplication keys/tie rule; complete
selection code and seed; unrounded cutoffs and scale constants; and prepared
CSV/metadata. Verify redistribution terms before distributing ocean package data.
The metadata must include `x_columns`, `m`, `log_tau_reference`, `log_center`,
and `log_scale`. The CSV must include `split`, `y`, `u_log_std`, `c`, and all X
columns. Preserve source row identifiers to audit selection and train/test
separation; these cannot be reconstructed from current metadata alone.

```{r check-ocean}
OCEAN_REALDATA_DIR <- DATA_WORKFLOW_DIR
source(file.path(DATA_WORKFLOW_DIR, "ocean_data_helpers.R"), local = TRUE)
ocean_dir <- Sys.getenv("OCEAN_DATA_DIR", file.path(DATA_WORKFLOW_DIR, "data", "prepared"))
if (!all(file.exists(file.path(ocean_dir, c("study1_data.csv", "meta.json"))))) {
  cat("Ocean unavailable: supply study1_data.csv and meta.json through OCEAN_DATA_DIR.\n")
} else {
  ocean <- load_ocean_prepared(ocean_dir)
  stopifnot(nrow(ocean$train) == 100L, nrow(ocean$test) == 300L,
            is.finite(ocean$log_center), is.finite(ocean$log_scale), ocean$log_scale > 0)
  allocation <- read_ocean_class_allocation(file.path(DATA_WORKFLOW_DIR, "data", "class6_proportional_calibration_allocation.csv"))
  calibration <- make_ocean_nested_calibration(ocean$c_train, allocation,
                                              c(10L, 25L, 50L), seed = 20260805L)
  stopifnot(all(calibration[["10"]] %in% calibration[["25"]]),
            all(calibration[["25"]] %in% calibration[["50"]]))
  cat("PASS: ocean loader, manuscript split sizes, scale, and nested calibration allocation.\n")
}
```

## What a complete raw-data release must contain

For each study, preserve an acquisition README, preprocessing R script with
explicit configuration, input checksums, aggregate inclusion/exclusion counts,
transformation constants, split-generation settings, prepared-output checksums,
and R session information. Run preprocessing separately from inference; an
analysis rerun should not silently rebuild cohorts or folds. Once the original
scripts are supplied, integrate them here and verify outputs against the freeze.
