---
title: "Simulation (ordinal input)"
author: "Penghui Fu, Sheng Jiang"
date: "`r Sys.Date()`"
output:
  pdf_document: 
    toc: true
    number_sections: true
header-includes:
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}} 
fontsize: 12pt
params:
  main.dir: 'script/main-code/'
  fig.dir: 'penghui-local/output/figures/'
  save.filename: 'penghui-local/output/MCMC/EIV-ordinal-cos-10.rds'
---

```{r setup, include=FALSE}
library(here)
knitr::opts_knit$set(root.dir = here())
knitr::opts_chunk$set(collapse = TRUE, message = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff=60), fig.align = 'center', fig.width = 8, fig.height = 6)
```

# Load libraries and generate data

This .rmd file runs a simulation for EIV-GP regression with an ordinal input. To begin with, load some packages and functions.

```{r}
library(mvtnorm)
library(laGP)
library(lhs)
library(GA)
library(scatterplot3d)
library(doParallel)
library(ggplot2)
source(paste0(params$main.dir, 'basics-common.R'))
source(paste0(params$main.dir, 'basics-ordinal.R'))
source(paste0(params$main.dir, 'plot-config.R'))
```

Set basic parameters for simulation.

```{r}
n <- 100  # sample size
m <- 6  # number of classes
d.x <- 1  # dimension of x

# true parameters for the data-generating process
sigma.error2.true <- 0.1^2
beta.true <- c(pi, 0)
sigma.u2.true <- (0.5*pi)^2
```

Generate other parameters and data.

```{r}
set.seed(1)
code <- 'cos'  # code for the response function
# generate x, u, c ----
x.ub <- 2
x.lb <- -2
X <- (x.ub - x.lb)*maximinLHS(n, d.x) + x.lb 
X.tilde <- cbind(1, X)
u.lb <- 0
u.ub <- 2*pi
U <- matrix(DrawTrunNormal(X.tilde%*%beta.true, sqrt(sigma.u2.true), u.lb, u.ub), n, 1)
#U <- matrix(X.tilde%*%beta.true+sqrt(sigma.u2.true)*rnorn(n), n, 1)
Tau.true <- (1:(m-1))/m*(u.ub-u.lb) + u.lb
c <- apply(U, 1, function(u) match(T, c(u-Tau.true,-Inf) <= 0))  
# generate y ----
f.true <- ResponseFunction(X, U, code)
y <- f.true + sqrt(sigma.error2.true)*rnorm(n) 
sd.y <- sd(y)
y <- y/sd.y  # standardize the response
# select observed u ----
n.obs <- floor(10)  # total number
n.obs.c <- rep(0, m)  # number for each class
n.min <- 2
for (i in 1:m) {
  if (i < m) {
    n.obs.c[i] <- max(floor(sum(c == i)/n*n.obs), n.min)
  } else {
    n.obs.c[i] <- max(n.obs - sum(n.obs.c[-m]), n.min)
  }
}
u.obs.idx <- NULL
for (i in 1:m) {
  u.obs.idx <- c(u.obs.idx, sample(c(1:n)[c == i], n.obs.c[i], FALSE))
}
u.mis.idx <- c(1:n)[-u.obs.idx]
U.obs <- U[u.obs.idx, ]
n.obs <- sum(n.obs.c)
```

# Visualization

```{r fig.cap = 'True response function', echo=FALSE}
PlotResponse(code = code, nx = 100, lb.u = min(U), ub.u = max(U), lb.x = min(X), ub.x = max(X))
```

```{r}
u_df <- data.frame(x = X, u = U, y = y*sd.y, class = as.character(c), observed = rep("FALSE", n))
u_df$observed[u.obs.idx] <- "TRUE"
```

```{r fig.cap = 'Scatterplot of y vs. u', echo=FALSE}
ggfig <- 
ggplot(u_df, aes(x = u, color = class)) + 
  geom_point(size = 3.5, aes(y = y)) + 
  scale_color_viridis_d(name = "Class") + 
  geom_function(fun = function(u) ResponseFunction(X=0, U=u, code=code), color = '#FF7F0E', linewidth = 1) +
  geom_vline(xintercept = Tau.true, linewidth = 1) + 
  coord_cartesian(xlim = c(min(U), max(U)), ylim = c(min(u_df$y), max(u_df$y))) +
  labs(color = "Class", x = 'Latent u') + 
  my_theme
#ggsave(paste0(params$fig.dir, 'Scatter-y-vs-u.pdf'), plot = ggfig, width = 8, height = 6)

ggfig <- 
ggplot(u_df[u_df$observed=="TRUE",], aes(x = u, y = y, color = class)) +
  geom_point(size = 3.5) +
  scale_color_viridis_d(name = "Class") + 
  geom_function(fun = function(u) ResponseFunction(X=0, U=u, code=code), color = '#FF7F0E', linewidth = 1) +
  coord_cartesian(xlim = c(min(U), max(U)), ylim = c(min(u_df$y), max(u_df$y))) +
  labs(color = "Class") + 
  my_theme
#ggsave(paste0(params$fig.dir, 'Scatter-y-vs-u-obs.pdf'), plot = ggfig, width = 8, height = 6)

ggfig <- 
ggplot(u_df, aes(x = c, y = y, color = class)) +
  geom_point(size = 3.5) +
  scale_color_viridis_d(name = "Class") + 
  labs(color = "Class", x = 'Class') + 
  my_theme
#ggsave(paste0(params$fig.dir, 'Scatter-y-vs-c.pdf'), plot = ggfig, width = 8, height = 6)
```

```{r fig.cap = 'Scatterplot of u versus x', echo=FALSE}
ggfig <- 
  ggplot(u_df, aes(x = X, y = u, color = class, shape = observed)) +
  geom_point(size = 3) +
  scale_color_viridis_d(name = "Class") + 
  geom_abline(slope = beta.true[2], intercept = beta.true[1], linewidth = 1, color = 'brown') +
  labs(color = "Class", shape = 'Observation') + 
  my_theme
#ggsave(paste0(params$fig.dir, 'Scatter-u-vs-x.pdf'), plot = ggfig, width = 8, height = 6)
```

```{r fig.cap = 'Barplot of c', echo=FALSE}
c_df <- data.frame(class = 1:m, value = rep(0,m))
for (i in 1:m) {
  c_df$value[i] <- sum(c == i)
}
bar_plot <- 
  ggplot(c_df, aes(x = class, y = value, fill = as.character(class))) +
  geom_bar(stat = "identity", color = "black", width = 0.7) +
  scale_fill_viridis_d(name = "Class") +  
  labs(x = "Class", y = "Count") +
  my_theme 
#ggsave(paste0(params$fig.dir, 'class-bar-plot.pdf'), plot = bar_plot, width = 8, height = 6)
```

```{r fig.cap = 'Scatterplot of y vs. x-c-u-obs', echo=FALSE}
ggfig <- Plot.Data(c, X, y, U, u.obs.idx)
ggsave(paste0(params$fig.dir, 'Scatter-y-x-c.pdf'), plot = ggfig[[1]], width = 8, height = 6)
ggsave(paste0(params$fig.dir, 'Scatter-y-x-u-obs.pdf'), plot = ggfig[[2]], width = 8, height = 6)
```

# Fix hyperparameters 

```{r, eval = FALSE}
v0 <- 1
g <- 1
IG.a.u <- 1
IG.b.u <- 1
IG.a.error <- 2.5/2
IG.b.error <- 2.5*0.084/2
```

# Obtain GP hyperparameters by MLE

Fit MLE for the GP hyperparameters using the limited observed data. We use package 'laGP' (see Gramacy's book).

```{r}
W <- cbind(X, U)
gp <- newGPsep(W[u.obs.idx,], y[u.obs.idx], d = 0.1, g = 0.01, dK = TRUE)
gp.mle <- mleGPsep(gp, param = "both", tmin = c(1e-3, 1e-3), tmax=c(1e3, 10^2))
theta.mle.obs <- 1/gp.mle$theta[1:(d.x+1)]
g.mle <- gp.mle$theta[length(gp.mle$theta)] 
sigma2.mle <- (t(y[u.obs.idx])%*%solve(KernelMatrix(W[u.obs.idx,], c(theta.mle.obs, 1)) + g.mle*diag(n.obs))%*%y[u.obs.idx]/n.obs) |> drop()
sigma.error2.mle.obs <- sigma2.mle*g.mle
rho.mle.obs <- sqrt(1/g.mle)
```

```{r echo = FALSE}
print(paste0('The MLE (from observed U) of lengthscale for x is ', theta.mle.obs[1:d.x]))
print(paste0('The MLE (from observed U) of lengthscale for u is ', theta.mle.obs[d.x+1]))
print(paste0('The MLE (from observed U) of nugget (variance) is ', sigma.error2.mle.obs))
print(paste0('The MLE (from observed U) of SN ratio is ', rho.mle.obs))
```

```{r}
# call the destructor function otherwise memory will leak.
deleteGPsep(gp)
```

(Optional) For comparison, we can also obtain the MLE of GP hyperparameters using the complete data. 

```{r, eval = FALSE}
W <- cbind(X, U)
gp <- newGPsep(W, y, d = 0.1, g = 0.01, dK = TRUE)
gp.mle <- mleGPsep(gp, param = "both", tmin = c(1e-3, 1e-3), tmax=c(1e3, var(y)))
theta.mle <- 1/gp.mle$theta[1:(d.x+1)]
g.mle <- gp.mle$theta[length(gp.mle$theta)] 
sigma2.mle <- (t(y)%*%solve(KernelMatrix(W, c(theta.mle, 1)) + g.mle*diag(n))%*%(y)/n) |> drop()
sigma.error2.mle <- sigma2.mle*g.mle
rho.mle <- sqrt(1/g.mle)
rm(list=c('W','gp','gp.mle','g.mle','sigma2.mle'))
```

```{r echo = FALSE, eval = FALSE}
print(paste0('The MLE of lengthscale for x is ', theta.mle[1:d.x]))
print(paste0('The MLE of lengthscale for u is ', theta.mle[d.x+1]))
print(paste0('The MLE of nugget (variance) is ', sigma.error2.mle))
print(paste0('The MLE of SN ratio is ', rho.mle))
```

# (Optional) Obtain GP hyperparameters by Empirical Bayes

Alternatively, we can obtain GP hyperparameters by EB, which is much slower than the MLE approach using the observed $u$.

Create candidates for GP hyperparameters (by grid).

```{r, eval = FALSE}
hyper.mat <- c() # each column of hyper.mat is a candidate of hyperparameter (theta,rho)
for (i in exp(log(2)*c(-6:1))) { # length scale for x
  for (j in exp(log(2)*c(-6:1))) { # lengthscale for u
    for(l in exp(log(2)*c(2:5))) { # signal-to-noise rho
      hyper.mat <- cbind(hyper.mat, c(rep(i, d.x), j, l))
    }
  }
}
rm(list=c('i','j','l'))
```

Generate Monte Carlo samples of $u$ given $\{x,c,u_\text{obs}\}$.

```{r, eval = FALSE}
n.mc.eb <- 10000
sigma.u20 <- 1/rgamma(1, shape = IG.a.u, rate = IG.b.u)
beta0 <- rep(0, d.x+1)
Ini.par <- GetInitial(n, m, c, u.obs.idx, U.obs, u.lb, u.ub) 
Tau0 <- Ini.par$Tau0
U0 <- Ini.par$U0
samples.eb <- Gibbs.EB(seed = 345, n.mc = n.mc.eb, 
                       beta0 = beta0, sigma.u20 = sigma.u20, Tau0 = Tau0, U0 = U0, 
                       X = X, c = c, u.obs.idx = u.obs.idx, U.obs = U.obs, u.lb = u.lb, u.ub = u.ub,
                       v0 = v0, g = g, IG.a.u = IG.a.u, IG.b.u = IG.b.u)
```

Estimate the marginal likelihood (using parallel computing).

```{r, eval = FALSE}
cl <- parallel::makeCluster(8)
doParallel::registerDoParallel(cl)
eb.evidence <- foreach(i = 1:ncol(hyper.mat), .combine = 'cbind') %dopar% {
  EstimateEvidence(theta = hyper.mat[1:(d.x+1), i], rho = hyper.mat[d.x+1+1, i],
                   U.hist = samples.eb$U.hist,
                   IG.a.error = IG.a.error, IG.b.error = IG.b.error,
                   X = X, y = y)
}
parallel::stopCluster(cl)
rm(cl)
```

Select the optimal one.

```{r, eval = FALSE}
skip <- 10
eb.evidence.mean <- apply(eb.evidence[seq(1, nrow(eb.evidence), by = skip), ], 2, mean)
eb.evidence.sd <- apply(eb.evidence[seq(1, nrow(eb.evidence), by = skip), ], 2, sd)
theta.eb <- hyper.mat[1:(d.x+1), which.max(eb.evidence.mean)]
rho.eb <- hyper.mat[d.x+1+1, which.max(eb.evidence.mean)]
```

Plot the Monte Carlo estimates of the marginal likelihood for each candidate. 

```{r echo = FALSE, eval = FALSE}
plot(log(eb.evidence.mean), 
     ylab = 'Log Evidence', 
     xlab = 'Hyper-parameter index')
     #ylim = c(min(eb.evidence.mean - 2*eb.evidence.sd/sqrt(params$n.mc)), 
     #         max(eb.evidence.mean + 2*eb.evidence.sd/sqrt(params$n.mc))))
points(which.max(eb.evidence.mean), log(max(eb.evidence.mean)), col = 'blue', pch = 16)
#segments(1:ncol(hyper.mat), eb.evidence.mean - 2*eb.evidence.sd/sqrt(params$n.mc),
#         1:ncol(hyper.mat), eb.evidence.mean + 2*eb.evidence.sd/sqrt(params$n.mc))
```

```{r echo = FALSE, eval = FALSE}
plot(eb.evidence.mean, 
     ylab = 'Evidence', 
     xlab = 'Hyper-para index',
     ylim = c(min(eb.evidence.mean - 2*eb.evidence.sd/sqrt(n.mc.eb)), 
              max(eb.evidence.mean + 2*eb.evidence.sd/sqrt(n.mc.eb))), pch = 16, cex = 0.5)
segments(1:ncol(hyper.mat), eb.evidence.mean - 2*eb.evidence.sd/sqrt(n.mc.eb),
         1:ncol(hyper.mat), eb.evidence.mean + 2*eb.evidence.sd/sqrt(n.mc.eb))
points(which.max(eb.evidence.mean), max(eb.evidence.mean), col = 'blue', pch = 16)
```

# Bayesian Inference by Gibbs

Initialize.

```{r}
set.seed(999)
beta.hist <- matrix(rep(0, d.x+1), d.x+1, 1)
sigma.u2.hist <- 1/rgamma(1, shape = IG.a.u, rate = IG.b.u)
sigma.error2.hist <- 1/rgamma(1, shape = IG.a.error, rate = IG.b.error)
Ini.par <- GetInitial(n, m, c, u.obs.idx, U.obs, u.lb, u.ub) 
Tau.hist <- matrix(Ini.par$Tau0, m-1, 1)
U.hist <- matrix(Ini.par$U0, n, 1)
accep.hist <- NULL
f.hist <- NULL

K = 1000  # number of iteration for a grand cycle
lmd <- 0.01  # step size, adjusted after each cycle 
stepsize.hist <- data.frame(stepsize = lmd, k = 1)
time0 <- proc.time()
time.hist <- data.frame(user = double(), system = double(), elapsed = double())
```

Run the Gibbs sampler.

```{r}
for (cycles in 1:60) {
  # Initialize the current cycle
  beta0 <- beta.hist[, ncol(beta.hist)]
  sigma.u20 <- sigma.u2.hist[length(sigma.u2.hist)]
  sigma.error20 <- sigma.error2.hist[length(sigma.error2.hist)]
  Tau0 <- Tau.hist[, ncol(Tau.hist)]
  U0 <- U.hist[, ncol(U.hist)]
  
  # Run Gibbs for the current cycle 
  Gibbs1 <- Gibbs(seed = NULL, num.iter = K, lmd = lmd, 
                  beta0 = beta0, sigma.u20 = sigma.u20, sigma.error20 = sigma.error20, Tau0 = Tau0, U0 = U0, 
                  X = X, c = c, y = y, u.obs.idx = u.obs.idx, U.obs = U.obs, u.lb = u.lb, u.ub = u.ub,
                  v0 = v0, g = g, IG.a.u = IG.a.u, IG.b.u = IG.b.u, IG.a.error = IG.a.error, IG.b.error = IG.b.error,
                  theta = theta.mle.obs, rho = rho.mle.obs, simu.f = FALSE, design.map = FALSE) # theta = theta.eb, rho = rho.eb
  
  # save results
  beta.hist <- cbind(beta.hist, Gibbs1$beta.hist)
  sigma.u2.hist <- c(sigma.u2.hist, Gibbs1$sigma.u2.hist)
  sigma.error2.hist <- c(sigma.error2.hist, Gibbs1$sigma.error2.hist)
  Tau.hist <- cbind(Tau.hist, Gibbs1$Tau.hist)
  U.hist <- cbind(U.hist, Gibbs1$U.hist)
  accep.hist <- c(accep.hist, Gibbs1$accep)
  f.hist <- cbind(f.hist, Gibbs1$f.hist)
  
  # adjust stepsize and record time
  if (mean(Gibbs1$accep) <= 0.2) {
    lmd <- lmd / 1.5
    stepsize.hist[nrow(stepsize.hist) + 1, ] <- c(lmd, K*cycles+1)
  }else if (mean(Gibbs1$accep) > 0.5){
    lmd <- lmd * 1.5
    stepsize.hist[nrow(stepsize.hist) + 1, ] <- c(lmd, K*cycles+1)
  }
  time.hist[cycles, ] <- c(proc.time() - time0)[1:3]
}
```

# Save results 

```{r}
saveRDS(list(X = X, y = y, sd.y = sd.y, c = c, x.ub = x.ub, x.lb = x.lb,
             v0 = v0, g = g, IG.a.u = IG.a.u, IG.b.u = IG.b.u, IG.a.error = IG.a.error, IG.b.error = IG.b.error, u.lb = u.lb, u.ub = u.ub,
             theta.mle.obs = theta.mle.obs, rho.mle.obs = rho.mle.obs,
             #theta.eb = theta.eb, rho.eb = rho.eb,
             beta.hist = beta.hist, beta.true = beta.true,
             sigma.u2.hist = sigma.u2.hist, sigma.u2.true = sigma.u2.true,
             sigma.error2.hist = sigma.error2.hist, sigma.error2.true = sigma.error2.true,
             Tau.hist = Tau.hist, Tau.true = Tau.true,
             U.hist = U.hist, U.true = U, u.obs.idx = u.obs.idx, U.obs = U.obs,
             f.hist = f.hist, f.true = f.true,
             accep.hist = accep.hist, stepsize.hist = stepsize.hist, time.hist = time.hist), 
        file = params$save.filename) 
```

Shutdown the computer.

```{r echo = FALSE}
# system('shutdown -s')
```