---
title: "Toledo-inspired ADNI EIV-GP: frozen repeated three-fold experiment"
output:
  html_document:
    toc: true
    toc_depth: 3
date: "2026-09-10"
---

This notebook evaluates a frozen observational ADNI application using eivGP
0.3.1 and the computation section's published UC-GP, LVGP, and EzGP competitors.
Cohort construction and frozen fold generation are the collaborator's separate
preprocessing task. Model comparisons use identical held-out participants.

## Scientific motivation and main-text plan

Amyloid PET measures brain amyloid burden, while CSF and plasma assays provide
related fluid biomarkers. Their relationship matters for learning about amyloid
burden when measurements are incomplete. A head-to-head ADNI biomarker study describes paired plasma, CSF, and amyloid-PET
measurements and evaluates blood tests against Alzheimer pathology
([Head-to-head comparison of leading blood tests, 2024](https://pmc.ncbi.nlm.nih.gov/articles/PMC11567821/)).
These motivate the relationship; this frozen cohort is not asserted to reproduce
those papers' cohorts or establish clinical diagnostic validity.

Here Y is PET Centiloid, U is CSF Aβ42, X comprises age, sex, and APOE4 dose,
and C contains three-level plasma Aβ42/40 and GFAP. CSF is naturally missing
for 362 of 495 participants. The ordinal inputs are constructed from continuous
assays; this study does not claim the assays are intrinsically ordinal.

The compact main text uses `main/fig1_exploration.pdf`,
`main/table1_prediction.md` (plus CSV), and `main/fig2_CSF_inference.pdf`.
All detailed fold scores, failures, paired contrasts, and validation summaries
are in `appendix/`; fold-specific diagnostics and checkpoints remain in fold
folders. The reporting script never selects the best fold, filters poor scores,
or writes an automatic claim of superiority.

The primary shared task predicts Y from test X,C with test CSF withheld for all
methods, reporting the full cohort and the naturally missing-CSF subgroup.
EIV-GP can also exploit observed *training* CSF, whereas the literature methods
use training X,C,Y. This is an explicit difference in information use, not a
controlled isolation of joint inference. Available-test-CSF prediction is an
appendix deployment scenario: EIV-GP can use it and the categorical competitors
retain their X,C predictions. The unique-to-this-comparison outputs are a
physical-CSF posterior and a CSF-coordinate response surface, not a claim that
all other statistical models lack these capabilities.

No truth is available for naturally missing CSF or the response surface. CSF
recovery is checked by hiding measured CSF in held-out participants, with neither
their U nor Y passed to the prospective imputer. A response-free LM-CSF benchmark
is reported in the appendix. This check cannot establish recovery for naturally
missing CSF. The surface figure uses prespecified fold 1, the training observed-U
10th–90th percentile range, median training age, female=0, and APOE4 dose=1;
its pointwise intervals describe association and do not imply a causal effect.

The study retains repeats 2 and 3 until the collaborator clarifies repeat 1's
role in development. Data-driven preprocessing and screening history remain
necessary qualifications; freezing data does not itself exclude leakage.

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)

## EDIT THIS ONE LINE to run repeat 3 instead of the default repeat 2.
DEFAULT_REPEAT_ID <- 2L

## Environment variables override the notebook defaults without editing it:
## ADNI_REPEAT_ID=3, EIVGP_N_CORES=4, or EIVGP_SMOKE_TEST=1.
env_int <- function(name, default, minimum = 0L) {
  raw <- Sys.getenv(name, unset = as.character(default))
  if (!grepl("^[0-9]+$", raw)) stop(name, " must be an integer.")
  x <- suppressWarnings(as.integer(raw))
  if (length(x) != 1L || is.na(x) || x < minimum) stop(name, " is invalid.")
  x
}
env_flag <- function(name, default = FALSE) {
  x <- tolower(Sys.getenv(name, unset = if (default) "true" else "false"))
  if (!x %in% c("true", "false", "1", "0", "yes", "no")) stop(name, " is invalid.")
  x %in% c("true", "1", "yes")
}

REPEAT_ID <- env_int("ADNI_REPEAT_ID", DEFAULT_REPEAT_ID, 1L)
N_CORES <- env_int("EIVGP_N_CORES", 4L, 1L)
SMOKE_TEST <- env_flag("EIVGP_SMOKE_TEST", FALSE)
stopifnot(REPEAT_ID %in% 2:3)

if (SMOKE_TEST) {
  N_CHAINS <- 4L; BURN <- 2L; INITIAL_ITER <- 8L
  EXTENSION_ITER <- 4L; CHECKPOINT_EVERY <- 3L
  FOLDS_TO_RUN <- 1L; PREDICTION_DRAWS <- 8L
} else {
  N_CHAINS <- 4L; BURN <- 4000L; INITIAL_ITER <- 20000L
  EXTENSION_ITER <- 10000L; CHECKPOINT_EVERY <- 2000L
  FOLDS_TO_RUN <- 1:3; PREDICTION_DRAWS <- 1000L
}

## Frozen sampler choice. A matched ADNI pilot found that block sizes 16 and
## 32 reduced wall time per transition but worsened missing-U mixing enough
## that they did not improve convergence per equal waiting time. Keep the
## package reference schedule for the confirmatory run.
SAMPLER_CONTROL <- list(
  u_block_size = 8L,
  cross_every = 10L,
  ess_max_steps = 10000L,
  dictionary_mode = "conditional",
  max_dictionary_size = 125L,
  gp_block_schur = TRUE,
  threshold_update = "gibbs"
)

R_HAT_LIMIT <- 1.01
BULK_ESS_LIMIT <- 400
TAIL_ESS_LIMIT <- 400
MCSE_SD_RATIO_LIMIT <- 0.10

input_path <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) "")
if (length(input_path) != 1L || is.na(input_path)) input_path <- ""
PROJECT_DIR <- Sys.getenv("ADNI_PROJECT_DIR", if (nzchar(input_path)) dirname(normalizePath(input_path)) else normalizePath(getwd()))
DATA_DIR <- Sys.getenv("ADNI_DATA_DIR", file.path(PROJECT_DIR, "data"))
DATA_FILE <- file.path(DATA_DIR, "toledo_adni_cohort_n495.csv")
FOLD_FILE <- file.path(DATA_DIR, "toledo_adni_balanced_repeated_3fold.csv")
OUTPUT_ROOT <- file.path(Sys.getenv("ADNI_OUTPUT_DIR", file.path(PROJECT_DIR, "outputs")), paste0(
  if (SMOKE_TEST) "smoke_" else "", "repeat", REPEAT_ID
))
dir.create(OUTPUT_ROOT, recursive = TRUE, showWarnings = FALSE)
RUN_LOCK <- file.path(OUTPUT_ROOT, ".run-lock")
LOCK_OWNED <- FALSE
if (!dir.create(RUN_LOCK, showWarnings = FALSE)) stop("Run already locked: ", RUN_LOCK,
  ". If a job was killed, verify it has stopped before removing this directory.")
LOCK_OWNED <- TRUE
writeLines(paste(Sys.info()[["nodename"]], Sys.getpid()), file.path(RUN_LOCK, "owner"))


## Dependency installation is a separate, explicit setup step.
EIVGP_TESTED_COMMIT <- "a3aa0f697f97d4bdf027507ac3c9b952644affb9"
if (!requireNamespace("eivGP", quietly = TRUE)) stop("Run setup.R before analysis.")
pkg <- utils::packageDescription("eivGP")
if (!identical(pkg$Version, "0.3.1")) {
  stop("eivGP version 0.3.1 is required. Run setup.R.")
}
suppressPackageStartupMessages(library(eivGP))

source(file.path(PROJECT_DIR, "adni_case_study_helpers.R"), local = TRUE)
for (d in c("main", "appendix")) dir.create(file.path(OUTPUT_ROOT, d), showWarnings = FALSE)
case_packages <- adni_case_preflight(SMOKE_TEST)
package_files <- list.files(system.file("R", package = "eivGP"), full.names = TRUE)
adni_write_csv(data.frame(file = basename(package_files), md5 = unname(tools::md5sum(package_files))),
  file.path(OUTPUT_ROOT, "appendix", "eivGP_code_fingerprints.csv"))
adni_write_csv(case_packages, file.path(OUTPUT_ROOT, "appendix", "package_versions.csv"))
adni_write_csv(data.frame(schema = ADNI_CASE_SCHEMA, smoke = SMOKE_TEST,
  prediction_draws = PREDICTION_DRAWS, repeat_id = REPEAT_ID,
  primary_target = "Y | test X,C (no test CSF)",
  competitors = paste(ADNI_METHODS, collapse = ";")),
  file.path(OUTPUT_ROOT, "appendix", "case_config.csv"))

cat("Project:", PROJECT_DIR, "\n")
cat("Repeat:", REPEAT_ID, " | folds:", paste(FOLDS_TO_RUN, collapse = ","), "\n")
cat("Mode:", if (SMOKE_TEST) "SMOKE" else "PRODUCTION", "\n")
cat("eivGP:", as.character(utils::packageVersion("eivGP")), "\n")
```

## Frozen data

```{r data}
dat <- read.csv(DATA_FILE, stringsAsFactors = FALSE, check.names = FALSE)
folds <- read.csv(FOLD_FILE, stringsAsFactors = FALSE, check.names = FALSE)
folds <- folds[folds$repeat_id == REPEAT_ID, c("RID", "fold")]

stopifnot(
  nrow(dat) == 495L,
  sum(dat$R == 1L) == 133L,
  !anyDuplicated(dat$RID),
  nrow(folds) == 495L,
  !anyDuplicated(folds$RID),
  setequal(dat$RID, folds$RID),
  setequal(unique(folds$fold), 1:3),
  unname(tools::md5sum(DATA_FILE)) == "d10b980253304620efea9f2a013166a9",
  unname(tools::md5sum(FOLD_FILE)) == "895a5759c7baf22e1771516c0aa84d70"
)

X_COLUMNS <- c("x_age_years", "x_female", "x_apoe4_dose")
C_COLUMNS <- c("c_ab42_ab40_3", "c_gfap_3")
Y_COLUMN <- "y_centiloid"
U_COLUMN <- "u_csf_abeta42"

make_C <- function(z) {
  out <- data.frame(
    c_ab42_ab40_3 = ordered(z$c_ab42_ab40_3, levels = 1:3),
    c_gfap_3 = ordered(z$c_gfap_3, levels = 1:3),
    check.names = FALSE
  )
  names(out) <- C_COLUMNS
  out
}
```

## Exploratory overview

These descriptive plots illustrate data availability and within-pattern outcome
variation. They are not model-selection criteria or evidence that a particular
method will win.

```{r exploration, fig.width=12, fig.height=4}
adni_eda(dat, OUTPUT_ROOT, SMOKE_TEST)
knitr::include_graphics(file.path(OUTPUT_ROOT, "main", "fig1_exploration.pdf"))
```

## Lightweight workflow helpers

The package does not export a trace-plot function. The helpers below plot the
stored public chain arrays directly. They do **not** call `diagnose_eivgp()`.

```{r helpers}
atomic_save_rds <- function(object, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  tmp <- tempfile(paste0(basename(path), "-"), tmpdir = dirname(path))
  on.exit(if (file.exists(tmp)) unlink(tmp), add = TRUE)
  saveRDS(object, tmp)
  if (!file.rename(tmp, path)) stop("Could not publish ", path)
}

atomic_write_lines <- function(lines, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  tmp <- tempfile(paste0(basename(path), "-"), tmpdir = dirname(path))
  on.exit(if (file.exists(tmp)) unlink(tmp), add = TRUE)
  writeLines(lines, tmp)
  if (!file.rename(tmp, path)) stop("Could not publish ", path)
}

stamp <- function(run_dir, message) {
  line <- sprintf("- [%s] %s", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"), message)
  cat(line, "\n", file = file.path(run_dir, "PROGRESS.md"), append = TRUE)
  cat(line, "\n")
  flush.console()
}

set_status <- function(run_dir, fold, stage, detail) {
  atomic_write_lines(c(
    paste0("updated_at=", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")),
    paste0("repeat=", REPEAT_ID), paste0("fold=", fold),
    paste0("stage=", stage), paste0("detail=", detail)
  ), file.path(run_dir, "CURRENT_STATUS.txt"))
  stamp(run_dir, paste("Fold", fold, "|", stage, "|", detail))
}

light_diagnostics <- function(fit, fold_dir, stage) {
  common <- c("parameter", "rhat", "ess_bulk", "ess_tail", "mcse_sd_ratio")
  parts <- list(
    hyper = subset(fit$diagnostics$rhat_hyper,
                   !grepl("^(J\\[|dictionary_occupancy)", parameter))[, common],
    loading = fit$diagnostics$rhat_A[, common],
    threshold = fit$diagnostics$rhat_tau[, common],
    missing_U = fit$diagnostics$rhat_U[, common]
  )
  tab <- do.call(rbind, lapply(names(parts), function(g) {
    z <- parts[[g]]; z$group <- g; z
  }))
  rownames(tab) <- NULL
  finite <- with(tab, is.finite(rhat) & is.finite(ess_bulk) &
                   is.finite(ess_tail) & is.finite(mcse_sd_ratio))
  passed <- all(finite) && all(tab$rhat <= R_HAT_LIMIT) &&
    all(tab$ess_bulk >= BULK_ESS_LIMIT) && all(tab$ess_tail >= TAIL_ESS_LIMIT) &&
    all(tab$mcse_sd_ratio <= MCSE_SD_RATIO_LIMIT)
  summary <- data.frame(
    stage = stage,
    n_iter = fit$diagnostics$summary$n_iter,
    saved_per_chain = fit$diagnostics$summary$saved_per_chain,
    max_rhat = if (all(finite)) max(tab$rhat) else NA_real_,
    min_bulk_ess = if (all(finite)) min(tab$ess_bulk) else NA_real_,
    min_tail_ess = if (all(finite)) min(tab$ess_tail) else NA_real_,
    max_mcse_sd_ratio = if (all(finite)) max(tab$mcse_sd_ratio) else NA_real_,
    light_gate_passed = passed,
    stringsAsFactors = FALSE
  )
  write.csv(tab, file.path(fold_dir, paste0("diagnostics_", stage, "_parameters.csv")),
            row.names = FALSE)
  write.csv(summary, file.path(fold_dir, paste0("diagnostics_", stage, "_summary.csv")),
            row.names = FALSE)
  list(table = tab, summary = summary, passed = passed)
}

trace_series <- function(fit) {
  out <- list(V = fit$mcmc$samples_by_chain$V, r = fit$mcmc$samples_by_chain$r)
  H <- fit$mcmc$samples_by_chain$logtheta
  for (j in seq_len(ncol(H[[1L]]))) {
    nm <- colnames(H[[1L]])[j]
    out[[nm]] <- lapply(H, function(x) x[, j])
  }
  A <- fit$mcmc$samples_by_chain$A
  for (j in seq_len(dim(A[[1L]])[2L]))
    out[[paste0("A[", j, ",1]")]] <- lapply(A, function(x) x[, j, 1L])
  T <- fit$mcmc$samples_by_chain$tau
  for (j in seq_len(ncol(T[[1L]])))
    out[[colnames(T[[1L]])[j]]] <- lapply(T, function(x) x[, j])
  out
}

draw_trace_panel <- function(series, path, title) {
  png(path, width = 2400, height = 1800, res = 180)
  on.exit(dev.off(), add = TRUE)
  old <- par(mfrow = c(4, 4), mar = c(2.4, 2.7, 2.1, 0.7), oma = c(0, 0, 2, 0))
  on.exit(par(old), add = TRUE)
  cols <- c("#0072B2", "#D55E00", "#009E73", "#CC79A7")
  for (nm in names(series)) {
    chains <- series[[nm]]
    n <- length(chains[[1L]])
    ids <- unique(as.integer(round(seq(1, n, length.out = min(n, 2000L)))))
    ylim <- range(unlist(lapply(chains, function(x) x[ids])), finite = TRUE)
    plot(ids, chains[[1L]][ids], type = "l", col = cols[1L], xlab = "retained draw",
         ylab = "", main = nm, ylim = ylim)
    if (length(chains) > 1L) for (cc in 2:length(chains))
      lines(ids, chains[[cc]][ids], col = cols[cc])
  }
  mtext(title, outer = TRUE, cex = 1.2)
}

draw_convergence_plots <- function(fit, diagnostic, fold_dir, stage) {
  draw_trace_panel(trace_series(fit),
                   file.path(fold_dir, paste0("trace_key_", stage, ".png")),
                   paste("Repeat", REPEAT_ID, "fold", basename(fold_dir), stage, "key traces"))

  udiag <- fit$diagnostics$rhat_U
  worst <- head(udiag$global_index[order(udiag$rhat, decreasing = TRUE)], 6L)
  U <- fit$mcmc$samples_by_chain$U
  useries <- setNames(lapply(worst, function(j) lapply(U, function(x) x[, j, 1L])),
                      paste0("U[", worst, "]"))
  draw_trace_panel(useries,
                   file.path(fold_dir, paste0("trace_worst_U_", stage, ".png")),
                   paste("Worst missing-U traces by raw R-hat:", stage))

  tab <- diagnostic$table
  png(file.path(fold_dir, paste0("rhat_overview_", stage, ".png")),
      width = 1800, height = 1100, res = 180)
  cols <- c(hyper = "#0072B2", loading = "#D55E00",
            threshold = "#009E73", missing_U = "#777777")
  plot(seq_len(nrow(tab)), tab$rhat, pch = 16, col = cols[tab$group],
       xlab = "raw parameter (grouped)", ylab = "rank-normalized split R-hat",
       main = paste("Repeat", REPEAT_ID, "fold", basename(fold_dir), stage),
       ylim = range(c(1, tab$rhat, R_HAT_LIMIT), finite = TRUE))
  abline(h = R_HAT_LIMIT, col = "red", lty = 2, lwd = 2)
  legend("topright", legend = names(cols), col = cols, pch = 16, bty = "n")
  dev.off()
}

balanced_draw_ids <- function(fit, total = 500L) {
  info <- fit$mcmc$mcmc_draw_info
  chains <- sort(unique(info$chain))
  target <- max(1L, floor(total / length(chains)))
  sort(unique(unlist(lapply(chains, function(cc) {
    a <- which(info$chain == cc)
    a[unique(as.integer(round(seq(1, length(a), length.out = min(target, length(a))))))]
  }), use.names = FALSE)))
}

score_predictions <- function(z) {
  split_z <- c(list(overall = z), split(z, paste0("R", z$R)))
  do.call(rbind, lapply(names(split_z), function(g) {
    x <- split_z[[g]]
    data.frame(
      group = g, n = nrow(x),
      RMSE = sqrt(mean((x$y_true - x$pred_mean)^2)),
      MAE = mean(abs(x$y_true - x$pred_mean)),
      Coverage95 = mean(x$y_true >= x$pred_lo95 & x$y_true <= x$pred_hi95),
      Width95 = mean(x$pred_hi95 - x$pred_lo95),
      stringsAsFactors = FALSE
    )
  }))
}
```

## Run repeat `r REPEAT_ID`

The first recoverable checkpoint is at iteration 6,000 (4,000 warm-up plus
2,000 retained transitions). Thereafter the same `fit_latest.rds` is atomically
overwritten every 2,000 transitions to save disk space. Each replacement fit
contains all earlier retained draws and a valid continuation state.

```{r run, results='asis'}
all_metrics <- list()
for (fold in FOLDS_TO_RUN) {
  fold_dir <- file.path(OUTPUT_ROOT, paste0("fold_", fold))
  dir.create(fold_dir, recursive = TRUE, showWarnings = FALSE)
  unlink(file.path(fold_dir, "CASE_STUDY_COMPLETED.txt"))
  set_status(fold_dir, fold, "prepare", "Preparing frozen train/test data.")

  test_ids <- folds$RID[folds$fold == fold]
  train <- dat[!(dat$RID %in% test_ids), , drop = FALSE]
  test <- dat[dat$RID %in% test_ids, , drop = FALSE]
  train <- train[order(train$RID), , drop = FALSE]
  test <- test[order(test$RID), , drop = FALSE]
  stopifnot(nrow(train) == 330L, nrow(test) == 165L)

  X_train <- as.matrix(train[, X_COLUMNS, drop = FALSE])
  C_train <- make_C(train)
  U_train <- matrix(train[[U_COLUMN]], ncol = 1L,
                    dimnames = list(NULL, U_COLUMN))
  U_train[train$R == 0L, 1L] <- NA_real_

  config <- data.frame(
    repeat_id = REPEAT_ID, fold = fold,
    data_md5 = unname(tools::md5sum(DATA_FILE)),
    folds_md5 = unname(tools::md5sum(FOLD_FILE)),
    n_train = nrow(train), n_test = nrow(test),
    n_train_observed_U = sum(train$R == 1L),
    n_train_missing_U = sum(train$R == 0L),
    engine = "multivariate", latent_dim = 1L, ident = "none",
    kernel = "se", standardize_U = TRUE,
    n_chains = N_CHAINS, burn = BURN, initial_iter = INITIAL_ITER,
    extension_iter = EXTENSION_ITER, checkpoint_every = CHECKPOINT_EVERY,
    u_block_size = SAMPLER_CONTROL$u_block_size,
    cross_every = SAMPLER_CONTROL$cross_every,
    gp_block_schur = SAMPLER_CONTROL$gp_block_schur,
    dictionary_mode = SAMPLER_CONTROL$dictionary_mode,
    max_dictionary_size = SAMPLER_CONTROL$max_dictionary_size,
    ess_max_steps = SAMPLER_CONTROL$ess_max_steps,
    threshold_update = SAMPLER_CONTROL$threshold_update,
    eivgp_version = as.character(utils::packageVersion("eivGP")),
    tested_commit = EIVGP_TESTED_COMMIT,
    installed_commit = if (is.null(pkg$RemoteSha)) "unrecorded" else pkg$RemoteSha,
    stringsAsFactors = FALSE
  )
  config_path <- file.path(fold_dir, "run_config.csv")
  if (file.exists(file.path(fold_dir, "fit_latest.rds"))) {
    if (!file.exists(config_path)) stop("Checkpoint has no run configuration.")
    old_config <- read.csv(config_path, stringsAsFactors = FALSE)
    if (!isTRUE(all.equal(old_config, config, check.attributes = FALSE)))
      stop("Checkpoint configuration mismatch; use a separate output directory.")
  }
  write.csv(config, config_path, row.names = FALSE)

  fit_path <- file.path(fold_dir, "fit_latest.rds")
  if (file.exists(fit_path)) {
    fit <- readRDS(fit_path)
    stopifnot(
      identical(fit$sampler_version, "0.3.1"),
      identical(fit$control$u_block_size, SAMPLER_CONTROL$u_block_size),
      identical(fit$control$cross_every, SAMPLER_CONTROL$cross_every),
      fit$diagnostics$summary$burn == BURN
    )
    stamp(fold_dir, paste("Loaded checkpoint at iteration", fit$diagnostics$summary$n_iter))
  } else {
    first_target <- min(INITIAL_ITER, BURN + CHECKPOINT_EVERY)
    set_status(fold_dir, fold, "fit", paste("Initial fit to iteration", first_target))
    started <- Sys.time()
    fit <- fit_eivgp(
      X = X_train, y = train[[Y_COLUMN]], C = C_train, U_obs = U_train,
      engine = "multivariate", latent_dim = 1L, ident = "none",
      standardize_U = TRUE, store_scores = FALSE, kernel = "se",
      parallel = TRUE, n_cores = N_CORES, n_chains = N_CHAINS,
      n_iter = first_target, burn = BURN, thin = 1L,
      seed = 202620000L + 100L * REPEAT_ID + fold,
      verbose = FALSE, sampler_control = SAMPLER_CONTROL
    )
    atomic_save_rds(fit, fit_path)
    elapsed <- as.numeric(difftime(Sys.time(), started, units = "secs"))
    stamp(fold_dir, sprintf("Saved checkpoint at %d; segment %.1f min.",
                            first_target, elapsed / 60))
  }

  extend_to <- function(fit, target) {
    while (fit$diagnostics$summary$n_iter < target) {
      current <- fit$diagnostics$summary$n_iter
      add <- min(CHECKPOINT_EVERY, target - current)
      set_status(fold_dir, fold, "continue",
                 sprintf("Iteration %d -> %d; checkpoint will overwrite fit_latest.rds.",
                         current, current + add))
      started <- Sys.time()
      fit <- continue_eivgp(fit, n_iter = add, parallel = TRUE,
                            n_cores = N_CORES, verbose = FALSE)
      atomic_save_rds(fit, fit_path)
      elapsed <- as.numeric(difftime(Sys.time(), started, units = "secs"))
      remaining <- target - fit$diagnostics$summary$n_iter
      eta <- if (remaining > 0L) elapsed / add * remaining / 60 else 0
      stamp(fold_dir, sprintf(
        "Saved iteration %d; last %d took %.1f min; ETA to %d about %.1f min.",
        fit$diagnostics$summary$n_iter, add, elapsed / 60, target, eta
      ))
    }
    fit
  }

  fit <- extend_to(fit, INITIAL_ITER)
  current_iter <- fit$diagnostics$summary$n_iter
  final_stage <- paste0("iter", current_iter)
  diagnostic <- light_diagnostics(fit, fold_dir, final_stage)
  draw_convergence_plots(fit, diagnostic, fold_dir, final_stage)
  if (!diagnostic$passed && current_iter < INITIAL_ITER + EXTENSION_ITER) {
    fit <- extend_to(fit, INITIAL_ITER + EXTENSION_ITER)
    final_stage <- paste0("iter", fit$diagnostics$summary$n_iter)
    diagnostic <- light_diagnostics(fit, fold_dir, final_stage)
    draw_convergence_plots(fit, diagnostic, fold_dir, final_stage)
  }
  stamp(fold_dir, paste(final_stage, "convergence gate:", diagnostic$passed))

  set_status(fold_dir, fold, "prediction", paste("Predicting from", final_stage, "fit."))
  X_test <- as.matrix(test[, X_COLUMNS, drop = FALSE])
  C_test <- make_C(test)
  U_test <- matrix(test[[U_COLUMN]], ncol = 1L,
                   dimnames = list(NULL, U_COLUMN))
  U_test[test$R == 0L, 1L] <- NA_real_
  draw_ids <- balanced_draw_ids(fit, PREDICTION_DRAWS)
  response_draws <- predict_eivgp(
    fit, new_X = X_test, new_C = C_test, new_U = U_test,
    new_U_scale = "raw", target = "response", draw_ids = draw_ids,
    seed = 202630000L + 100L * REPEAT_ID + fold
  )
  pred <- data.frame(
    repeat_id = REPEAT_ID, fold = fold, final_stage = final_stage,
    convergence_passed = diagnostic$passed,
    RID = test$RID, diagnosis = test$diagnosis, R = test$R,
    y_true = test[[Y_COLUMN]], pred_mean = colMeans(response_draws),
    pred_sd = apply(response_draws, 2L, stats::sd),
    pred_lo95 = apply(response_draws, 2L, stats::quantile, probs = 0.025),
    pred_hi95 = apply(response_draws, 2L, stats::quantile, probs = 0.975),
    n_prediction_draws = nrow(response_draws), stringsAsFactors = FALSE
  )
  write.csv(pred, file.path(fold_dir, "eivgp_predictions.csv"), row.names = FALSE)
  metrics <- score_predictions(pred)
  metrics$convergence_passed <- diagnostic$passed
  metrics$repeat_id <- REPEAT_ID; metrics$fold <- fold; metrics$final_stage <- final_stage
  write.csv(metrics, file.path(fold_dir, "eivgp_metrics.csv"), row.names = FALSE)
  all_metrics[[as.character(fold)]] <- metrics
  atomic_write_lines(c(
    paste0("completed_at=", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")),
    paste0("repeat=", REPEAT_ID), paste0("fold=", fold),
    paste0("convergence_passed=", diagnostic$passed),
    paste0("final_stage=", final_stage)
  ), file.path(fold_dir, "EIV_COMPLETED.txt"))
  set_status(fold_dir, fold, "EIV_complete", paste("EIV prediction completed at", final_stage))
  cat("\n### Fold", fold, "\n\n")
  print(knitr::kable(metrics, digits = 3))
  set_status(fold_dir, fold, "comparison", "Running shared prediction and CSF inference tasks.")
  adni_extended_fold(fit, train, test, draw_ids, fold_dir, REPEAT_ID, fold,
                     SMOKE_TEST, diagnostic$passed, response_draws)
  atomic_write_lines(c(paste0("completed_at=", Sys.time()),
    paste0("schema=", ADNI_CASE_SCHEMA), paste0("convergence_passed=", diagnostic$passed)),
    file.path(fold_dir, "CASE_STUDY_COMPLETED.txt"))
  set_status(fold_dir, fold, "complete", "EIV fit, comparison, and additional inference finished; inspect method status.")
  rm(fit, response_draws)
  invisible(gc())
}

combined_metrics <- do.call(rbind, all_metrics)
write.csv(combined_metrics, file.path(OUTPUT_ROOT, "all_fold_metrics.csv"), row.names = FALSE)
```

## Compact main results and appendix export

Published competitors follow the numerical/computation section: UC-GP
(Qian et al., 2008; Zhou et al., 2011) via kergp, LVGP (Zhang et al., 2020)
via LVGP, and EzGP (Xiao et al., 2021) via EzGP. The installed eivGP 0.3.1 public adapters
use their default production controls and training-only tuning. Exact package
code fingerprints and versions are recorded because 0.3.1 development builds
can differ in optimizer rescue schedules.
Failed fits are listed, not replaced by substitute implementations. SMOKE
runs use explicitly reduced budgets to exercise code, never to rank methods.

The primary table uses the intersection of folds with valid predictions from
all four methods. Incomplete intersections are labelled provisional; per-method
available-fold results are separately tabulated. MCMC warnings are retained.
Scores are pooled over out-of-fold participants within each repeat. No independent
replication standard errors are computed from folds or repeated participants.

```{r report, results='asis'}
main_table <- adni_case_report(OUTPUT_ROOT, FOLDS_TO_RUN, SMOKE_TEST)
print(knitr::kable(main_table, digits = 3))
```

## Session information

```{r session-info}
writeLines(capture.output(sessionInfo()), file.path(OUTPUT_ROOT, "sessionInfo.txt"))
sessionInfo()
unlink(RUN_LOCK, recursive = TRUE)
```
