---
title: "Prepare the Toledo-inspired ADNI cohort and frozen 3 x 3 folds"
output:
  html_document:
    toc: true
    toc_depth: 3
date: "2026-09-11"
---

## Purpose and data-use boundary

This notebook reconstructs the participant-level inputs used by the
Toledo-inspired ADNI EIV-GP analysis. It creates, **only in a local secure
directory**:

1. `toledo_adni_cohort_n495.csv`;
2. `toledo_adni_balanced_repeated_3fold.csv`;
3. aggregate validation and provenance files.

The generated CSV files contain ADNI participant-level or derived
participant-level data. Each investigator who
runs this notebook must have approved ADNI access and must comply with the
[ADNI Data Use Agreement](https://adni.loni.usc.edu/wp-content/themes/adni_2023/documents/ADNI_Data_Use_Agreement.pdf).


## Raw ADNI files required

Download the following tables through the investigator's own approved
[LONI IDA/ADNI account](https://adni.loni.usc.edu/data-samples/adni-data/).
The displayed download filename may have a date suffix. The notebook searches
recursively by the stable table name, so a file such as
`UCBERKELEY_AMY_6MM_01Aug2026.csv` is accepted.

| Stable table name | ADNI area / description | Columns used |
|---|---|---|
| `UPENN_PLASMA_FUJIREBIO_QUANTERIX` | Biofluid Biomarkers; UPENN plasma Aβ42, Aβ40, p-tau217, NfL and GFAP measured by Fujirebio/Quanterix | `RID`, `EXAMDATE`, `pT217_F`, `pT217_AB42_F`, `AB42_AB40_F`, `GFAP_Q` |
| `UCBERKELEY_AMY_6MM` | PET numerical data; UC Berkeley amyloid PET 6 mm analysis | `RID`, `SCANDATE`, `CENTILOIDS`, `SUMMARY_SUVR`, `TRACER`, `qc_flag` |
| `DXSUM` | Clinical Assessments; diagnostic summary | `RID`, `EXAMDATE`, `DIAGNOSIS` |
| `PTDEMOG` | Subject Characteristics; participant demographics | `RID`, `VISDATE`, `PTGENDER`, `PTEDUCAT`, `PTDOB` |
| `APOERES` | Genetics / Biospecimen; ApoE genotyping results | `RID`, `GENOTYPE` |
| `UPENNBIOMK_ROCHE_ELECSYS` | Biofluid Biomarkers; UPENN CSF Roche Elecsys biomarkers | `RID`, `EXAMDATE`, `ABETA42`, `BATCH` |

For the exact frozen experiment, use the same ADNI snapshot used in the
original analysis (the PET, diagnosis, and demographics exports were dated
01-Aug-2026). ADNI tables are updated over time. A newer snapshot can change
the eligible cohort and therefore cannot silently replace the frozen
confirmatory cohort.

The preferred input is the original CSV exported from IDA. For local backward
compatibility, this notebook can also read a matching `.rda` file or a matching
CSV stored inside a ZIP archive. Those compatibility options do not authorize
redistribution of the data.

## Configuration

Before rendering, set two environment variables. Example:

```r
Sys.setenv(
  ADNI_RAW_DIR = "/secure/path/to/my/adni-downloads",
  ADNI_TOLEDO_DATA_DIR = "/secure/path/to/my/adni-derived/toledo"
)
rmarkdown::render("prepare_adni_toledo_cohort.Rmd")
```

Alternatively, place the raw downloads in a local `raw-data/` directory beside
this Rmd. The default output is `local-data/` beside this Rmd. Both directory
names must be excluded from Git.

Set `ADNI_STRICT_FROZEN=0` only for an explicitly labelled cohort-update
sensitivity analysis. Such output is not the frozen n=495 experiment.

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE,
  results = "markup", error = FALSE
)

input_path <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) "")
PROJECT_DIR <- if (nzchar(input_path)) {
  dirname(normalizePath(input_path))
} else {
  normalizePath(getwd())
}

env_flag <- function(name, default = FALSE) {
  value <- tolower(Sys.getenv(name, unset = if (default) "true" else "false"))
  if (!value %in% c("true", "false", "1", "0", "yes", "no")) {
    stop(name, " must be true/false, yes/no, or 1/0.")
  }
  value %in% c("true", "1", "yes")
}

RAW_DIR <- Sys.getenv(
  "ADNI_RAW_DIR", unset = file.path(PROJECT_DIR, "raw-data")
)
OUTPUT_DIR <- Sys.getenv(
  "ADNI_TOLEDO_DATA_DIR", unset = file.path(PROJECT_DIR, "local-data")
)
STRICT_FROZEN <- env_flag("ADNI_STRICT_FROZEN", TRUE)

if (!dir.exists(RAW_DIR)) {
  stop(
    "ADNI raw-data directory not found: ", RAW_DIR, "\n",
    "Set ADNI_RAW_DIR to the investigator's secure local ADNI download directory."
  )
}
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
RAW_DIR <- normalizePath(RAW_DIR)
OUTPUT_DIR <- normalizePath(OUTPUT_DIR)

cat("Raw directory:", RAW_DIR, "\n")
cat("Local derived-data directory:", OUTPUT_DIR, "\n")
cat("Strict frozen reproduction:", STRICT_FROZEN, "\n")
```

## Input discovery and validation helpers

The environment-variable overrides below are optional. They are useful if the
raw directory contains multiple versions of the same ADNI table:

```text
ADNI_PLASMA_FILE
ADNI_PET_FILE
ADNI_DIAGNOSIS_FILE
ADNI_DEMOGRAPHICS_FILE
ADNI_APOE_FILE
ADNI_CSF_FILE
```

Each override should point to the exact local CSV, RDA, or ZIP file. When a ZIP
is supplied, the notebook selects the member beginning with the corresponding
stable table name.

```{r input-helpers}
escape_regex <- function(x) {
  gsub("([][{}()+*^$|\\\\?.])", "\\\\\\1", x)
}

stable_file_pattern <- function(stem, extension = "csv|rda") {
  ## Match the stable IDA name or the usual DDMonYYYY export suffix. This
  ## deliberately rejects project-derived files such as *_baseline_selected.csv.
  paste0(
    "^", escape_regex(stem),
    "(?:_[0-9]{2}[A-Za-z]{3}[0-9]{4})?\\.(", extension, ")$"
  )
}

zip_candidate <- function(zip_path, stem) {
  listing <- utils::unzip(zip_path, list = TRUE)$Name
  hit <- listing[grepl(
    stable_file_pattern(stem, "csv"), basename(listing), ignore.case = TRUE
  )]
  if (!length(hit)) return(NULL)
  if (length(hit) > 1L) {
    stop(
      "ZIP contains multiple candidates for ", stem, ": ", zip_path, "\n",
      paste(hit, collapse = "\n")
    )
  }
  list(path = zip_path, member = hit[[1L]], kind = "zip_csv")
}

discover_source <- function(stem, override_env) {
  override <- Sys.getenv(override_env, unset = "")
  if (nzchar(override)) {
    if (!file.exists(override)) stop(override_env, " does not exist: ", override)
    override <- normalizePath(override)
    if (grepl("\\.zip$", override, ignore.case = TRUE)) {
      answer <- zip_candidate(override, stem)
      if (is.null(answer)) stop("No ", stem, " CSV found inside ", override)
      return(answer)
    }
    kind <- if (grepl("\\.csv$", override, ignore.case = TRUE)) {
      "csv"
    } else if (grepl("\\.rda$", override, ignore.case = TRUE)) {
      "rda"
    } else {
      stop(override_env, " must identify a CSV, RDA, or ZIP file.")
    }
    return(list(path = override, member = NA_character_, kind = kind))
  }

  files <- list.files(RAW_DIR, recursive = TRUE, full.names = TRUE)
  direct_csv <- files[
    grepl(stable_file_pattern(stem, "csv"), basename(files), ignore.case = TRUE)
  ]
  direct_rda <- files[
    grepl(stable_file_pattern(stem, "rda"), basename(files), ignore.case = TRUE)
  ]
  zip_files <- files[grepl("\\.zip$", files, ignore.case = TRUE)]
  inside_zip <- Filter(Negate(is.null), lapply(zip_files, zip_candidate, stem = stem))

  candidates <- c(
    lapply(direct_csv, function(x) list(path = x, member = NA_character_, kind = "csv")),
    inside_zip,
    lapply(direct_rda, function(x) list(path = x, member = NA_character_, kind = "rda"))
  )
  if (!length(candidates)) {
    stop(
      "Could not find ADNI table ", stem, " under ", RAW_DIR, ".\n",
      "Download it from IDA or set ", override_env, " to its exact path."
    )
  }

  ## Prefer CSV over compatibility RDA. If there is more than one candidate at
  ## the best available level, require an explicit file override instead of
  ## silently mixing ADNI snapshots.
  rank <- vapply(candidates, function(x) {
    switch(x$kind, csv = 1L, zip_csv = 2L, rda = 3L)
  }, integer(1))
  candidates <- candidates[rank == min(rank)]
  if (length(candidates) > 1L) {
    labels <- vapply(candidates, function(x) {
      paste0(x$path, if (!is.na(x$member)) paste0("::", x$member) else "")
    }, character(1))
    stop(
      "Multiple equally preferred sources found for ", stem, ".\n",
      "Set ", override_env, " to one exact file:\n", paste(labels, collapse = "\n")
    )
  }
  candidates[[1L]]
}

read_source <- function(source, stem) {
  if (source$kind == "csv") {
    out <- utils::read.csv(
      source$path, stringsAsFactors = FALSE, check.names = FALSE
    )
  } else if (source$kind == "zip_csv") {
    out <- utils::read.csv(
      unz(source$path, source$member),
      stringsAsFactors = FALSE, check.names = FALSE
    )
  } else {
    env <- new.env(parent = emptyenv())
    loaded <- load(source$path, envir = env)
    object_name <- if (stem %in% loaded) stem else loaded[[1L]]
    out <- as.data.frame(env[[object_name]], stringsAsFactors = FALSE)
  }
  out
}

require_columns <- function(data, columns, label) {
  missing <- setdiff(columns, names(data))
  if (length(missing)) {
    stop(label, " is missing required columns: ", paste(missing, collapse = ", "))
  }
  invisible(data)
}

parse_adni_date <- function(x) {
  if (inherits(x, "Date")) return(x)
  x <- trimws(as.character(x))
  x[x %in% c("", "NA", "N/A")] <- NA_character_
  out <- as.Date(rep(NA_character_, length(x)))
  for (fmt in c("%Y-%m-%d", "%m/%d/%Y", "%m/%d/%y")) {
    use <- is.na(out) & !is.na(x)
    out[use] <- as.Date(x[use], format = fmt)
  }
  out
}

parse_month_year_midpoint <- function(x) {
  if (inherits(x, "Date")) return(x)
  x <- trimws(as.character(x))
  out <- as.Date(rep(NA_character_, length(x)))
  mm_yyyy <- grepl("^[0-9]{1,2}/[0-9]{4}$", x)
  out[mm_yyyy] <- as.Date(paste0("15/", x[mm_yyyy]), format = "%d/%m/%Y")
  yyyy_mm <- grepl("^[0-9]{4}-[0-9]{1,2}$", x)
  out[yyyy_mm] <- as.Date(paste0(x[yyyy_mm], "-15"), format = "%Y-%m-%d")
  full <- is.na(out) & !is.na(x)
  out[full] <- parse_adni_date(x[full])
  out
}

as_numeric <- function(x) suppressWarnings(as.numeric(as.character(x)))

nearest_match <- function(anchor, measure, anchor_date = "anchor_date",
                          measure_date = "measure_date", lower = -183L,
                          upper = 183L, label, value_names = "value") {
  paired <- merge(
    anchor[, c("row_id", "RID", anchor_date)], measure, by = "RID"
  )
  paired$delta_days <- as.numeric(paired[[measure_date]] - paired[[anchor_date]])
  paired <- paired[
    paired$delta_days >= lower & paired$delta_days <= upper,
    , drop = FALSE
  ]
  paired <- paired[
    order(paired$row_id, abs(paired$delta_days), paired[[measure_date]]),
    , drop = FALSE
  ]
  paired <- paired[
    !duplicated(paired$row_id),
    c("row_id", measure_date, value_names, "delta_days"),
    drop = FALSE
  ]
  names(paired)[2L] <- paste0(label, "_date")
  names(paired)[match(value_names, names(paired))] <- paste0(label, "_", value_names)
  names(paired)[ncol(paired)] <- paste0(label, "_delta_days")
  paired
}

atomic_write_csv <- function(data, path) {
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  tmp <- tempfile(paste0(basename(path), "-"), tmpdir = dirname(path))
  on.exit(if (file.exists(tmp)) unlink(tmp), add = TRUE)
  utils::write.csv(data, tmp, row.names = FALSE)
  if (!file.rename(tmp, path)) stop("Could not publish ", path)
  invisible(path)
}
```

## Load the six authorized raw tables

No participant rows are printed in this report.

```{r load-inputs}
specs <- list(
  plasma = list(
    stem = "UPENN_PLASMA_FUJIREBIO_QUANTERIX",
    env = "ADNI_PLASMA_FILE",
    columns = c("RID", "EXAMDATE", "pT217_F", "pT217_AB42_F",
                "AB42_AB40_F", "GFAP_Q")
  ),
  pet = list(
    stem = "UCBERKELEY_AMY_6MM", env = "ADNI_PET_FILE",
    columns = c("RID", "SCANDATE", "CENTILOIDS", "SUMMARY_SUVR",
                "TRACER", "qc_flag")
  ),
  diagnosis = list(
    stem = "DXSUM", env = "ADNI_DIAGNOSIS_FILE",
    columns = c("RID", "EXAMDATE", "DIAGNOSIS")
  ),
  demographics = list(
    stem = "PTDEMOG", env = "ADNI_DEMOGRAPHICS_FILE",
    columns = c("RID", "VISDATE", "PTGENDER", "PTEDUCAT", "PTDOB")
  ),
  apoe = list(
    stem = "APOERES", env = "ADNI_APOE_FILE",
    columns = c("RID", "GENOTYPE")
  ),
  csf = list(
    stem = "UPENNBIOMK_ROCHE_ELECSYS", env = "ADNI_CSF_FILE",
    columns = c("RID", "EXAMDATE", "ABETA42", "BATCH")
  )
)

sources <- lapply(specs, function(x) discover_source(x$stem, x$env))
tables <- Map(function(source, spec) {
  data <- read_source(source, spec$stem)
  require_columns(data, spec$columns, spec$stem)
  data
}, sources, specs)

source_manifest <- do.call(rbind, lapply(names(sources), function(name) {
  source <- sources[[name]]
  info <- file.info(source$path)
  data.frame(
    role = name,
    stable_table = specs[[name]]$stem,
    local_file = basename(source$path),
    archive_member = if (is.na(source$member)) "" else source$member,
    file_bytes = info$size,
    file_md5 = unname(tools::md5sum(source$path)),
    rows_read = nrow(tables[[name]]),
    stringsAsFactors = FALSE
  )
}))
atomic_write_csv(
  source_manifest,
  file.path(OUTPUT_DIR, "toledo_adni_raw_source_manifest_LOCAL_ONLY.csv")
)

cat("Validated six raw ADNI tables.\n")
cat("Raw rows by table:\n")
print(source_manifest[, c("stable_table", "rows_read")], row.names = FALSE)
```

## Build the Toledo analysis cohort

The construction order reproduces the original frozen analysis:

1. use the plasma visit as the temporal anchor;
2. match the nearest QC-passed amyloid PET scan within +/-183 days;
3. match the nearest diagnosis to PET within +/-183 days;
4. attach demographics and APOE4 dose;
5. retain the closest eligible plasma-PET landmark per RID **before** reading
   CSF availability;
6. define observed `U` as Roche Elecsys CSF Aβ42 within +/-30 days of PET;
7. retain florbetapir (`FBP`) PET only;
8. derive plasma Aβ42/Aβ40 and GFAP tertiles from the complete n=495 cohort
   without using `Y`, `U`, or `R`.

```{r construct-cohort}
plasma0 <- tables$plasma
anchor <- data.frame(
  RID = as.integer(as_numeric(plasma0$RID)),
  anchor_date = parse_adni_date(plasma0$EXAMDATE),
  plasma_ptau217 = as_numeric(plasma0$pT217_F),
  plasma_ptau217_ab42 = as_numeric(plasma0$pT217_AB42_F),
  stringsAsFactors = FALSE
)
anchor <- anchor[
  complete.cases(anchor) & anchor$plasma_ptau217 > 0 &
    anchor$plasma_ptau217_ab42 > 0,
  , drop = FALSE
]
anchor$c_ptau217_3 <- as.integer(cut(
  anchor$plasma_ptau217,
  breaks = c(-Inf, 0.128, 0.299, Inf),
  labels = 1:3, ordered_result = TRUE
))
anchor$c_ptau217_ab42_3 <- as.integer(cut(
  anchor$plasma_ptau217_ab42,
  breaks = c(-Inf, 0.0055, 0.0086, Inf),
  labels = 1:3, ordered_result = TRUE
))
anchor$row_id <- seq_len(nrow(anchor))

## Keep these two continuous plasma measurements for the final ordinal proxies.
plasma_proxy <- data.frame(
  RID = as.integer(as_numeric(plasma0$RID)),
  anchor_date = parse_adni_date(plasma0$EXAMDATE),
  plasma_ab42_ab40 = as_numeric(plasma0$AB42_AB40_F),
  plasma_gfap = as_numeric(plasma0$GFAP_Q),
  stringsAsFactors = FALSE
)
plasma_proxy <- plasma_proxy[
  !duplicated(plasma_proxy[, c("RID", "anchor_date")]),
  , drop = FALSE
]

pet0 <- tables$pet
pet <- data.frame(
  RID = as.integer(as_numeric(pet0$RID)),
  measure_date = parse_adni_date(pet0$SCANDATE),
  centiloid = as_numeric(pet0$CENTILOIDS),
  summary_suvr = as_numeric(pet0$SUMMARY_SUVR),
  tracer = as.character(pet0$TRACER),
  qc_flag = as.integer(as_numeric(pet0$qc_flag)),
  stringsAsFactors = FALSE
)
pet <- pet[
  pet$qc_flag == 2L & is.finite(pet$centiloid) & !is.na(pet$measure_date),
  , drop = FALSE
]
pet_match <- nearest_match(
  anchor, pet, label = "pet",
  value_names = c("centiloid", "summary_suvr", "tracer")
)
parent <- merge(anchor, pet_match, by = "row_id")
names(parent)[names(parent) == "pet_centiloid"] <- "y_centiloid"
names(parent)[names(parent) == "pet_summary_suvr"] <- "y_summary_suvr"
names(parent)[names(parent) == "pet_tracer"] <- "tracer"

diagnosis0 <- tables$diagnosis
diagnosis <- data.frame(
  RID = as.integer(as_numeric(diagnosis0$RID)),
  measure_date = parse_adni_date(diagnosis0$EXAMDATE),
  value = as.integer(as_numeric(diagnosis0$DIAGNOSIS)),
  stringsAsFactors = FALSE
)
diagnosis <- diagnosis[
  diagnosis$value %in% 1:3 & !is.na(diagnosis$measure_date),
  , drop = FALSE
]
diagnosis_match <- nearest_match(
  parent, diagnosis, anchor_date = "pet_date", label = "diagnosis"
)
parent <- merge(parent, diagnosis_match, by = "row_id")
names(parent)[names(parent) == "diagnosis_value"] <- "diagnosis_code"

demographics0 <- tables$demographics
demographics <- data.frame(
  RID = as.integer(as_numeric(demographics0$RID)),
  VISDATE = parse_adni_date(demographics0$VISDATE),
  PTGENDER = as.integer(as_numeric(demographics0$PTGENDER)),
  PTEDUCAT = as_numeric(demographics0$PTEDUCAT),
  PTDOB = as.character(demographics0$PTDOB),
  stringsAsFactors = FALSE
)
demographics <- demographics[
  demographics$PTGENDER %in% c(1L, 2L) &
    is.finite(demographics$PTEDUCAT) & !is.na(demographics$PTDOB),
  , drop = FALSE
]
demographics <- demographics[order(demographics$RID, demographics$VISDATE), ]
demographics <- demographics[
  !duplicated(demographics$RID),
  c("RID", "PTGENDER", "PTEDUCAT", "PTDOB"),
  drop = FALSE
]
parent <- merge(parent, demographics, by = "RID")
birth_midmonth <- parse_month_year_midpoint(parent$PTDOB)
parent$x_age_years <- as.numeric(parent$pet_date - birth_midmonth) / 365.25
parent$x_female <- as.integer(parent$PTGENDER == 2L)
parent$x_education_years <- parent$PTEDUCAT

apoe0 <- tables$apoe
apoe <- data.frame(
  RID = as.integer(as_numeric(apoe0$RID)),
  GENOTYPE = as.character(apoe0$GENOTYPE),
  stringsAsFactors = FALSE
)
apoe <- apoe[!is.na(apoe$GENOTYPE) & nzchar(apoe$GENOTYPE), , drop = FALSE]
apoe <- apoe[!duplicated(apoe$RID), , drop = FALSE]
apoe$x_apoe4_dose <- lengths(regmatches(
  apoe$GENOTYPE, gregexpr("4", apoe$GENOTYPE)
))
parent <- merge(parent, apoe, by = "RID")

required <- c(
  "y_centiloid", "x_age_years", "x_female", "x_apoe4_dose",
  "c_ptau217_3", "c_ptau217_ab42_3", "diagnosis_code"
)
parent <- parent[
  apply(parent[, required, drop = FALSE], 1L, function(x) all(is.finite(x))),
  , drop = FALSE
]

## This is the important no-selection-on-U step: choose one landmark per RID
## before CSF is read or R is defined.
parent <- parent[
  order(parent$RID, abs(parent$pet_delta_days), parent$pet_date),
  , drop = FALSE
]
parent <- parent[!duplicated(parent$RID), , drop = FALSE]
parent$row_id <- seq_len(nrow(parent))

csf0 <- tables$csf
csf <- data.frame(
  RID = as.integer(as_numeric(csf0$RID)),
  measure_date = parse_adni_date(csf0$EXAMDATE),
  abeta42 = as_numeric(csf0$ABETA42),
  batch = as.character(csf0$BATCH),
  stringsAsFactors = FALSE
)
csf <- csf[
  is.finite(csf$abeta42) & csf$abeta42 > 0 & !is.na(csf$measure_date),
  , drop = FALSE
]
csf_match <- nearest_match(
  parent, csf, anchor_date = "pet_date", lower = -30L, upper = 30L,
  label = "u_csf", value_names = c("abeta42", "batch")
)
cohort <- merge(parent, csf_match, by = "row_id", all.x = TRUE)
cohort$u_log_csf_abeta42 <- log(cohort$u_csf_abeta42)
cohort$R <- as.integer(is.finite(cohort$u_csf_abeta42))
cohort$tracer_fbp <- as.integer(toupper(cohort$tracer) == "FBP")

keep <- c(
  "RID", "anchor_date", "pet_date", "tracer", "tracer_fbp",
  "y_centiloid", "y_summary_suvr", "R", "u_csf_abeta42",
  "u_log_csf_abeta42", "u_csf_batch", "c_ptau217_3",
  "c_ptau217_ab42_3", "plasma_ptau217", "plasma_ptau217_ab42",
  "x_age_years", "x_female", "x_education_years", "x_apoe4_dose",
  "diagnosis_code", "GENOTYPE", "pet_delta_days", "u_csf_date",
  "u_csf_delta_days"
)
cohort <- cohort[order(cohort$RID), keep]
cohort <- cohort[cohort$tracer_fbp == 1L, , drop = FALSE]
cohort$diagnosis <- c(`1` = "CN", `2` = "MCI", `3` = "Dementia")[
  as.character(cohort$diagnosis_code)
]

cohort <- merge(
  cohort, plasma_proxy, by = c("RID", "anchor_date"), all.x = TRUE,
  sort = FALSE
)
cohort <- cohort[order(cohort$RID), , drop = FALSE]
if (any(!is.finite(cohort$plasma_ab42_ab40)) ||
    any(!is.finite(cohort$plasma_gfap)) ||
    any(cohort$plasma_ab42_ab40 <= 0) || any(cohort$plasma_gfap <= 0)) {
  stop("Eligible cohort contains missing/invalid plasma Aβ42/Aβ40 or GFAP values.")
}

## Exploratory ordered measurements: complete-cohort marginal tertiles only.
## Y, U, and R do not enter the cutpoint calculation.
ab42_ab40_cutpoints <- unname(stats::quantile(
  cohort$plasma_ab42_ab40, c(1 / 3, 2 / 3)
))
gfap_cutpoints <- unname(stats::quantile(
  cohort$plasma_gfap, c(1 / 3, 2 / 3)
))
cohort$c_ab42_ab40_3 <- as.integer(cut(
  cohort$plasma_ab42_ab40,
  c(-Inf, ab42_ab40_cutpoints, Inf),
  labels = 1:3, ordered_result = TRUE
))
cohort$c_gfap_3 <- as.integer(cut(
  cohort$plasma_gfap,
  c(-Inf, gfap_cutpoints, Inf),
  labels = 1:3, ordered_result = TRUE
))

## Preserve the byte-level schema of the frozen file. In the historical
## two-stage pipeline anchor_date was re-parsed as Date, whereas pet_date and
## u_csf_date remained ISO character columns after an intermediate CSV read.
cohort$pet_date <- as.character(cohort$pet_date)
cohort$u_csf_date <- as.character(cohort$u_csf_date)

cohort_file <- file.path(OUTPUT_DIR, "toledo_adni_cohort_n495.csv")
atomic_write_csv(cohort, cohort_file)

cohort_summary <- data.frame(
  n = nrow(cohort),
  observed_U = sum(cohort$R == 1L),
  missing_U = sum(cohort$R == 0L),
  missing_fraction = mean(cohort$R == 0L),
  ab42_ab40_tertile_1 = ab42_ab40_cutpoints[[1L]],
  ab42_ab40_tertile_2 = ab42_ab40_cutpoints[[2L]],
  gfap_tertile_1 = gfap_cutpoints[[1L]],
  gfap_tertile_2 = gfap_cutpoints[[2L]],
  cohort_md5 = unname(tools::md5sum(cohort_file)),
  stringsAsFactors = FALSE
)

if (STRICT_FROZEN) {
  stopifnot(
    nrow(cohort) == 495L,
    !anyDuplicated(cohort$RID),
    sum(cohort$R == 1L) == 133L,
    all.equal(ab42_ab40_cutpoints, c(0.0782, 0.0883333333333333),
              tolerance = 1e-12) == TRUE,
    all.equal(gfap_cutpoints, c(127.966666666667, 215.9),
              tolerance = 1e-10) == TRUE
  )
}

cat("Cohort constructed. Aggregate summary only:\n")
print(cohort_summary, row.names = FALSE)
```

## Construct the frozen balanced 3 repeats x 3 folds

This is the exact balance-only procedure used for the current `balanced-v2`
split. It never fits a prediction model and never selects a split based on
held-out predictive performance.

Each fold has exactly 165 participants. Balance targets include diagnosis,
natural U availability, outcome range, sex, APOE4 dose, the two-proxy joint
cells, age, observed-U range, and coarse observed-U/amyloid-PET curve regions.
The use of `Y` and observed `U` here is only for allocation balance; it does not
change cohort membership and does not evaluate a fitted model.

```{r construct-folds}
cohort <- cohort[order(cohort$RID), , drop = FALSE]
n <- nrow(cohort)
r1 <- cohort$R == 1L
r0 <- !r1

bin_rank <- function(x, k) {
  pmin(k, pmax(1L, ceiling(rank(x, ties.method = "average") / length(x) * k)))
}
one_hot <- function(x, prefix) {
  matrix <- stats::model.matrix(~ factor(x) - 1L)
  colnames(matrix) <- paste0(prefix, seq_len(ncol(matrix)))
  matrix
}
zscore <- function(x) as.numeric(scale(x))

blocks <- list()
block_weights <- list()
add_block <- function(x, weight) {
  blocks[[length(blocks) + 1L]] <<- as.matrix(x)
  block_weights[[length(block_weights) + 1L]] <<- rep(
    weight, ncol(as.matrix(x))
  )
}

add_block(one_hot(cohort$diagnosis_code, "diagnosis_"), 3)
add_block(one_hot(interaction(cohort$diagnosis_code, cohort$R), "diagnosis_R_"), 3)
add_block(one_hot(bin_rank(cohort$y_centiloid, 5L), "Yq5_"), 3)
add_block(one_hot(cohort$x_female, "sex_"), 1.5)
add_block(one_hot(cohort$x_apoe4_dose, "apoe_"), 2)
add_block(one_hot(interaction(
  cohort$c_ab42_ab40_3, cohort$c_gfap_3
), "Ccell_"), 2)
add_block(cbind(
  age_z = zscore(cohort$x_age_years),
  Y_z = zscore(cohort$y_centiloid),
  Y_z2 = zscore(cohort$y_centiloid)^2
), 1)

u_q5 <- integer(n)
u_q5[r1] <- bin_rank(cohort$u_csf_abeta42[r1], 5L)
y_q3_r1 <- integer(n)
y_q3_r1[r1] <- bin_rank(cohort$y_centiloid[r1], 3L)
u_q3 <- integer(n)
u_q3[r1] <- bin_rank(cohort$u_csf_abeta42[r1], 3L)

r1_uq <- matrix(0, n, 5L)
r1_uq[cbind(which(r1), u_q5[r1])] <- 1
colnames(r1_uq) <- paste0("R1_Uq5_", 1:5)
add_block(r1_uq, 5)

r1_curve <- matrix(0, n, 9L)
curve_id <- (u_q3[r1] - 1L) * 3L + y_q3_r1[r1]
r1_curve[cbind(which(r1), curve_id)] <- 1
colnames(r1_curve) <- paste0("R1_UxY_region_", 1:9)
add_block(r1_curve, 4)

cell_id <- (cohort$c_ab42_ab40_3 - 1L) * 3L + cohort$c_gfap_3
r1_cell <- matrix(0, n, 9L)
r1_cell[cbind(which(r1), cell_id[r1])] <- 1
colnames(r1_cell) <- paste0("R1_Ccell_", 1:9)
add_block(r1_cell, 6)

r0_cell <- matrix(0, n, 9L)
r0_cell[cbind(which(r0), cell_id[r0])] <- 1
colnames(r0_cell) <- paste0("R0_Ccell_", 1:9)
add_block(r0_cell, 3)

u_z <- numeric(n)
u_z[r1] <- zscore(cohort$u_csf_abeta42[r1])
y_r1_z <- numeric(n)
y_r1_z[r1] <- zscore(cohort$y_centiloid[r1])
add_block(cbind(
  R1_U_z = u_z, R1_U_z2 = u_z^2,
  R1_Y_z = y_r1_z, R1_UxY = u_z * y_r1_z
), 2)

feature <- do.call(cbind, blocks)
weight <- unlist(block_weights, use.names = FALSE)
target <- colSums(feature) / 3
feature_scale <- pmax(
  1,
  sqrt(colSums((feature - rep(colMeans(feature), each = n))^2) / 3)
)
contribution <- function(sums) {
  sum(weight * ((sums - target) / feature_scale)^2)
}

make_initial <- function(seed, repeat_id) {
  set.seed(seed)
  assignment <- integer(n)
  r1_counts <- c(45L, 44L, 44L)[
    ((seq_len(3L) + repeat_id - 2L) %% 3L) + 1L
  ]
  r0_counts <- 165L - r1_counts
  assignment[sample(which(r1))] <- rep(1:3, r1_counts)
  assignment[sample(which(r0))] <- rep(1:3, r0_counts)
  assignment
}

improve <- function(assignment, seed, attempts = 100000L) {
  set.seed(seed)
  sums <- do.call(rbind, lapply(1:3, function(k) {
    colSums(feature[assignment == k, , drop = FALSE])
  }))
  score <- sum(vapply(1:3, function(k) contribution(sums[k, ]), numeric(1)))
  best <- assignment
  best_score <- score
  pools <- list(which(r0), which(r1))

  for (step in seq_len(attempts)) {
    pool <- pools[[sample.int(2L, 1L)]]
    pair <- sample(pool, 2L)
    a <- assignment[pair[1L]]
    b <- assignment[pair[2L]]
    if (a == b) next
    old <- contribution(sums[a, ]) + contribution(sums[b, ])
    delta <- feature[pair[2L], ] - feature[pair[1L], ]
    new_a <- sums[a, ] + delta
    new_b <- sums[b, ] - delta
    new <- contribution(new_a) + contribution(new_b)
    temperature <- 0.02 * (1 - step / attempts)
    if (new <= old || stats::runif(1) < exp(
      (old - new) / max(temperature, 1e-8)
    )) {
      assignment[pair] <- assignment[rev(pair)]
      sums[a, ] <- new_a
      sums[b, ] <- new_b
      score <- score + new - old
      if (score < best_score) {
        best <- assignment
        best_score <- score
      }
    }
  }
  list(fold = best, score = best_score)
}

adjusted_rand <- function(a, b) {
  table_ab <- table(a, b)
  choose2 <- function(x) x * (x - 1) / 2
  index <- sum(choose2(table_ab))
  row_sum <- sum(choose2(rowSums(table_ab)))
  column_sum <- sum(choose2(colSums(table_ab)))
  total <- choose2(length(a))
  expected <- row_sum * column_sum / total
  (index - expected) / ((row_sum + column_sum) / 2 - expected)
}

RNGkind("Mersenne-Twister", "Inversion", "Rejection")
candidate_pool <- vector("list", 3L)
for (repeat_id in 1:3) {
  candidates <- vector("list", 12L)
  for (candidate_id in seq_along(candidates)) {
    seed <- 202609170L + 10000L * repeat_id + candidate_id
    candidates[[candidate_id]] <- improve(
      make_initial(seed, repeat_id), seed + 500000L
    )
  }
  candidate_pool[[repeat_id]] <- candidates
}

selected <- vector("list", 3L)
scores_1 <- vapply(candidate_pool[[1L]], `[[`, numeric(1), "score")
selected[[1L]] <- candidate_pool[[1L]][[which.min(scores_1)]]
for (repeat_id in 2:3) {
  scores <- vapply(candidate_pool[[repeat_id]], `[[`, numeric(1), "score")
  eligible <- which(scores <= min(scores) * 1.10)
  diversity <- vapply(eligible, function(candidate_id) {
    max(vapply(selected[seq_len(repeat_id - 1L)], function(old) {
      abs(adjusted_rand(
        old$fold, candidate_pool[[repeat_id]][[candidate_id]]$fold
      ))
    }, numeric(1)))
  }, numeric(1))
  selected[[repeat_id]] <- candidate_pool[[repeat_id]][[
    eligible[which.min(diversity)]
  ]]
}

folds <- do.call(rbind, lapply(1:3, function(repeat_id) {
  data.frame(
    repeat_id = repeat_id,
    RID = cohort$RID,
    fold = selected[[repeat_id]]$fold
  )
}))
folds <- folds[order(folds$repeat_id, folds$RID), , drop = FALSE]

fold_audit <- list()
audit_index <- 0L
for (repeat_id in 1:3) for (fold_id in 1:3) {
  ids <- folds$RID[folds$repeat_id == repeat_id & folds$fold == fold_id]
  test <- cohort[cohort$RID %in% ids, , drop = FALSE]
  train_r1 <- cohort[!cohort$RID %in% ids & cohort$R == 1L, , drop = FALSE]
  test_r1 <- test[test$R == 1L, , drop = FALSE]
  train_cell <- table(factor(
    cell_id[match(train_r1$RID, cohort$RID)], levels = 1:9
  ))
  audit_index <- audit_index + 1L
  fold_audit[[audit_index]] <- data.frame(
    repeat_id = repeat_id,
    fold = fold_id,
    test_n = nrow(test),
    test_R1 = sum(test$R == 1L),
    train_R1 = nrow(train_r1),
    age_SMD = (mean(test$x_age_years) - mean(cohort$x_age_years)) /
      stats::sd(cohort$x_age_years),
    Y_SMD = (mean(test$y_centiloid) - mean(cohort$y_centiloid)) /
      stats::sd(cohort$y_centiloid),
    U_SMD_R1 = (mean(test_r1$u_csf_abeta42) -
                  mean(cohort$u_csf_abeta42[r1])) /
      stats::sd(cohort$u_csf_abeta42[r1]),
    min_train_R1_Ccell = min(train_cell),
    stringsAsFactors = FALSE
  )
}
fold_audit <- do.call(rbind, fold_audit)

repeat_diversity <- data.frame(
  repeat_a = c(1L, 1L, 2L),
  repeat_b = c(2L, 3L, 3L),
  adjusted_rand = c(
    adjusted_rand(selected[[1L]]$fold, selected[[2L]]$fold),
    adjusted_rand(selected[[1L]]$fold, selected[[3L]]$fold),
    adjusted_rand(selected[[2L]]$fold, selected[[3L]]$fold)
  )
)

fold_file <- file.path(
  OUTPUT_DIR, "toledo_adni_balanced_repeated_3fold.csv"
)
atomic_write_csv(folds, fold_file)
atomic_write_csv(
  fold_audit,
  file.path(OUTPUT_DIR, "toledo_adni_fold_balance_summary_LOCAL_ONLY.csv")
)
atomic_write_csv(
  repeat_diversity,
  file.path(OUTPUT_DIR, "toledo_adni_repeat_diversity_LOCAL_ONLY.csv")
)

stopifnot(
  nrow(folds) == 3L * nrow(cohort),
  !anyDuplicated(folds[, c("repeat_id", "RID")]),
  all(table(folds$repeat_id) == nrow(cohort)),
  setequal(folds$fold, 1:3)
)
if (STRICT_FROZEN) {
  stopifnot(
    nrow(folds) == 1485L,
    all(table(folds$repeat_id, folds$fold) == 165L),
    all(fold_audit$test_R1 %in% 44:45),
    all(fold_audit$train_R1 %in% 88:89),
    all(fold_audit$min_train_R1_Ccell >= 2L),
    max(abs(repeat_diversity$adjusted_rand)) < 0.01
  )
}

cat("Frozen fold construction completed. Aggregate audit only:\n")
print(fold_audit, row.names = FALSE)
cat("\nCross-repeat adjusted Rand indices:\n")
print(repeat_diversity, row.names = FALSE)
cat("\nFold file MD5:", unname(tools::md5sum(fold_file)), "\n")
```

## Final frozen-reproduction check and local manifest

For the 01-Aug-2026 source snapshot, the expected output hashes are included as
reproducibility checks. A hash mismatch is not repaired by editing the cohort or
fold files manually. First check raw-table versions, column definitions, locale,
and R version.

```{r final-check}
EXPECTED_COHORT_MD5 <- "d10b980253304620efea9f2a013166a9"
EXPECTED_FOLD_MD5 <- "895a5759c7baf22e1771516c0aa84d70"
cohort_md5 <- unname(tools::md5sum(cohort_file))
fold_md5 <- unname(tools::md5sum(fold_file))

if (STRICT_FROZEN && cohort_md5 != EXPECTED_COHORT_MD5) {
  stop(
    "The cohort passed structural checks but does not reproduce the frozen ",
    "cohort byte-for-byte. Expected ", EXPECTED_COHORT_MD5,
    "; obtained ", cohort_md5, ". Check the raw ADNI snapshot."
  )
}
if (STRICT_FROZEN && fold_md5 != EXPECTED_FOLD_MD5) {
  stop(
    "The folds passed structural checks but do not reproduce the frozen split ",
    "byte-for-byte. Expected ", EXPECTED_FOLD_MD5,
    "; obtained ", fold_md5, ". Do not use these folds for the frozen run."
  )
}

manifest_lines <- c(
  paste0("created_at=", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")),
  paste0("strict_frozen=", STRICT_FROZEN),
  "cohort_rule=plasma anchor; nearest QC-pass amyloid PET within 183 days; one closest FBP landmark per RID before CSF exposure",
  "observed_U_rule=nearest Roche Elecsys CSF ABETA42 within 30 days of PET",
  "proxy_rule=complete-cohort marginal tertiles; no Y/U/R in cutpoint calculation",
  "fold_rule=balanced-v2; 3 repeats x 3 folds; balance/support only; no fitted-model performance",
  paste0("cohort_n=", nrow(cohort)),
  paste0("observed_U_n=", sum(cohort$R == 1L)),
  paste0("cohort_md5=", cohort_md5),
  paste0("fold_md5=", fold_md5),
  paste0("R_version=", R.version.string)
)
writeLines(
  manifest_lines,
  file.path(OUTPUT_DIR, "toledo_adni_preprocessing_manifest_LOCAL_ONLY.txt")
)

cat("Frozen preprocessing reproduced successfully.\n")
cat("Cohort:", cohort_file, "\n")
cat("Folds:", fold_file, "\n")
cat("Cohort MD5:", cohort_md5, "\n")
cat("Fold MD5:", fold_md5, "\n")
cat("Do not upload the generated files to Git or GitHub.\n")
```

## Session information

The report records the local software environment but never prints participant
rows.

```{r session-info}
sessionInfo()
```
