---
title: "5Study1_helper"
output: html_document
---

#Function for Study I

Including:

- General Utility

- Slice Sampler

- Data Generating Process

- GP likelihood and prediction

- EIVGP Sampler

- MCMC diagnostics

- Main EIVGP fitting function

- Baseline GP model

- Predictive scoring

## General Utilities

```{r, send_to = 'R/Study_1_Helper.R'}
#' Load or install a package
#'
#' Checks whether a package is installed. If not, installs it.
#' Then loads the package with suppressed startup messages.
#'
#' @param pkg Character string, name of the package.
#' @return Invisibly loads the package.
#' @export
load_or_install <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

#' Safe maximum
#'
#' Returns the maximum of a numeric vector, or `NA_real_` if the vector
#' is empty or contains no finite values.
#'
#' @param x Numeric vector.
#' @return A single numeric value, the maximum, or `NA_real_`.
#' @export
safe_max <- function(x) {
  if (length(x) == 0 || all(!is.finite(x))) return(NA_real_)
  max(x, na.rm = TRUE)
}

#' Safe minimum
#'
#' Returns the minimum of a numeric vector, or `NA_real_` if the vector
#' is empty or contains no finite values.
#'
#' @param x Numeric vector.
#' @return A single numeric value, the minimum, or `NA_real_`.
#' @export
safe_min <- function(x) {
  if (length(x) == 0 || all(!is.finite(x))) return(NA_real_)
  min(x, na.rm = TRUE)
}

#' Safe median
#'
#' Returns the median of a numeric vector, or `NA_real_` if the vector
#' is empty or contains no finite values.
#'
#' @param x Numeric vector.
#' @return A single numeric value, the median, or `NA_real_`.
#' @export
safe_median <- function(x) {
  if (length(x) == 0 || all(!is.finite(x))) return(NA_real_)
  median(x, na.rm = TRUE)
}

#' Maximin Latin hypercube sample (1D)
#'
#' Generates a one-dimensional Latin hypercube sample of size `n`
#' using a random shift and random permutation to maximise the minimum
#' distance between points.
#'
#' @param n Integer, sample size.
#' @param lower Numeric, lower bound of the interval (default -2).
#' @param upper Numeric, upper bound of the interval (default 2).
#' @return Numeric vector of length `n`.
#' @export
maximin_lhs_1d <- function(n, lower = -2, upper = 2) {
  z <- (seq_len(n) - runif(n)) / n
  z <- sample(z)
  lower + (upper - lower) * z
}

#' Classify continuous values
#'
#' Discretises a numeric vector `u` into classes according to
#' thresholds `tau`. Values less than or equal to the first threshold
#' belong to class 1, those between the first and second to class 2,
#' and so on.
#'
#' @param u Numeric vector.
#' @param tau Numeric vector of increasing thresholds.
#' @return Integer vector of class labels (1, 2, ...).
#' @export
make_class <- function(u, tau) {
  as.integer(cut(u, breaks = c(-Inf, tau, Inf), labels = FALSE))
}

#' Vectorised truncated normal random generation
#'
#' Draws random samples from a normal distribution truncated to
#' the interval [`lower`, `upper`], with element-wise parameters.
#'
#' @param mean Numeric vector of means.
#' @param sd Numeric vector of standard deviations (positive).
#' @param lower Numeric vector of lower truncation points.
#' @param upper Numeric vector of upper truncation points.
#' @return Numeric vector of the same length as the longest input.
#' @export
rtruncnorm_vec <- function(mean, sd, lower, upper) {
  n <- max(length(mean), length(lower), length(upper))
  mean <- rep(mean, length.out = n)
  sd <- rep(sd, length.out = n)
  lower <- rep(lower, length.out = n)
  upper <- rep(upper, length.out = n)
  
  pl <- pnorm((lower - mean) / sd)
  pu <- pnorm((upper - mean) / sd)
  
  pl <- pmax(pl, 0)
  pu <- pmin(pu, 1)
  
  width <- pmax(pu - pl, .Machine$double.eps)
  uu <- pl + runif(n) * width
  uu <- pmin(pmax(uu, .Machine$double.eps), 1 - .Machine$double.eps)
  
  mean + sd * qnorm(uu)
}

#' Pairwise squared Euclidean distances
#'
#' Computes the matrix of squared Euclidean distances between
#' the rows of `a` and `b`. If `b` is `NULL`, returns the
#' pairwise distances within `a`.
#'
#' @param a Numeric matrix or data frame.
#' @param b Optional second matrix or data frame.
#' @return A numeric matrix of squared distances.
#' @export
pairwise_sqdist <- function(a, b = NULL) {
  a <- as.matrix(a)
  if (is.null(b)) b <- a
  b <- as.matrix(b)
  
  aa <- rowSums(a^2)
  bb <- rowSums(b^2)
  
  out <- outer(aa, bb, "+") - 2 * tcrossprod(a, b)
  pmax(out, 0)
}

#' Safe Cholesky decomposition
#'
#' Computes the Cholesky factor of a symmetric positive
#' (semi)definite matrix, adding increasing amounts of jitter
#' if the decomposition fails numerically.
#'
#' @param A A symmetric matrix.
#' @param jitter Base amount of jitter to add (default `1e-8`).
#' @return Upper triangular Cholesky factor.
#' @export
safe_chol <- function(A, jitter = 1e-8) {
  n <- nrow(A)
  for (k in 0:8) {
    jj <- jitter * 10^k
    ans <- try(chol(A + jj * diag(n)), silent = TRUE)
    if (!inherits(ans, "try-error")) return(ans)
  }
  stop("Cholesky decomposition failed.")
}

#' Solve linear system from Cholesky factor
#'
#' Solves `A x = b` given the upper triangular Cholesky factor
#' `U` of `A`.
#'
#' @param U Upper triangular Cholesky factor.
#' @param b Right-hand side vector or matrix.
#' @return Solution `x`.
#' @export
solve_chol <- function(U, b) {
  backsolve(U, forwardsolve(t(U), b))
}
```

## Slice Sampler
```{r, send_to = 'R/Study_1_Helper.R'}
#' Bounded slice sampler update
#'
#' Performs a single slice sampling step with stepping-out and
#' shrinkage on a bounded interval. The slice is constructed
#' around the current point `x0` using the given log-density
#' function `logf`.
#'
#' @param x0 Numeric scalar, current state.
#' @param logf Function that returns the log-density at a point.
#' @param w Estimated typical slice width (default 1).
#' @param lower Lower bound (default `-Inf`).
#' @param upper Upper bound (default `Inf`).
#' @param max_steps_out Maximum number of stepping-out steps (default 50).
#' @param max_iter Maximum iterations for the shrinkage phase (default 200).
#'
#' @return A list with components `x` (new state) and `n_eval`
#'   (number of log-density evaluations).
#' @export
bounded_slice_update <- function(x0, logf, w = 1,
                                 lower = -Inf, upper = Inf,
                                 max_steps_out = 50,
                                 max_iter = 200) {
  if (upper <= lower) {
    return(list(x = x0, n_eval = 0))
  }
  
  eps <- 1e-12
  if (is.finite(lower)) x0 <- max(x0, lower + eps)
  if (is.finite(upper)) x0 <- min(x0, upper - eps)
  
  f0 <- logf(x0)
  n_eval <- 1L
  
  if (!is.finite(f0)) {
    return(list(x = x0, n_eval = n_eval))
  }
  
  logy <- f0 + log(runif(1))
  
  L <- x0 - runif(1) * w
  R <- L + w
  
  if (is.finite(lower)) L <- max(L, lower)
  if (is.finite(upper)) R <- min(R, upper)
  
  J <- floor(runif(1) * max_steps_out)
  K <- max_steps_out - 1L - J
  
  while (J > 0 && (!is.finite(lower) || L > lower)) {
    fL <- logf(L)
    n_eval <- n_eval + 1L
    if (!is.finite(fL) || fL <= logy) break
    L <- L - w
    if (is.finite(lower)) L <- max(L, lower)
    J <- J - 1L
  }
  
  while (K > 0 && (!is.finite(upper) || R < upper)) {
    fR <- logf(R)
    n_eval <- n_eval + 1L
    if (!is.finite(fR) || fR <= logy) break
    R <- R + w
    if (is.finite(upper)) R <- min(R, upper)
    K <- K - 1L
  }
  
  for (iter in seq_len(max_iter)) {
    x1 <- runif(1, L, R)
    f1 <- logf(x1)
    n_eval <- n_eval + 1L
    
    if (is.finite(f1) && f1 >= logy) {
      return(list(x = x1, n_eval = n_eval))
    }
    
    if (x1 < x0) {
      L <- x1
    } else {
      R <- x1
    }
  }
  
  list(x = x0, n_eval = n_eval)
}

#' Full-interval slice sampler update
#'
#' Performs a single slice sampling step using the full interval
#' `[lower, upper]` as the initial slice. Both `lower` and `upper`
#' must be finite. The algorithm shrinks the interval toward the
#' current point until a valid sample is found.
#'
#' @param x0 Numeric scalar, current state.
#' @param logf Function that returns the log-density at a point.
#' @param lower Finite lower bound.
#' @param upper Finite upper bound.
#' @param max_iter Maximum iterations for the shrinkage phase (default 200).
#'
#' @return A list with components `x` (new state) and `n_eval`
#'   (number of log-density evaluations). Stops if `lower` or `upper`
#'   is not finite.
#' @export
full_interval_slice_update <- function(x0, logf, lower, upper,
                                       max_iter = 200) {
  if (!is.finite(lower) || !is.finite(upper)) {
    stop("full_interval_slice_update requires finite lower and upper bounds.")
  }
  
  if (upper <= lower) {
    return(list(x = x0, n_eval = 0))
  }
  
  eps <- 1e-12
  lower_eps <- lower + eps
  upper_eps <- upper - eps
  
  if (upper_eps <= lower_eps) {
    return(list(x = x0, n_eval = 0))
  }
  
  x0 <- min(max(x0, lower_eps), upper_eps)
  
  f0 <- logf(x0)
  n_eval <- 1L
  
  if (!is.finite(f0)) {
    return(list(x = x0, n_eval = n_eval))
  }
  
  logy <- f0 + log(runif(1))
  L <- lower_eps
  R <- upper_eps
  
  for (iter in seq_len(max_iter)) {
    x1 <- runif(1, L, R)
    f1 <- logf(x1)
    n_eval <- n_eval + 1L
    
    if (is.finite(f1) && f1 >= logy) {
      return(list(x = x1, n_eval = n_eval))
    }
    
    if (x1 < x0) {
      L <- x1
    } else {
      R <- x1
    }
  }
  
  list(x = x0, n_eval = n_eval)
}
```

## Data Generating Process
```{r, send_to = 'R/Study_1_Helper.R'}
############################################################
## Data-generating mechanism
############################################################

#' True regression function (1D simulation)
#'
#' Computes the regression function for the one-dimensional simulation
#' study. Under the *active* scenario the function depends on both
#' `x` and `u`; under the *inactive* scenario it depends only on `u`.
#'
#' @param x Numeric vector, covariate.
#' @param u Numeric vector, latent variable.
#' @param scenario Character, either `"active"` or `"inactive"`.
#' @return Numeric vector of function values.
#' @export
f0_1d <- function(x, u, scenario = c("active", "inactive")) {
  scenario <- match.arg(scenario)
  
  if (scenario == "inactive") {
    return(cos(0.9 * pi * u))
  }
  
  0.5 * sin(pi * x / 2) +
    cos(0.9 * pi * u) +
    0.35 * (x / 2) * sin(0.9 * pi * u)
}

#' Simulate 1D heterogeneous data
#'
#' Generates training and test data according to the one-dimensional
#' heterogeneous regression model. A Latin hypercube design is used for
#' the training covariate `x`; the latent variable `u` is drawn from a
#' standard normal and discretised into `m` classes using true
#' thresholds `tau_true`. The function repeatedly samples until every
#' class contains at least 3 training points.
#'
#' @param n Integer, number of training observations (default 100).
#' @param n_test Integer, number of test observations (default 500).
#' @param m Integer, number of latent classes (default 6).
#' @param scenario Character, `"active"` or `"inactive"`.
#' @param sigma_eps Numeric, residual standard deviation (default 0.1).
#' @param seed Optional integer for random seed.
#'
#' @return A list with components `train` (data frame), `test`
#'   (data frame), `tau_true`, `sigma_eps`, `scenario`, and `m`.
#' @export
simulate_1d_data <- function(n = 100,
                             n_test = 500,
                             m = 6,
                             scenario = c("active", "inactive"),
                             sigma_eps = 0.1,
                             seed = NULL) {
  scenario <- match.arg(scenario)
  
  if (!is.null(seed)) {
    set.seed(seed)
  }
  
  tau_true <- c(-5 / 3, -10 / 9, 0, 10 / 9, 5 / 3)
  
  repeat {
    x <- maximin_lhs_1d(n, lower = -2, upper = 2)
    u <- rnorm(n)
    c_ord <- make_class(u, tau_true)
    
    if (all(tabulate(c_ord, nbins = m) >= 3)) break
  }
  
  f <- f0_1d(x, u, scenario = scenario)
  y <- f + rnorm(n, mean = 0, sd = sigma_eps)
  
  x_test <- runif(n_test, -2, 2)
  u_test <- rnorm(n_test)
  c_test <- make_class(u_test, tau_true)
  f_test <- f0_1d(x_test, u_test, scenario = scenario)
  y_test <- f_test + rnorm(n_test, mean = 0, sd = sigma_eps)
  
  list(
    train = data.frame(
      x = x,
      u = u,
      c = c_ord,
      y = y,
      f = f
    ),
    test = data.frame(
      x = x_test,
      u = u_test,
      c = c_test,
      y = y_test,
      f = f_test
    ),
    tau_true = tau_true,
    sigma_eps = sigma_eps,
    scenario = scenario,
    m = m
  )
}

#' Create nested calibration sets
#'
#' From a sample of size `n`, generates a list of nested index sets
#' corresponding to increasing calibration sizes given by `calib_grid`.
#' The observations are randomly permuted, and for each `k` in the grid
#' the first `k` indices are returned (sorted). `k = 0` yields an
#' empty integer vector.
#'
#' @param n Integer, total sample size.
#' @param calib_grid Integer vector of calibration sizes (must include 0
#'   or positive values up to `n`).
#' @param seed Optional integer for random seed.
#'
#' @return A named list of integer vectors; the name of each element is
#'   the character representation of `k`.
#' @export
make_nested_calibration_sets <- function(n, calib_grid, seed = NULL) {
  if (!is.null(seed)) {
    set.seed(seed)
  }
  
  ord <- sample(seq_len(n))
  
  out <- lapply(calib_grid, function(k) {
    if (k == 0) integer(0) else sort(ord[seq_len(k)])
  })
  
  names(out) <- as.character(calib_grid)
  out
}

```

## GP likelihood and prediction
```{r, send_to = 'R/Study_1_Helper.R'}


#' Compute GP log-likelihood and auxiliary quantities (1D)
#'
#' Evaluates the Gaussian process log-marginal likelihood for the
#' one-dimensional model assuming a separable covariance function.
#' The covariance matrix is \eqn{\sigma_\varepsilon^2(\rho^2 R + I)},
#' where \eqn{R} depends on the squared distances in `x` and `u`.
#'
#' @param y Numeric vector of responses.
#' @param u Numeric vector, latent covariate.
#' @param Dx Numeric matrix of squared distances in the `x`-space.
#' @param logtheta Numeric vector of length 3: log-transformed
#'   \eqn{\rho}, \eqn{\theta_x}, \eqn{\theta_u}.
#' @param sigma2_eps Error variance \eqn{\sigma_\varepsilon^2}.
#'
#' @return A list with components `loglik` (log-likelihood value),
#'   `R` (correlation matrix), `A` (scaled covariance matrix),
#'   `cholA` (Cholesky factor of `A`), `Ainv_y` (solution of `A x = y`),
#'   `logdetA` (log determinant of `A`), and `quad` (quadratic form).
#' @export
gp_state_1d <- function(y, u, Dx, logtheta, sigma2_eps) {
  n <- length(y)
  
  rho <- exp(logtheta[1])
  theta_x <- exp(logtheta[2])
  theta_u <- exp(logtheta[3])
  
  Du <- pairwise_sqdist(matrix(u, ncol = 1))
  R <- exp(-theta_x * Dx - theta_u * Du)
  
  A <- rho^2 * R + diag(n)
  
  U <- safe_chol(A)
  Ainv_y <- solve_chol(U, y)
  
  logdetA <- 2 * sum(log(diag(U)))
  quad <- sum(y * Ainv_y)
  
  loglik <- -0.5 * (
    n * log(2 * pi * sigma2_eps) +
      logdetA +
      quad / sigma2_eps
  )
  
  list(
    loglik = loglik,
    R = R,
    A = A,
    cholA = U,
    Ainv_y = Ainv_y,
    logdetA = logdetA,
    quad = quad
  )
}

#' GP predictive draw (1D)
#'
#' Computes the predictive mean and variance for new inputs
#' `x_star`, `u_star` given training data. The predictive
#' distribution can be for the latent function (if `noisy = FALSE`)
#' or for noisy observations (if `noisy = TRUE`, adding the error
#' variance).
#'
#' @param x_train Numeric vector of training covariate `x`.
#' @param u_train Numeric vector of training latent variable `u`.
#' @param y_train Numeric vector of training responses.
#' @param x_star Numeric vector of test points for `x`.
#' @param u_star Numeric vector of test points for `u`.
#' @param logtheta Numeric vector of log-transformed GP hyperparameters.
#' @param sigma2_eps Error variance.
#' @param noisy Logical, if `TRUE` the predictive variance includes
#'   \eqn{\sigma_\varepsilon^2} (default `FALSE`).
#'
#' @return A list with components `mean` (predictive mean vector) and
#'   `var` (predictive variance vector).
#' @export
gp_predict_draw <- function(x_train, u_train, y_train,
                            x_star, u_star,
                            logtheta, sigma2_eps,
                            noisy = FALSE) {
  n <- length(y_train)
  
  rho <- exp(logtheta[1])
  theta_x <- exp(logtheta[2])
  theta_u <- exp(logtheta[3])
  
  Dx_train <- pairwise_sqdist(matrix(x_train, ncol = 1))
  Du_train <- pairwise_sqdist(matrix(u_train, ncol = 1))
  
  R <- exp(-theta_x * Dx_train - theta_u * Du_train)
  K <- rho^2 * sigma2_eps * R
  C <- K + sigma2_eps * diag(n)
  
  U <- safe_chol(C)
  alpha <- solve_chol(U, y_train)
  
  Dxs <- pairwise_sqdist(
    matrix(x_star, ncol = 1),
    matrix(x_train, ncol = 1)
  )
  Dus <- pairwise_sqdist(
    matrix(u_star, ncol = 1),
    matrix(u_train, ncol = 1)
  )
  
  R_star <- exp(-theta_x * Dxs - theta_u * Dus)
  K_star <- rho^2 * sigma2_eps * R_star
  
  mu <- as.numeric(K_star %*% alpha)
  
  v <- forwardsolve(t(U), t(K_star))
  var_lat <- rho^2 * sigma2_eps - colSums(v^2)
  var_lat <- pmax(var_lat, 1e-10)
  
  if (noisy) {
    var_lat <- var_lat + sigma2_eps
  }
  
  list(mean = mu, var = var_lat)
}

```

## EIV-GP sampler

Note I elinminated original global variable such as "a_eps0". Rather, turns in corresponding functions.
```{r, send_to = 'R/Study_1_Helper.R'}
#' Create default MCMC control parameters
#'
#' Chooses sampling intensities (fraction of local updates, block-
#' wise elliptical slice sampling frequency, etc.) according to a
#' preset profile.
#'
#' @param n Integer, total number of observations.
#' @param n_mis Integer, number of unobserved (latent) `u` values.
#' @param preset Character, one of `"fast"`, `"balanced"` or
#'   `"thorough"`.
#'
#' @return A list of control parameters.
#' @export
make_default_control <- function(n, n_mis,
                                 preset = c("fast", "balanced", "thorough")) {
  preset <- match.arg(preset)
  
  if (preset == "fast") {
    local_frac <- 0.03
    theta_update_every <- 10L
    block_ess_every <- 2L
    n_blocks_per_iter <- 1L
    global_ess_every <- 50L
    full_local_every <- 100L
  }
  
  if (preset == "balanced") {
    local_frac <- 0.06
    theta_update_every <- 5L
    block_ess_every <- 1L
    n_blocks_per_iter <- 1L
    global_ess_every <- 25L
    full_local_every <- 50L
  }
  
  if (preset == "thorough") {
    local_frac <- 0.10
    theta_update_every <- 3L
    block_ess_every <- 1L
    n_blocks_per_iter <- 2L
    global_ess_every <- 10L
    full_local_every <- 25L
  }
  
  local_per_iter <- min(
    n_mis,
    max(3L, ceiling(local_frac * max(n_mis, 1L)))
  )
  
  ess_block_size <- min(
    n_mis,
    max(5L, ceiling(sqrt(max(n_mis, 1L))))
  )
  
  list(
    preset = preset,
    local_per_iter = local_per_iter,
    full_local_every = full_local_every,
    ess_block_size = ess_block_size,
    block_ess_every = block_ess_every,
    n_blocks_per_iter = n_blocks_per_iter,
    global_ess_every = global_ess_every,
    theta_update_every = theta_update_every,
    theta_slice_width_init = c(1.0, 1.0, 1.0),
    adapt_theta_width = TRUE,
    adapt_every = 100L,
    adapt_window = 500L,
    theta_width_min = 0.20,
    theta_width_max = 2.50
  )
}

#' Check ordinal constraints
#'
#' Verifies that each element of `u` lies in the interval defined
#' by the thresholds `tau` and the class label `c_ord`.
#'
#' @param u Numeric vector, latent variable.
#' @param c_ord Integer vector of class labels.
#' @param tau Numeric vector of thresholds.
#' @return Logical scalar, `TRUE` if all constraints are satisfied.
#' @export
check_constraints_1d <- function(u, c_ord, tau) {
  lower <- c(-Inf, tau)[c_ord]
  upper <- c(tau, Inf)[c_ord]
  
  all(u > lower & u <= upper)
}

#' Log-prior for GP hyperparameters
#'
#' Evaluates the log-prior density for the three log-transformed GP
#' hyperparameters (log-rho, log-theta_x, log-theta_u). The prior is
#' a product of independent normal densities truncated to a bounded
#' region.
#'
#' @param logtheta Numeric vector of length 3.
#' @param mean Prior mean vector (default from original code).
#' @param sd Prior standard deviation vector (default from original code).
#' @param bounds 3x2 matrix of lower and upper bounds for each
#'   log-parameter (default from original code).
#' @return Log-prior value, or `-Inf` if outside bounds.
#' @export
log_prior_logtheta <- function(logtheta,
                               mean = c(log(3), log(0.5), log(0.5)),
                               sd = c(1.5, 1.5, 1.5),
                               bounds = rbind(
                                 log_rho     = c(log(0.05), log(100)),
                                 log_theta_x = c(log(1e-4), log(100)),
                                 log_theta_u = c(log(1e-4), log(100))
                               )) {
  if (any(logtheta < bounds[, 1]) ||
      any(logtheta > bounds[, 2])) {
    return(-Inf)
  }
  
  sum(dnorm(
    logtheta,
    mean = mean,
    sd = sd,
    log = TRUE
  ))
}

#' Log-likelihood with ordinal constraints (1D)
#'
#' Computes the GP log-likelihood only if the latent `u` satisfies the
#' ordinal constraints given by `c_ord` and `tau`.
#'
#' @param y Numeric vector of responses.
#' @param u Numeric vector, latent variable.
#' @param Dx Squared distance matrix in x.
#' @param c_ord Integer class labels.
#' @param tau Numeric vector of thresholds.
#' @param logtheta Log GP hyperparameters.
#' @param sigma2_eps Error variance.
#' @return Log-likelihood value, or `-Inf` if constraints violated.
#' @export
gp_loglik_with_constraints_1d <- function(y, u, Dx, c_ord, tau,
                                          logtheta, sigma2_eps) {
  if (!check_constraints_1d(u, c_ord, tau)) {
    return(-Inf)
  }
  
  gp_state_1d(y, u, Dx, logtheta, sigma2_eps)$loglik
}

#' Log-posterior for GP hyperparameters (1D)
#'
#' Sum of log-prior and log-likelihood for the GP parameters.
#' By default uses the original prior constants.
#'
#' @param y Response vector.
#' @param u Latent variable vector.
#' @param Dx Squared distance matrix.
#' @param logtheta Log hyperparameters.
#' @param sigma2_eps Error variance.
#' @param prior_mean Prior mean (optional).
#' @param prior_sd Prior sd (optional).
#' @param bounds Prior bounds (optional).
#' @return Log posterior value.
#' @export
theta_logpost_1d <- function(y, u, Dx, logtheta, sigma2_eps,
                             prior_mean = c(log(3), log(0.5), log(0.5)),
                             prior_sd = c(1.5, 1.5, 1.5),
                             bounds = rbind(
                               log_rho     = c(log(0.05), log(100)),
                               log_theta_x = c(log(1e-4), log(100)),
                               log_theta_u = c(log(1e-4), log(100))
                             )) {
  lp <- log_prior_logtheta(logtheta, mean = prior_mean, sd = prior_sd,
                           bounds = bounds)
  
  if (!is.finite(lp)) {
    return(-Inf)
  }
  
  gp_state_1d(y, u, Dx, logtheta, sigma2_eps)$loglik + lp
}

#' Initialize thresholds for ordinal classes
#'
#' Creates a starting value for the thresholds `tau` based on
#' empirical quantiles of the class assignments, adjusted by any
#' observed calibrated `u` values.
#'
#' @param c_ord Integer class labels.
#' @param u_obs Observed latent values (for calibration).
#' @param calib_idx Indices of calibrated observations.
#' @param m Number of classes.
#' @param tau_bound Absolute bound for threshold values.
#' @return Numeric vector of `m-1` thresholds.
#' @export
initialize_tau_1d <- function(c_ord, u_obs, calib_idx, m, tau_bound = 8) {
  n <- length(c_ord)
  
  counts <- tabulate(c_ord, nbins = m)
  probs <- cumsum(counts)[1:(m - 1)] / n
  probs <- pmin(pmax(probs, 0.03), 0.97)
  
  tau <- qnorm(probs)
  tau <- pmin(pmax(tau, -tau_bound + 1e-3), tau_bound - 1e-3)
  
  if (length(calib_idx) > 0) {
    obs_c <- c_ord[calib_idx]
    obs_u <- u_obs[calib_idx]
    
    eps_ord <- 1e-4
    
    for (j in seq_len(m - 1)) {
      lower <- max(c(-tau_bound, obs_u[obs_c <= j]), na.rm = TRUE)
      upper <- min(c( tau_bound, obs_u[obs_c >  j]), na.rm = TRUE)
      
      if (lower >= upper) {
        stop("Calibrated u values are incompatible with ordinal labels.")
      }
      
      tau[j] <- min(max(tau[j], lower + eps_ord), upper - eps_ord)
    }
  }
  
  tau
}

#' Gibbs update for thresholds (1D)
#'
#' Samples new thresholds conditional on the current latent `u` and
#' class labels `c_ord`.
#'
#' @param tau Current threshold vector.
#' @param u Latent variable vector.
#' @param c_ord Integer class labels.
#' @param m Number of classes.
#' @param tau_bound Absolute bound.
#' @return Updated threshold vector.
#' @export
update_tau_1d <- function(tau, u, c_ord, m, tau_bound = 8) {
  tau_new <- tau
  
  for (j in seq_len(m - 1)) {
    L <- max(
      c(
        -tau_bound,
        if (j > 1) tau_new[j - 1] else -Inf,
        u[c_ord <= j]
      ),
      na.rm = TRUE
    )
    
    U <- min(
      c(
        tau_bound,
        if (j < m - 1) tau_new[j + 1] else Inf,
        u[c_ord > j]
      ),
      na.rm = TRUE
    )
    
    if (is.finite(L) && is.finite(U) && L < U) {
      tau_new[j] <- runif(1, L, U)
    }
  }
  
  tau_new
}

#' Sample error variance from inverse-gamma conditional
#'
#' Updates \eqn{\sigma_\varepsilon^2} under the conjugate inverse-gamma
#' prior with shape `a_eps0` and rate `b_eps0`.
#'
#' @param y Response vector.
#' @param u Latent variable vector.
#' @param Dx Squared distance matrix.
#' @param logtheta Log GP hyperparameters.
#' @param a_eps0 Prior shape (default 2).
#' @param b_eps0 Prior rate (default 0.05).
#' @return Sampled error variance.
#' @export
sample_sigma2_eps_1d <- function(y, u, Dx, logtheta,
                                 a_eps0 = 2, b_eps0 = 0.05) {
  n <- length(y)
  
  rho <- exp(logtheta[1])
  theta_x <- exp(logtheta[2])
  theta_u <- exp(logtheta[3])
  
  Du <- pairwise_sqdist(matrix(u, ncol = 1))
  R <- exp(-theta_x * Dx - theta_u * Du)
  
  A <- rho^2 * R + diag(n)
  
  U <- safe_chol(A)
  Ainv_y <- solve_chol(U, y)
  quad <- sum(y * Ainv_y)
  
  shape <- a_eps0 + n / 2
  rate <- b_eps0 + 0.5 * quad
  
  1 / rgamma(1, shape = shape, rate = rate)
}

#' Block elliptical slice sampling update for latent u
#'
#' Updates a block of latent `u` values using elliptical slice sampling
#' that respects the GP likelihood and ordinal constraints.
#'
#' @param y Response vector.
#' @param u Current latent vector.
#' @param Dx Squared distance matrix.
#' @param c_ord Class labels.
#' @param tau Thresholds.
#' @param logtheta Log GP hyperparameters.
#' @param sigma2_eps Error variance.
#' @param block_idx Indices of the block to update.
#' @param max_try Maximum number of angle proposals.
#' @return List with new `u`, number of likelihood evaluations,
#'   and acceptance flag.
#' @export
update_u_ess_block_1d <- function(y, u, Dx, c_ord, tau, logtheta, sigma2_eps,
                                  block_idx, max_try = 300L) {
  if (length(block_idx) == 0) {
    return(list(u = u, n_eval = 0L, accepted = TRUE))
  }
  
  u_block <- u[block_idx]
  
  loglik_fun <- function(u_block_prop) {
    u_prop <- u
    u_prop[block_idx] <- u_block_prop
    
    gp_loglik_with_constraints_1d(
      y = y,
      u = u_prop,
      Dx = Dx,
      c_ord = c_ord,
      tau = tau,
      logtheta = logtheta,
      sigma2_eps = sigma2_eps
    )
  }
  
  loglik_cur <- loglik_fun(u_block)
  
  if (!is.finite(loglik_cur)) {
    stop("Current latent state violates constraints or has invalid likelihood.")
  }
  
  nu <- rnorm(length(u_block))
  
  logy <- loglik_cur + log(runif(1))
  
  angle <- runif(1, 0, 2 * pi)
  angle_min <- angle - 2 * pi
  angle_max <- angle
  
  n_eval <- 1L
  
  for (try_id in seq_len(max_try)) {
    u_block_prop <- u_block * cos(angle) + nu * sin(angle)
    
    loglik_prop <- loglik_fun(u_block_prop)
    n_eval <- n_eval + 1L
    
    if (is.finite(loglik_prop) && loglik_prop >= logy) {
      u_new <- u
      u_new[block_idx] <- u_block_prop
      return(list(u = u_new, n_eval = n_eval, accepted = TRUE))
    }
    
    if (angle < 0) {
      angle_min <- angle
    } else {
      angle_max <- angle
    }
    
    angle <- runif(1, angle_min, angle_max)
  }
  
  list(u = u, n_eval = n_eval, accepted = FALSE)
}

#' Local slice update for latent u (z-space)
#'
#' Updates a subset of the latent `u` coordinates one-by-one using
#' full-interval slice sampling in the z-space (probit scale), which
#' automatically respects ordinal bounds.
#'
#' @param y Response vector.
#' @param u Current latent vector.
#' @param Dx Squared distance matrix.
#' @param c_ord Class labels.
#' @param tau Thresholds.
#' @param logtheta Log GP hyperparameters.
#' @param sigma2_eps Error variance.
#' @param update_idx Indices of coordinates to update.
#' @return List with new `u` and number of likelihood evaluations.
#' @export
update_u_local_z_slice_1d <- function(y, u, Dx, c_ord, tau,
                                      logtheta, sigma2_eps,
                                      update_idx) {
  n_eval_total <- 0L
  eps_z <- 1e-12
  
  update_idx <- sample(update_idx)
  
  for (i in update_idx) {
    lower_u <- c(-Inf, tau)[c_ord[i]]
    upper_u <- c(tau, Inf)[c_ord[i]]
    
    z_lower <- pnorm(lower_u)
    z_upper <- pnorm(upper_u)
    
    z_lower <- max(z_lower, eps_z)
    z_upper <- min(z_upper, 1 - eps_z)
    
    if (z_upper <= z_lower) next
    
    z0 <- pnorm(u[i])
    z0 <- min(max(z0, z_lower + eps_z), z_upper - eps_z)
    
    logf_z <- function(z) {
      if (z <= z_lower || z >= z_upper) {
        return(-Inf)
      }
      
      u_prop_i <- qnorm(z)
      
      if (!(u_prop_i > lower_u && u_prop_i <= upper_u)) {
        return(-Inf)
      }
      
      u_prop <- u
      u_prop[i] <- u_prop_i
      
      gp_state_1d(y, u_prop, Dx, logtheta, sigma2_eps)$loglik
    }
    
    ans <- full_interval_slice_update(
      x0 = z0,
      logf = logf_z,
      lower = z_lower,
      upper = z_upper,
      max_iter = 200L
    )
    
    n_eval_total <- n_eval_total + ans$n_eval
    u[i] <- qnorm(ans$x)
  }
  
  list(u = u, n_eval = n_eval_total)
}

#' Slice sampling for log GP hyperparameters
#'
#' Updates each component of `logtheta` using bounded slice sampling.
#' Uses the original prior bounds by default.
#'
#' @param y Response vector.
#' @param u Latent variable vector.
#' @param Dx Squared distance matrix.
#' @param logtheta Current log hyperparameters.
#' @param sigma2_eps Error variance.
#' @param theta_slice_width Slice widths per component.
#' @param bounds 3x2 matrix of parameter bounds (default from original code).
#' @return List with updated `logtheta` and number of evaluations.
#' @export
update_logtheta_slice_1d <- function(y, u, Dx, logtheta, sigma2_eps,
                                     theta_slice_width,
                                     bounds = rbind(
                                       log_rho     = c(log(0.05), log(100)),
                                       log_theta_x = c(log(1e-4), log(100)),
                                       log_theta_u = c(log(1e-4), log(100))
                                     )) {
  n_eval_total <- 0L
  
  for (j in seq_along(logtheta)) {
    logf_j <- function(val) {
      lt <- logtheta
      lt[j] <- val
      
      theta_logpost_1d(y, u, Dx, lt, sigma2_eps)
    }
    
    ans <- bounded_slice_update(
      x0 = logtheta[j],
      logf = logf_j,
      w = theta_slice_width[j],
      lower = bounds[j, 1],
      upper = bounds[j, 2],
      max_steps_out = 30L,
      max_iter = 100L
    )
    
    logtheta[j] <- ans$x
    n_eval_total <- n_eval_total + ans$n_eval
  }
  
  list(logtheta = logtheta, n_eval = n_eval_total)
}
```

## MCMC diagnostics
```{r, send_to = 'R/Study_1_Helper.R'}
#' Compute split-Rhat convergence diagnostic
#'
#' Splits each chain in `chain_list` into two halves, pools the
#' resulting sequences into a single matrix, and computes the
#' potential scale reduction factor (Rhat) using the split-chain
#' formula.  Only the most recent `min(chain lengths)` iterations
#' are used to ensure equal lengths.  The function returns `NA`
#' if there are fewer than two chains, if the effective sample size
#' is too small, or if the within-chain variance is zero or invalid.
#'
#' @param chain_list A list of numeric vectors, each representing
#'   one MCMC chain (can contain `NA`/`Inf` values, which are removed).
#'
#' @return A single numeric value, the split-Rhat statistic, or
#'   `NA_real_` if the diagnostic cannot be computed.
#' @export
split_rhat <- function(chain_list) {
  chain_list <- lapply(chain_list, function(x) as.numeric(x[is.finite(x)]))
  
  lens <- vapply(chain_list, length, integer(1))
  n0 <- min(lens)
  
  if (length(chain_list) < 2 || n0 < 20) {
    return(NA_real_)
  }
  
  chain_list <- lapply(chain_list, function(x) tail(x, n0))
  
  n_half <- floor(n0 / 2)
  
  if (n_half < 10) {
    return(NA_real_)
  }
  
  split_mat <- do.call(
    rbind,
    lapply(chain_list, function(x) {
      rbind(
        x[seq_len(n_half)],
        x[(n0 - n_half + 1):n0]
      )
    })
  )
  
  n_split <- ncol(split_mat)
  
  chain_means <- rowMeans(split_mat)
  chain_vars <- apply(split_mat, 1, var)
  
  W <- mean(chain_vars)
  B <- n_split * var(chain_means)
  
  if (!is.finite(W) || W <= 0) {
    return(NA_real_)
  }
  
  var_hat <- ((n_split - 1) / n_split) * W + B / n_split
  
  sqrt(var_hat / W)
}

#' Effective sample size via initial positive sequence
#'
#' Estimates the effective sample size (ESS) of a univariate MCMC
#' sample using the initial positive sequence (IPS) estimator.
#' Non-finite values are silently removed.  A maximum lag is
#' chosen automatically if not supplied.
#'
#' @param x Numeric vector of MCMC draws.
#' @param max_lag Integer, maximum lag for autocorrelation estimation
#'   (default `NULL` chooses `min(n-1, 1000)`).
#'
#' @return A numeric value, the estimated effective sample size
#'   (at least 1, at most the sample size `n`), or `NA_real_` if
#'   the estimate cannot be computed.
#' @export
ess_ips <- function(x, max_lag = NULL) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  
  n <- length(x)
  
  if (n < 5) {
    return(NA_real_)
  }
  
  sx <- sd(x)
  
  if (!is.finite(sx)) {
    return(NA_real_)
  }
  
  if (sx == 0) {
    return(n)
  }
  
  if (is.null(max_lag)) {
    max_lag <- min(n - 1L, 1000L)
  } else {
    max_lag <- min(n - 1L, as.integer(max_lag))
  }
  
  if (max_lag < 1L) {
    return(n)
  }
  
  ac <- tryCatch(
    as.numeric(stats::acf(
      x,
      lag.max = max_lag,
      plot = FALSE,
      demean = TRUE
    )$acf),
    error = function(e) NA_real_
  )
  
  if (length(ac) <= 1 || all(!is.finite(ac))) {
    return(NA_real_)
  }
  
  ac <- ac[-1]
  ac <- ac[is.finite(ac)]
  
  if (length(ac) == 0) {
    return(n)
  }
  
  mm <- floor(length(ac) / 2)
  
  if (mm >= 1) {
    pair_sums <- ac[2 * seq_len(mm) - 1] + ac[2 * seq_len(mm)]
    first_nonpos <- which(pair_sums <= 0)[1]
    
    if (is.na(first_nonpos)) {
      use_lag <- 2 * mm
    } else {
      use_lag <- 2 * (first_nonpos - 1)
    }
  } else {
    use_lag <- ifelse(ac[1] > 0, 1, 0)
  }
  
  if (use_lag > 0) {
    tau_int <- 1 + 2 * sum(ac[seq_len(use_lag)])
  } else {
    tau_int <- 1
  }
  
  tau_int <- max(tau_int, 1)
  ess <- n / tau_int
  min(max(ess, 1), n)
}
```

## Main EIV-GP fitting function

```{r, send_to = 'R/Study_1_Helper.R'}

#' Fit the EIV-GP model for 1D heterogeneous regression
#'
#' Runs multiple MCMC chains for the errors-in-variables Gaussian
#' process model with ordinal latent classes.  Handles missing latent
#' variables, calibration observations, and full posterior sampling
#' of thresholds, GP hyperparameters, and error variance.
#'
#' @param x_raw Numeric vector of raw covariate values.
#' @param y_raw Numeric vector of raw response values.
#' @param c_ord Integer vector of ordinal class labels (1 … m).
#' @param u_true Optional numeric vector of true latent `u` values
#'   (required if `calib_idx` is non-empty).
#' @param calib_idx Integer vector of indices where `u` is observed
#'   (calibration set).
#' @param m Integer, number of ordinal classes (default 6).
#' @param tau_true Optional numeric vector of true thresholds.
#' @param n_iter Total number of MCMC iterations per chain.
#' @param burn Number of burn-in iterations.
#' @param thin Thinning interval.
#' @param n_chains Number of independent chains.
#' @param preset Character, one of `"fast"`, `"balanced"`, `"thorough"`.
#' @param seed Integer seed for reproducibility.
#' @param parallel_chains Logical; if `TRUE` and OS is not Windows,
#'   chains are run in parallel using `mclapply`.
#' @param verbose Logical; if `TRUE`, prints progress messages.
#'
#' @return A list with components `data` (scaled data and metadata),
#'   `control` (MCMC control settings), `mcmc` (posterior samples),
#'   and `diagnostics` (convergence diagnostics).
#' @export
fit_eivgp_1d <- function(x_raw,
                         y_raw,
                         c_ord,
                         u_true = NULL,
                         calib_idx = integer(0),
                         m = 6L,
                         tau_true = NULL,
                         n_iter = 6000L,
                         burn = 2000L,
                         thin = 1L,
                         n_chains = 4L,
                         preset = "balanced",
                         seed = 1L,
                         parallel_chains = FALSE,
                         verbose = FALSE) {
  set.seed(seed)
  
  n <- length(y_raw)
  
  calib_idx <- sort(as.integer(calib_idx))
  miss_idx <- setdiff(seq_len(n), calib_idx)
  
  if (length(calib_idx) > 0 && is.null(u_true)) {
    stop("u_true must be provided when calibration observations are used.")
  }
  
  # Re-define the hyperparameter bounds locally
  # (originally a global variable, now kept within the function)
  theta_log_bounds <- rbind(
    log_rho     = c(log(0.05), log(100)),
    log_theta_x = c(log(1e-4), log(100)),
    log_theta_u = c(log(1e-4), log(100))
  )
  
  x_center <- mean(x_raw)
  x_scale <- sd(x_raw)
  x <- as.numeric((x_raw - x_center) / x_scale)
  
  y_center <- mean(y_raw)
  y_scale <- sd(y_raw)
  y <- as.numeric((y_raw - y_center) / y_scale)
  
  u_obs <- rep(NA_real_, n)
  
  if (length(calib_idx) > 0) {
    u_obs[calib_idx] <- u_true[calib_idx]
  }
  
  tau_bound <- 8
  
  control <- make_default_control(
    n = n,
    n_mis = length(miss_idx),
    preset = preset
  )
  
  Dx <- pairwise_sqdist(matrix(x, ncol = 1))
  n_save <- floor((n_iter - burn) / thin)
  
  initialize_chain_state <- function(chain_seed) {
    set.seed(chain_seed)
    
    tau0 <- initialize_tau_1d(
      c_ord = c_ord,
      u_obs = u_obs,
      calib_idx = calib_idx,
      m = m,
      tau_bound = tau_bound
    )
    
    u0 <- u_obs
    
    if (length(miss_idx) > 0) {
      lower_all0 <- c(-Inf, tau0)[c_ord]
      upper_all0 <- c(tau0, Inf)[c_ord]
      
      u0[miss_idx] <- rtruncnorm_vec(
        mean = rep(0, length(miss_idx)),
        sd = 1,
        lower = lower_all0[miss_idx],
        upper = upper_all0[miss_idx]
      )
    }
    
    if (length(miss_idx) > 0) {
      for (rr in seq_len(3)) {
        tau0 <- update_tau_1d(tau0, u0, c_ord, m, tau_bound)
        
        lower_all0 <- c(-Inf, tau0)[c_ord]
        upper_all0 <- c(tau0, Inf)[c_ord]
        
        u0[miss_idx] <- rtruncnorm_vec(
          mean = rep(0, length(miss_idx)),
          sd = 1,
          lower = lower_all0[miss_idx],
          upper = upper_all0[miss_idx]
        )
      }
    }
    
    sigma2_eps0 <- exp(log(0.05) + rnorm(1, 0, 0.5))
    sigma2_eps0 <- min(max(sigma2_eps0, 1e-4), 2)
    
    Du_init <- pairwise_sqdist(matrix(u0, ncol = 1))
    
    med_dx <- median(Dx[upper.tri(Dx)], na.rm = TRUE)
    med_du <- median(Du_init[upper.tri(Du_init)], na.rm = TRUE)
    
    theta_x_init <- ifelse(
      is.finite(med_dx) && med_dx > 0,
      1 / med_dx,
      0.5
    )
    
    theta_u_init <- ifelse(
      is.finite(med_du) && med_du > 0,
      1 / med_du,
      0.5
    )
    
    rho_init <- sqrt(max(var(y) / sigma2_eps0 - 1, 1))
    
    logtheta0 <- c(log(rho_init), log(theta_x_init), log(theta_u_init))
    logtheta0 <- logtheta0 + rnorm(3, mean = 0, sd = c(0.4, 0.7, 0.7))
    logtheta0 <- pmin(pmax(logtheta0, theta_log_bounds[, 1]),
                      theta_log_bounds[, 2])
    
    list(
      u_curr = u0,
      tau = tau0,
      sigma2_eps = sigma2_eps0,
      logtheta = logtheta0,
      theta_slice_width = control$theta_slice_width_init
    )
  }
  
  run_one_chain <- function(chain_id, chain_seed) {
    set.seed(chain_seed)
    
    state <- initialize_chain_state(chain_seed)
    
    u_curr <- state$u_curr
    tau <- state$tau
    sigma2_eps <- state$sigma2_eps
    logtheta <- state$logtheta
    theta_slice_width <- state$theta_slice_width
    
    samples_u_chain <- matrix(NA_real_, n_save, n)
    samples_tau_chain <- matrix(NA_real_, n_save, m - 1)
    samples_logtheta_chain <- matrix(NA_real_, n_save, 3)
    samples_sigma2_chain <- numeric(n_save)
    
    logtheta_trace_all <- matrix(NA_real_, n_iter, 3)
    
    block_ess_eval_total <- 0L
    block_ess_accept_total <- 0L
    block_ess_total <- 0L
    
    global_ess_eval_total <- 0L
    global_ess_accept_total <- 0L
    global_ess_total <- 0L
    
    local_eval_total <- 0L
    theta_eval_total <- 0L
    theta_update_total <- 0L
    
    save_id <- 0L
    
    for (iter in seq_len(n_iter)) {
      sigma2_eps <- sample_sigma2_eps_1d(y, u_curr, Dx, logtheta)
      
      tau <- update_tau_1d(tau, u_curr, c_ord, m, tau_bound)
      
      if (length(miss_idx) > 0 &&
          iter %% control$block_ess_every == 0) {
        for (bb in seq_len(control$n_blocks_per_iter)) {
          block_idx <- sample(
            miss_idx,
            min(control$ess_block_size, length(miss_idx))
          )
          
          ess <- update_u_ess_block_1d(
            y = y,
            u = u_curr,
            Dx = Dx,
            c_ord = c_ord,
            tau = tau,
            logtheta = logtheta,
            sigma2_eps = sigma2_eps,
            block_idx = block_idx
          )
          
          u_curr <- ess$u
          
          block_ess_eval_total <- block_ess_eval_total + ess$n_eval
          block_ess_accept_total <- block_ess_accept_total +
            as.integer(ess$accepted)
          block_ess_total <- block_ess_total + 1L
        }
      }
      
      if (length(miss_idx) > 0 &&
          control$global_ess_every > 0 &&
          iter %% control$global_ess_every == 0) {
        gess <- update_u_ess_block_1d(
          y = y,
          u = u_curr,
          Dx = Dx,
          c_ord = c_ord,
          tau = tau,
          logtheta = logtheta,
          sigma2_eps = sigma2_eps,
          block_idx = miss_idx
        )
        
        u_curr <- gess$u
        
        global_ess_eval_total <- global_ess_eval_total + gess$n_eval
        global_ess_accept_total <- global_ess_accept_total +
          as.integer(gess$accepted)
        global_ess_total <- global_ess_total + 1L
      }
      
      if (length(miss_idx) > 0) {
        if (iter %% control$full_local_every == 0) {
          local_idx <- miss_idx
        } else {
          local_idx <- sample(
            miss_idx,
            min(control$local_per_iter, length(miss_idx))
          )
        }
        
        loc <- update_u_local_z_slice_1d(
          y = y,
          u = u_curr,
          Dx = Dx,
          c_ord = c_ord,
          tau = tau,
          logtheta = logtheta,
          sigma2_eps = sigma2_eps,
          update_idx = local_idx
        )
        
        u_curr <- loc$u
        local_eval_total <- local_eval_total + loc$n_eval
      }
      
      if (iter %% control$theta_update_every == 0) {
        th <- update_logtheta_slice_1d(
          y = y,
          u = u_curr,
          Dx = Dx,
          logtheta = logtheta,
          sigma2_eps = sigma2_eps,
          theta_slice_width = theta_slice_width
        )
        
        logtheta <- th$logtheta
        
        theta_eval_total <- theta_eval_total + th$n_eval
        theta_update_total <- theta_update_total + 1L
      }
      
      logtheta_trace_all[iter, ] <- logtheta
      
      if (control$adapt_theta_width &&
          iter <= burn &&
          iter >= 200 &&
          iter %% control$adapt_every == 0) {
        lo <- max(1, iter - control$adapt_window + 1)
        recent <- logtheta_trace_all[lo:iter, , drop = FALSE]
        recent <- recent[complete.cases(recent), , drop = FALSE]
        
        if (nrow(recent) >= 50) {
          sds <- apply(recent, 2, sd)
          new_width <- 2 * sds
          new_width <- pmin(
            pmax(new_width, control$theta_width_min),
            control$theta_width_max
          )
          
          if (all(is.finite(new_width))) {
            theta_slice_width <- new_width
          }
        }
      }
      
      if (iter > burn && ((iter - burn) %% thin == 0)) {
        save_id <- save_id + 1L
        
        samples_u_chain[save_id, ] <- u_curr
        samples_tau_chain[save_id, ] <- tau
        samples_logtheta_chain[save_id, ] <- logtheta
        samples_sigma2_chain[save_id] <- sigma2_eps
      }
    }
    
    samples_u_chain <- samples_u_chain[seq_len(save_id), , drop = FALSE]
    samples_tau_chain <- samples_tau_chain[seq_len(save_id), , drop = FALSE]
    samples_logtheta_chain <- samples_logtheta_chain[seq_len(save_id), , drop = FALSE]
    samples_sigma2_chain <- samples_sigma2_chain[seq_len(save_id)]
    
    stats <- data.frame(
      chain = chain_id,
      seed = chain_seed,
      saved = save_id,
      block_ess_eval_total = block_ess_eval_total,
      block_ess_accept_total = block_ess_accept_total,
      block_ess_total = block_ess_total,
      global_ess_eval_total = global_ess_eval_total,
      global_ess_accept_total = global_ess_accept_total,
      global_ess_total = global_ess_total,
      local_eval_total = local_eval_total,
      theta_eval_total = theta_eval_total,
      theta_update_total = theta_update_total
    )
    
    list(
      chain_id = chain_id,
      seed = chain_seed,
      samples_u = samples_u_chain,
      samples_tau = samples_tau_chain,
      samples_logtheta = samples_logtheta_chain,
      samples_sigma2 = samples_sigma2_chain,
      theta_slice_width_final = theta_slice_width,
      stats = stats
    )
  }
  
  chain_seeds <- seed + 10000L * seq_len(n_chains)
  
  use_mclapply <- (
    parallel_chains &&
      .Platform$OS.type != "windows" &&
      n_chains > 1L
  )
  
  mc_cores <- min(
    n_chains,
    max(1L, parallel::detectCores(logical = TRUE) - 2L)
  )
  
  if (verbose) {
    cat("Running EIV-GP with", n_chains, "chain(s).\n")
  }
  
  mcmc_time <- system.time({
    if (use_mclapply) {
      chains <- parallel::mclapply(
        seq_len(n_chains),
        function(cc) {
          run_one_chain(chain_id = cc, chain_seed = chain_seeds[cc])
        },
        mc.cores = mc_cores,
        mc.set.seed = TRUE,
        mc.preschedule = FALSE
      )
    } else {
      chains <- lapply(
        seq_len(n_chains),
        function(cc) {
          run_one_chain(chain_id = cc, chain_seed = chain_seeds[cc])
        }
      )
    }
  })
  
  samples_by_chain <- list(
    u = lapply(chains, function(z) z$samples_u),
    tau = lapply(chains, function(z) z$samples_tau),
    logtheta = lapply(chains, function(z) z$samples_logtheta),
    sigma2 = lapply(chains, function(z) z$samples_sigma2)
  )
  
  samples_u <- do.call(rbind, samples_by_chain$u)
  samples_tau <- do.call(rbind, samples_by_chain$tau)
  samples_logtheta <- do.call(rbind, samples_by_chain$logtheta)
  samples_sigma2 <- unlist(samples_by_chain$sigma2)
  
  mcmc_draw_info <- data.frame(
    chain = rep(
      seq_len(n_chains),
      times = vapply(samples_by_chain$u, nrow, integer(1))
    ),
    draw_within_chain = unlist(
      lapply(samples_by_chain$u, function(mat) seq_len(nrow(mat)))
    )
  )
  
  chain_stats <- do.call(rbind, lapply(chains, function(z) z$stats))
  
  rhat_hyper <- data.frame(
    parameter = c("sigma_epsilon", "rho", "theta_x", "theta_u"),
    rhat = c(
      split_rhat(lapply(samples_by_chain$sigma2, function(v) sqrt(v))),
      split_rhat(lapply(samples_by_chain$logtheta, function(mat) exp(mat[, 1]))),
      split_rhat(lapply(samples_by_chain$logtheta, function(mat) exp(mat[, 2]))),
      split_rhat(lapply(samples_by_chain$logtheta, function(mat) exp(mat[, 3])))
    )
  )
  
  rhat_tau <- data.frame(
    parameter = paste0("tau", seq_len(m - 1)),
    rhat = sapply(seq_len(m - 1), function(j) {
      split_rhat(
        lapply(samples_by_chain$tau, function(mat) mat[, j])
      )
    })
  )
  
  if (length(miss_idx) > 0) {
    rhat_u <- data.frame(
      parameter = paste0("u[", miss_idx, "]"),
      global_index = miss_idx,
      rhat = sapply(seq_along(miss_idx), function(k) {
        jj <- miss_idx[k]
        
        split_rhat(
          lapply(samples_by_chain$u, function(mat) mat[, jj])
        )
      })
    )
  } else {
    rhat_u <- data.frame(
      parameter = character(0),
      global_index = integer(0),
      rhat = numeric(0)
    )
  }
  
  hyper_mat <- cbind(
    sigma_epsilon = sqrt(samples_sigma2),
    rho = exp(samples_logtheta[, 1]),
    theta_x = exp(samples_logtheta[, 2]),
    theta_u = exp(samples_logtheta[, 3])
  )
  
  tau_mat <- samples_tau
  colnames(tau_mat) <- paste0("tau", seq_len(ncol(tau_mat)))
  
  ess_key <- data.frame(
    parameter = colnames(cbind(hyper_mat, tau_mat)),
    ess = apply(cbind(hyper_mat, tau_mat), 2, ess_ips)
  )
  
  diagnostics_summary <- data.frame(
    n_chains = n_chains,
    n_iter = n_iter,
    burn = burn,
    saved_per_chain = mean(vapply(samples_by_chain$u, nrow, integer(1))),
    total_saved_draws = nrow(samples_u),
    max_rhat_hyper = safe_max(rhat_hyper$rhat),
    max_rhat_tau = safe_max(rhat_tau$rhat),
    median_rhat_missing_u = safe_median(rhat_u$rhat),
    max_rhat_missing_u = safe_max(rhat_u$rhat),
    min_ess_key = safe_min(ess_key$ess),
    time_seconds = as.numeric(mcmc_time["elapsed"])
  )
  
  list(
    data = list(
      x_raw = x_raw,
      x = x,
      x_center = x_center,
      x_scale = x_scale,
      y_raw = y_raw,
      y = y,
      y_center = y_center,
      y_scale = y_scale,
      c_ord = c_ord,
      u_true = u_true,
      u_obs = u_obs,
      calib_idx = calib_idx,
      miss_idx = miss_idx,
      tau_true = tau_true,
      m = m
    ),
    control = control,
    mcmc = list(
      samples_u = samples_u,
      samples_tau = samples_tau,
      samples_logtheta = samples_logtheta,
      samples_sigma2 = samples_sigma2,
      samples_by_chain = samples_by_chain,
      mcmc_draw_info = mcmc_draw_info,
      chain_stats = chain_stats
    ),
    diagnostics = list(
      rhat_hyper = rhat_hyper,
      rhat_tau = rhat_tau,
      rhat_u = rhat_u,
      ess_key = ess_key,
      summary = diagnostics_summary
    )
  )
}
```

## Baseline GP Model
```{r, send_to = 'R/Study_1_Helper.R'}
#' Fit a GP model via maximum likelihood (MLE)
#'
#' Estimates GP hyperparameters (variance, correlation scale,
#' length-scales) by minimising the negative log-marginal likelihood
#' using L-BFGS-B optimisation.
#'
#' @param X Numeric matrix of covariates.
#' @param y Numeric response vector.
#'
#' @return A list with elements `par` (MLE on log scale), `value`
#'   (minimum negative log-likelihood), `convergence` (optimisation
#'   convergence code), and copies of `X` and `y`.
#' @export
gp_mle_fit <- function(X, y) {
  X <- as.matrix(X)
  n <- nrow(X)
  d <- ncol(X)
  
  Dlist <- lapply(seq_len(d), function(j) {
    pairwise_sqdist(X[, j, drop = FALSE])
  })
  
  nll <- function(par) {
    log_sigma2 <- par[1]
    log_rho <- par[2]
    log_theta <- par[-c(1, 2)]
    
    sigma2 <- exp(log_sigma2)
    rho <- exp(log_rho)
    theta <- exp(log_theta)
    
    Rexp <- matrix(0, n, n)
    
    for (j in seq_len(d)) {
      Rexp <- Rexp + theta[j] * Dlist[[j]]
    }
    
    R <- exp(-Rexp)
    A <- rho^2 * R + diag(n)
    
    U <- try(safe_chol(A), silent = TRUE)
    
    if (inherits(U, "try-error")) {
      return(1e20)
    }
    
    Ainv_y <- solve_chol(U, y)
    logdetA <- 2 * sum(log(diag(U)))
    quad <- sum(y * Ainv_y)
    
    0.5 * (n * log(2 * pi * sigma2) + logdetA + quad / sigma2)
  }
  
  init <- c(log(0.05), log(3), rep(log(0.5), d))
  lower <- c(log(1e-5), log(0.05), rep(log(1e-4), d))
  upper <- c(log(5), log(100), rep(log(100), d))
  
  opt <- optim(
    par = init,
    fn = nll,
    method = "L-BFGS-B",
    lower = lower,
    upper = upper,
    control = list(maxit = 500)
  )
  
  list(
    par = opt$par,
    value = opt$value,
    convergence = opt$convergence,
    X = X,
    y = y
  )
}

#' Predict from a GP fitted via MLE
#'
#' Computes the predictive mean and variance for new inputs given
#' an object returned by `gp_mle_fit`. If `noisy = TRUE`, the error
#' variance is added to the predictive variance.
#'
#' @param fit A list returned by `gp_mle_fit`.
#' @param Xstar Numeric matrix of test inputs.
#' @param noisy Logical; if `TRUE`, predictive variance includes
#'   residual noise (default `FALSE`).
#'
#' @return A list with components `mean` and `var`.
#' @export
gp_mle_predict <- function(fit, Xstar, noisy = FALSE) {
  X <- fit$X
  y <- fit$y
  
  Xstar <- as.matrix(Xstar)
  
  n <- nrow(X)
  N <- nrow(Xstar)
  d <- ncol(X)
  
  par <- fit$par
  
  sigma2 <- exp(par[1])
  rho <- exp(par[2])
  theta <- exp(par[-c(1, 2)])
  
  Rexp <- matrix(0, n, n)
  
  for (j in seq_len(d)) {
    Rexp <- Rexp + theta[j] * pairwise_sqdist(X[, j, drop = FALSE])
  }
  
  R <- exp(-Rexp)
  K <- rho^2 * sigma2 * R
  C <- K + sigma2 * diag(n)
  
  U <- safe_chol(C)
  alpha <- solve_chol(U, y)
  
  Rstar_exp <- matrix(0, N, n)
  
  for (j in seq_len(d)) {
    Rstar_exp <- Rstar_exp +
      theta[j] * pairwise_sqdist(
        Xstar[, j, drop = FALSE],
        X[, j, drop = FALSE]
      )
  }
  
  Rstar <- exp(-Rstar_exp)
  Kstar <- rho^2 * sigma2 * Rstar
  
  mu <- as.numeric(Kstar %*% alpha)
  
  v <- forwardsolve(t(U), t(Kstar))
  var_lat <- rho^2 * sigma2 - colSums(v^2)
  var_lat <- pmax(var_lat, 1e-10)
  
  if (noisy) {
    var_lat <- var_lat + sigma2
  }
  
  list(mean = mu, var = var_lat)
}

#' Sample from the predictive distribution of an MLE-fitted GP
#'
#' Generates `n_draw` independent draws from the (noisy) predictive
#' distribution of a GP fitted by `gp_mle_fit`.
#'
#' @param fit Output of `gp_mle_fit`.
#' @param Xstar Test input matrix.
#' @param n_draw Number of posterior predictive draws.
#'
#' @return A numeric matrix with `n_draw` rows and `nrow(Xstar)` columns.
#' @export
sample_gp_mle_predictive <- function(fit, Xstar, n_draw = 1000) {
  pred <- gp_mle_predict(fit, Xstar = Xstar, noisy = TRUE)
  
  n_test <- length(pred$mean)
  
  matrix(
    rnorm(
      n_draw * n_test,
      mean = rep(pred$mean, each = n_draw),
      sd = rep(sqrt(pred$var), each = n_draw)
    ),
    nrow = n_draw,
    ncol = n_test
  )
}

#' Conditional mean scores for ordinal classes
#'
#' Computes the expected value of a standard normal random variable
#' truncated to each ordinal class based on empirical quantiles
#' of the class labels.
#'
#' @param c_ord Integer vector of class labels (1 … m).
#' @param m Number of classes.
#'
#' @return Numeric vector of length `m` giving the conditional
#'   mean for each class.
#' @export
conditional_mean_scores <- function(c_ord, m) {
  counts <- tabulate(c_ord, nbins = m)
  probs <- cumsum(counts) / sum(counts)
  probs <- pmin(pmax(probs[seq_len(m - 1)], 1e-4), 1 - 1e-4)
  
  tau_hat <- qnorm(probs)
  
  lower <- c(-Inf, tau_hat)
  upper <- c(tau_hat, Inf)
  
  denom <- pnorm(upper) - pnorm(lower)
  numer <- dnorm(lower) - dnorm(upper)
  
  numer / pmax(denom, .Machine$double.eps)
}

#' Monotone increasing scores from latent parameters
#'
#' Constructs a monotone increasing vector of length `m` starting
#' at 0 and ending at 1, based on a vector of unconstrained
#' log-ratio parameters `a`.
#'
#' @param a Numeric vector of length `m-2` (log-ratios).
#' @param m Number of categories (>= 2).
#'
#' @return Numeric vector of length `m`, strictly increasing from
#'   0 to 1.
#' @export
make_monotone_scores <- function(a, m) {
  if (m == 1) {
    return(0)
  }
  
  if (m == 2) {
    return(c(0, 1))
  }
  
  logits <- c(a, 0)
  e <- exp(logits - max(logits))
  inc <- e / sum(e)
  
  c(0, cumsum(inc))
}

#' Fit a GP with a learned ordinal embedding
#'
#' Estimates GP hyperparameters together with a learned monotone
#' embedding of the ordinal classes by maximum likelihood.
#' Multiple random starts are used to avoid local optima.
#'
#' @param x Numeric vector of continuous covariate.
#' @param c_ord Integer class labels.
#' @param y Numeric response vector.
#' @param m Number of classes.
#' @param n_starts Number of random starting values.
#'
#' @return A list with `par` (log-hyperparameters), `a` (log-ratios
#'   for embedding), `z_scores` (final embedding), `value`, etc.
#' @export
gp_mle_fit_learned_embedding <- function(x, c_ord, y, m,
                                         n_starts = 8L) {
  n <- length(y)
  
  Dx <- pairwise_sqdist(matrix(x, ncol = 1))
  
  nll <- function(par) {
    log_sigma2 <- par[1]
    log_rho <- par[2]
    log_theta_x <- par[3]
    log_theta_z <- par[4]
    
    a <- par[-(1:4)]
    
    sigma2 <- exp(log_sigma2)
    rho <- exp(log_rho)
    theta_x <- exp(log_theta_x)
    theta_z <- exp(log_theta_z)
    
    z_scores <- make_monotone_scores(a, m)
    z <- z_scores[c_ord]
    
    Dz <- pairwise_sqdist(matrix(z, ncol = 1))
    
    R <- exp(-theta_x * Dx - theta_z * Dz)
    A <- rho^2 * R + diag(n)
    
    U <- try(safe_chol(A), silent = TRUE)
    
    if (inherits(U, "try-error")) {
      return(1e20)
    }
    
    Ainv_y <- solve_chol(U, y)
    logdetA <- 2 * sum(log(diag(U)))
    quad <- sum(y * Ainv_y)
    
    0.5 * (n * log(2 * pi * sigma2) + logdetA + quad / sigma2)
  }
  
  lower <- c(
    log(1e-5),
    log(0.05),
    log(1e-4),
    log(1e-4),
    rep(-6, max(m - 2, 0))
  )
  
  upper <- c(
    log(5),
    log(100),
    log(100),
    log(100),
    rep(6, max(m - 2, 0))
  )
  
  make_init <- function() {
    c(
      log(0.05),
      log(3),
      log(0.5),
      log(0.5),
      rnorm(max(m - 2, 0), 0, 0.5)
    )
  }
  
  starts <- replicate(n_starts, make_init(), simplify = FALSE)
  
  opts <- lapply(starts, function(init) {
    optim(
      par = init,
      fn = nll,
      method = "L-BFGS-B",
      lower = lower,
      upper = upper,
      control = list(maxit = 500)
    )
  })
  
  best_id <- which.min(vapply(opts, function(z) z$value, numeric(1)))
  opt <- opts[[best_id]]
  
  z_scores <- make_monotone_scores(opt$par[-(1:4)], m)
  
  list(
    par = opt$par[1:4],
    a = opt$par[-(1:4)],
    z_scores = z_scores,
    value = opt$value,
    convergence = opt$convergence,
    x = x,
    c_ord = c_ord,
    y = y,
    m = m
  )
}

#' Predict from a GP with learned ordinal embedding
#'
#' Computes predictive mean and variance for new inputs using a
#' fitted learned-embedding GP model.
#'
#' @param fit Output of `gp_mle_fit_learned_embedding`.
#' @param x_star Numeric vector of new continuous covariates.
#' @param c_star Integer vector of new class labels.
#' @param noisy Logical; include residual noise (default `TRUE`).
#'
#' @return List with `mean` and `var`.
#' @export
gp_mle_predict_learned_embedding <- function(fit, x_star, c_star,
                                             noisy = TRUE) {
  X_train <- cbind(fit$x, fit$z_scores[fit$c_ord])
  X_star <- cbind(x_star, fit$z_scores[c_star])
  
  fake_fit <- list(
    par = fit$par,
    X = X_train,
    y = fit$y
  )
  
  gp_mle_predict(fake_fit, Xstar = X_star, noisy = noisy)
}

#' Sample from the predictive distribution of the learned-embedding GP
#'
#' @param fit Output of `gp_mle_fit_learned_embedding`.
#' @param x_star Numeric vector of new `x` values.
#' @param c_star Integer vector of new class labels.
#' @param n_draw Number of draws.
#'
#' @return A matrix of predictive samples (n_draw x n_test).
#' @export
sample_gp_learned_embedding_predictive <- function(fit, x_star, c_star,
                                                   n_draw = 1000) {
  pred <- gp_mle_predict_learned_embedding(
    fit,
    x_star = x_star,
    c_star = c_star,
    noisy = TRUE
  )
  
  n_test <- length(pred$mean)
  
  matrix(
    rnorm(
      n_draw * n_test,
      mean = rep(pred$mean, each = n_draw),
      sd = rep(sqrt(pred$var), each = n_draw)
    ),
    nrow = n_draw,
    ncol = n_test
  )
}

#' Fit baseline GP models with different embeddings
#'
#' Fits three baseline GP models on scaled data:
#' \describe{
#'   \item{GP-Gaussian}{Uses quantile-based Gaussian scores.}
#'   \item{GP-CondMean}{Uses conditional mean scores.}
#'   \item{GP-LearnedEmb}{Learns a monotone embedding by MLE.}
#' }
#'
#' @param x_raw Raw continuous covariate.
#' @param y_raw Raw response.
#' @param c_ord Integer class labels.
#' @param m Number of classes.
#' @param n_starts_learned Number of random starts for the learned
#'   embedding model.
#'
#' @return A list with centering/scaling information and the three
#'   fitted model objects.
#' @export
fit_embedding_baselines <- function(x_raw, y_raw, c_ord, m,
                                    n_starts_learned = 8L) {
  x_center <- mean(x_raw)
  x_scale <- sd(x_raw)
  x <- as.numeric((x_raw - x_center) / x_scale)
  
  y_center <- mean(y_raw)
  y_scale <- sd(y_raw)
  y <- as.numeric((y_raw - y_center) / y_scale)
  
  z_gauss <- qnorm(c_ord / (m + 1))
  
  fit_gauss <- gp_mle_fit(
    X = cbind(x, z_gauss),
    y = y
  )
  
  z_cm_scores <- conditional_mean_scores(c_ord, m)
  z_cm <- z_cm_scores[c_ord]
  
  fit_cm <- gp_mle_fit(
    X = cbind(x, z_cm),
    y = y
  )
  
  fit_learned <- gp_mle_fit_learned_embedding(
    x = x,
    c_ord = c_ord,
    y = y,
    m = m,
    n_starts = n_starts_learned
  )
  
  list(
    x_center = x_center,
    x_scale = x_scale,
    y_center = y_center,
    y_scale = y_scale,
    fit_gauss = fit_gauss,
    fit_cm = fit_cm,
    fit_learned = fit_learned,
    z_cm_scores = z_cm_scores
  )
}

#' Generate predictive samples from baseline GP models
#'
#' For each of the three baseline embeddings, produces `n_draw`
#' posterior predictive samples on the original scale.
#'
#' @param baselines Output of `fit_embedding_baselines`.
#' @param x_star_raw Raw test covariate.
#' @param c_star Integer test class labels.
#' @param m Number of classes.
#' @param n_draw Number of predictive draws per method.
#'
#' @return A named list of three matrices (`GP-Gaussian`,
#'   `GP-CondMean`, `GP-LearnedEmb`), each `n_draw` x `n_test`.
#' @export
predict_embedding_baseline_samples <- function(baselines,
                                               x_star_raw,
                                               c_star,
                                               m,
                                               n_draw = 1000) {
  x_star <- as.numeric((x_star_raw - baselines$x_center) / baselines$x_scale)
  
  z_gauss_star <- qnorm(c_star / (m + 1))
  
  draws_gauss_std <- sample_gp_mle_predictive(
    baselines$fit_gauss,
    Xstar = cbind(x_star, z_gauss_star),
    n_draw = n_draw
  )
  
  draws_gauss <- baselines$y_center + baselines$y_scale * draws_gauss_std
  
  z_cm_star <- baselines$z_cm_scores[c_star]
  
  draws_cm_std <- sample_gp_mle_predictive(
    baselines$fit_cm,
    Xstar = cbind(x_star, z_cm_star),
    n_draw = n_draw
  )
  
  draws_cm <- baselines$y_center + baselines$y_scale * draws_cm_std
  
  draws_learned_std <- sample_gp_learned_embedding_predictive(
    baselines$fit_learned,
    x_star = x_star,
    c_star = c_star,
    n_draw = n_draw
  )
  
  draws_learned <- baselines$y_center + baselines$y_scale * draws_learned_std
  
  list(
    `GP-Gaussian` = draws_gauss,
    `GP-CondMean` = draws_cm,
    `GP-LearnedEmb` = draws_learned
  )
}
```

## Predictive scoring
```{r, send_to = 'R/Study_1_Helper.R'}
#' Continuous ranked probability score (single observation)
#'
#' Computes the CRPS for a single observation given a set of
#' predictive draws.
#'
#' @param draws Numeric vector of predictive samples.
#' @param y True observation.
#'
#' @return CRPS value (non-negative).
#' @export
crps_sample_one <- function(draws, y) {
  draws <- sort(as.numeric(draws))
  S <- length(draws)
  
  term1 <- mean(abs(draws - y))
  
  weights <- 2 * seq_len(S) - S - 1
  mean_abs_pair <- 2 * sum(weights * draws) / S^2
  
  term1 - 0.5 * mean_abs_pair
}

#' Continuous ranked probability score (matrix of draws)
#'
#' Vectorised CRPS for a matrix of predictive draws (rows = draws,
#' columns = test points) and a vector of true values.
#'
#' @param draw_mat Numeric matrix, columns correspond to test points.
#' @param y Numeric vector of true values, length `ncol(draw_mat)`.
#'
#' @return Numeric vector of CRPS values per test point.
#' @export
crps_sample_matrix <- function(draw_mat, y) {
  vapply(seq_along(y), function(j) {
    crps_sample_one(draw_mat[, j], y[j])
  }, numeric(1))
}

#' Interval score for a central prediction interval
#'
#' @param lo Lower bound of the prediction interval.
#' @param hi Upper bound of the prediction interval.
#' @param y True observation.
#' @param alpha Significance level (default 0.05 for 95% interval).
#'
#' @return Interval score.
#' @export
interval_score <- function(lo, hi, y, alpha = 0.05) {
  width <- hi - lo
  
  width +
    2 / alpha * (lo - y) * (y < lo) +
    2 / alpha * (y - hi) * (y > hi)
}

#' Summarise predictive samples with common metrics
#'
#' Computes RMSE, MAE, 95% coverage, interval width, CRPS, and
#' interval score from a matrix of predictive draws and true
#' observations.
#'
#' @param draw_mat Numeric matrix (n_draw x n_test).
#' @param y_true True test responses.
#' @param method Character string, method name.
#' @param rep_id Replication identifier.
#' @param n_calib Calibration size.
#' @param scenario Scenario name.
#'
#' @return A data frame with one row and the listed summary metrics.
#' @export
summarize_predictive_samples <- function(draw_mat, y_true,
                                         method,
                                         rep_id,
                                         n_calib,
                                         scenario) {
  pred_mean <- colMeans(draw_mat)
  
  qs <- apply(
    draw_mat,
    2,
    quantile,
    probs = c(0.025, 0.975),
    names = FALSE
  )
  
  lo <- qs[1, ]
  hi <- qs[2, ]
  
  data.frame(
    rep = rep_id,
    scenario = scenario,
    n_calib = n_calib,
    method = method,
    RMSE = sqrt(mean((pred_mean - y_true)^2)),
    MAE = mean(abs(pred_mean - y_true)),
    Coverage95 = mean(y_true >= lo & y_true <= hi),
    Width95 = mean(hi - lo),
    CRPS = mean(crps_sample_matrix(draw_mat, y_true)),
    IntervalScore95 = mean(interval_score(lo, hi, y_true)),
    stringsAsFactors = FALSE
  )
}

#' Sample from the oracle predictive distribution
#'
#' Generates draws of `y*` from the true data-generating process
#' for a given x and class.
#'
#' @param x_star True x value.
#' @param c_star True class label.
#' @param tau_true True thresholds.
#' @param scenario `"active"` or `"inactive"`.
#' @param sigma_eps Error standard deviation.
#' @param n_draw Number of draws.
#'
#' @return A numeric vector of length `n_draw`.
#' @export
sample_oracle_y_star <- function(x_star, c_star, tau_true,
                                 scenario = "active",
                                 sigma_eps = 0.1,
                                 n_draw = 1000) {
  lower <- c(-Inf, tau_true)[c_star]
  upper <- c(tau_true, Inf)[c_star]
  
  u_star <- rtruncnorm_vec(
    mean = rep(0, n_draw),
    sd = 1,
    lower = rep(lower, n_draw),
    upper = rep(upper, n_draw)
  )
  
  f_star <- f0_1d(x_star, u_star, scenario = scenario)
  
  f_star + rnorm(n_draw, 0, sigma_eps)
}

#' Sample oracle test set predictions
#'
#' @param x_test Numeric vector of test x values.
#' @param c_test Integer vector of test class labels.
#' @param tau_true True thresholds.
#' @param scenario `"active"` or `"inactive"`.
#' @param sigma_eps Error standard deviation.
#' @param n_draw Number of draws per test point.
#'
#' @return A matrix with `n_draw` rows and `length(c_test)` columns.
#' @export
sample_oracle_test_y <- function(x_test,
                                 c_test,
                                 tau_true,
                                 scenario,
                                 sigma_eps,
                                 n_draw = 1000) {
  n_test <- length(c_test)
  out <- matrix(NA_real_, nrow = n_draw, ncol = n_test)
  
  for (j in seq_len(n_test)) {
    out[, j] <- sample_oracle_y_star(
      x_star = x_test[j],
      c_star = c_test[j],
      tau_true = tau_true,
      scenario = scenario,
      sigma_eps = sigma_eps,
      n_draw = n_draw
    )
  }
  
  out
}

#' Generate predictive draws from an EIV-GP fit
#'
#' Draws from the posterior predictive distribution of `y` for new
#' `x_test_raw` and `c_test` using selected posterior draws from the
#' fitted EIV-GP model.
#'
#' @param x_test_raw Raw test covariate vector.
#' @param c_test Integer test class labels.
#' @param fit_obj Output of `fit_eivgp_1d`.
#' @param draw_ids Indices of posterior draws to use.
#' @param n_per_draw Number of predictive samples per posterior draw.
#'
#' @return A matrix with `length(draw_ids) * n_per_draw` rows and
#'   `length(c_test)` columns, on the original response scale.
#' @export
sample_eiv_test_y <- function(x_test_raw,
                              c_test,
                              fit_obj,
                              draw_ids,
                              n_per_draw = 1L) {
  samples_u <- fit_obj$mcmc$samples_u
  samples_tau <- fit_obj$mcmc$samples_tau
  samples_logtheta <- fit_obj$mcmc$samples_logtheta
  samples_sigma2 <- fit_obj$mcmc$samples_sigma2
  
  x_train <- fit_obj$data$x
  y_train <- fit_obj$data$y
  y_center <- fit_obj$data$y_center
  y_scale <- fit_obj$data$y_scale
  x_center <- fit_obj$data$x_center
  x_scale <- fit_obj$data$x_scale
  
  x_test <- as.numeric((x_test_raw - x_center) / x_scale)
  
  n_test <- length(c_test)
  n_draw <- length(draw_ids) * n_per_draw
  
  out <- matrix(NA_real_, nrow = n_draw, ncol = n_test)
  
  row_id <- 0L
  
  for (s in draw_ids) {
    for (rr in seq_len(n_per_draw)) {
      row_id <- row_id + 1L
      
      tau_s <- samples_tau[s, ]
      
      lower <- c(-Inf, tau_s)[c_test]
      upper <- c(tau_s, Inf)[c_test]
      
      u_star <- rtruncnorm_vec(
        mean = rep(0, n_test),
        sd = 1,
        lower = lower,
        upper = upper
      )
      
      pred <- gp_predict_draw(
        x_train = x_train,
        u_train = samples_u[s, ],
        y_train = y_train,
        x_star = x_test,
        u_star = u_star,
        logtheta = samples_logtheta[s, ],
        sigma2_eps = samples_sigma2[s],
        noisy = TRUE
      )
      
      y_std <- pred$mean + sqrt(pred$var) * rnorm(n_test)
      out[row_id, ] <- y_center + y_scale * y_std
    }
  }
  
  out
}
```

