---
title: "Development numerical experiments — draft results"
output:
  html_document:
    toc: true
params:
  study: "study1"
  execute: false
  repository_dir: ""
  core_budget: 12
---

# Purpose

This document produces a small, reproducible draft of the numerical results.
These results support inspecting the analysis and writing the paper; they are
not final publication evidence. Both studies follow the same wrapper below.
The current algorithms come from the repository's `codes/` modules, which also
build eivmixgp. The historical eivGP launcher is not used here.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, cache = FALSE)
document <- knitr::current_input(dir = TRUE)
repository_dir <- params$repository_dir
if (!nzchar(repository_dir)) {
  repository_dir <- if (is.character(document) && length(document) == 1L && nzchar(document)) {
    file.path(dirname(document), "..")
  } else if (file.exists("codes/simulation_helpers.R")) getwd() else ".."
}
repository_dir <- normalizePath(repository_dir, mustWork = TRUE)
source(file.path(repository_dir, "codes", "00_diagnostics.R"), local = knitr::knit_global())
source(file.path(repository_dir, "codes", "simulation_helpers.R"), local = knitr::knit_global())
```

# Explicit experiment wrapper

The default profile uses three replications in two selected cells, four
chains per dataset, 500 warmup iterations, and 1,250 retained draws per chain.
Each chain runs 1,750 total iterations. Dataset and chain workers share the
explicit core budget. Existing diagnostic thresholds remain visible;
the short run is not required to pass them. The profile generates its own
frozen development data, then fits and aggregates. Subsequent runs verify
those data. See `DEVELOPMENT.md` for settings and the paper-writing cycle.

```{r configure}
run_simulation <- function(study, execute = FALSE, core_budget = 12L) {
  study <- match.arg(study, c("study1", "study2"))
  constructor <- if (study == "study1") study1_simulation_config else study2_simulation_config
  runner <- if (study == "study1") run_study1_simulation else run_study2_simulation
  config <- constructor(
    mode = "development", core_budget = core_budget,
    code_dir = file.path(repository_dir, "codes")
  )
  print(config[c("study", "mode", "parallel", "cells", "mcmc", "evaluation", "output_root", "data_root")])
  if (!execute) {
    config$mode <- "dry_run"
    config$stages <- character(0)
  }
  runner(config)
}
```

# Run and inspect

```{r run-development}
result <- run_simulation(params$study, params$execute, params$core_budget)
if (isTRUE(params$execute)) {
  print(result$run_dir)
  print(result$gates)
  print(list.files(file.path(result$run_dir, "combined")))
}
```

Review missing-method records and diagnostics alongside every draft table.
Use the full publication design only once the scientific settings and plotting
choices are settled. Real-data results remain a separate, deferred step.

```{r session}
sessionInfo()
```
