---
title: "Study I numerical experiments"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  run_mode: "dry_run"
  workers: 8
  stages: ""
  data_root: ""
  output_root: ""
---

# Purpose

This is the sole top-to-bottom reproduction document for the frozen Study I
publication design. Its computational implementation lives in `eivGP`; this
file contains configuration, one package call, and presentation only.

The committed default is a read-only dry run. To reproduce the paper, change
`run_mode` above to `publication` and render this document in a clean R
session. A resumable fit may set `stages` to `fit,aggregate`; the existing
frozen data are checksum-verified before fitting. Smoke and publication fits
automatically run the exact published-competitor and paired-design validators;
publication mode requires every competitor validation to succeed.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE,
  error = FALSE, cache = FALSE
)
document_file <- knitr::current_input(dir = TRUE)
repository_dir <- normalizePath(
  file.path(dirname(document_file), ".."), winslash = "/", mustWork = TRUE
)
package_dir <- file.path(repository_dir, "eivGP")
if (!requireNamespace("pkgload", quietly = TRUE)) {
  stop("Install pkgload to run the development version of eivGP.")
}
pkgload::load_all(package_dir, quiet = TRUE)
```

# Configuration

Only run mode, worker count, resume stages, and artifact locations are exposed.
The design cells, seeds, sample sizes, MCMC controls, estimands, method
eligibility, and diagnostic gates are frozen inside the package constructor.

```{r configure}
RUN_MODE <- as.character(params$run_mode)
WORKERS <- as.integer(params$workers)
STAGES <- if (nzchar(params$stages)) {
  trimws(strsplit(params$stages, ",", fixed = TRUE)[[1L]])
} else {
  NULL
}
DATA_ROOT <- if (nzchar(params$data_root)) {
  normalizePath(params$data_root, winslash = "/", mustWork = FALSE)
} else {
  file.path(repository_dir, "artifacts", "data-synthetic", "publication-v2", "study1")
}
OUTPUT_ROOT <- if (nzchar(params$output_root)) {
  normalizePath(params$output_root, winslash = "/", mustWork = FALSE)
} else {
  file.path(repository_dir, "artifacts", "simulation-runs", "study1")
}
CODE_DIR <- file.path(repository_dir, "script", "R-scripts")

STUDY1_CONFIG <- eivGP::study1_simulation_config(
  mode = RUN_MODE,
  code_dir = CODE_DIR,
  workers = WORKERS,
  data_root = DATA_ROOT,
  output_root = OUTPUT_ROOT,
  stages = STAGES
)
STUDY1_CONFIG
```

# Run

```{r run-study1}
STUDY1_RUN <- eivGP::run_study1_simulation(STUDY1_CONFIG)
```

# Auditable outputs

Dry-run output contains the resolved design, dependency preflight, task plan,
and status of any existing frozen data. Executed runs additionally expose the
fingerprinted run directory, verified input manifest, diagnostic gates, raw
results, paired comparisons, result checksums, and automatic validation records
under `config/`.

```{r summarize-study1}
if (identical(STUDY1_CONFIG$mode, "dry_run")) {
  print(STUDY1_RUN$preflight)
  print(STUDY1_RUN$runtime_preflight)
  print(STUDY1_RUN$task_plan)
  print(STUDY1_RUN$data_status)
} else {
  print(STUDY1_RUN$run_dir)
  print(STUDY1_RUN$gates)
  combined_dir <- file.path(STUDY1_RUN$run_dir, "combined")
  print(list.files(combined_dir, full.names = TRUE))
}
```

# Session information

```{r session-information}
sessionInfo()
```
