---
title: "Study II numerical experiments"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  run_mode: "dry_run"
  workers: 8
  stages: ""
  data_root: ""
  output_root: ""
---

# Purpose

This is the sole top-to-bottom reproduction document for the frozen Study II
publication design. Its computational implementation lives in `eivGP`; this
file contains configuration, one package call, and presentation only.

The committed default is a read-only dry run. To reproduce the paper, change
`run_mode` above to `publication` and render this document in a clean R
session. A resumable fit may set `stages` to `fit,aggregate`; every selected
manifest entry, design field, generator version, and checksum is verified
before dispatch. Smoke and publication fits automatically run the exact
published-competitor and paired-design validators; publication mode requires
every competitor validation to succeed.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE,
  error = FALSE, cache = FALSE
)
document_file <- knitr::current_input(dir = TRUE)
repository_dir <- normalizePath(
  file.path(dirname(document_file), ".."), winslash = "/", mustWork = TRUE
)
package_dir <- file.path(repository_dir, "eivGP")
if (!requireNamespace("pkgload", quietly = TRUE)) {
  stop("Install pkgload to run the development version of eivGP.")
}
pkgload::load_all(package_dir, quiet = TRUE)
```

# Configuration

Only run mode, worker count, resume stages, and artifact locations are exposed.
The proxy-dimension cells, calibration curve, controls, seeds, MCMC settings,
estimands, method eligibility, and diagnostic gates are frozen inside the
package constructor.

```{r configure}
RUN_MODE <- as.character(params$run_mode)
WORKERS <- as.integer(params$workers)
STAGES <- if (nzchar(params$stages)) {
  trimws(strsplit(params$stages, ",", fixed = TRUE)[[1L]])
} else {
  NULL
}
DATA_ROOT <- if (nzchar(params$data_root)) {
  normalizePath(params$data_root, winslash = "/", mustWork = FALSE)
} else {
  file.path(repository_dir, "artifacts", "data-synthetic", "publication-v2", "study2")
}
OUTPUT_ROOT <- if (nzchar(params$output_root)) {
  normalizePath(params$output_root, winslash = "/", mustWork = FALSE)
} else {
  file.path(repository_dir, "artifacts", "simulation-runs", "study2")
}
CODE_DIR <- file.path(repository_dir, "script", "R-scripts")

STUDY2_CONFIG <- eivGP::study2_simulation_config(
  mode = RUN_MODE,
  code_dir = CODE_DIR,
  workers = WORKERS,
  data_root = DATA_ROOT,
  output_root = OUTPUT_ROOT,
  stages = STAGES
)
STUDY2_CONFIG
```

# Run

```{r run-study2}
STUDY2_RUN <- eivGP::run_study2_simulation(STUDY2_CONFIG)
```

# Auditable outputs

Dry-run output contains the resolved design, dependency preflight, task plan,
and status of any existing frozen data. Executed runs additionally expose the
fingerprinted run directory, verified input manifest, diagnostic gates, raw
results, paired comparisons, prespecified contrasts, result checksums, and
automatic validation records under `config/`.

```{r summarize-study2}
if (identical(STUDY2_CONFIG$mode, "dry_run")) {
  print(STUDY2_RUN$preflight)
  print(STUDY2_RUN$runtime_preflight)
  print(STUDY2_RUN$task_plan)
  print(STUDY2_RUN$data_status)
} else {
  print(STUDY2_RUN$run_dir)
  print(STUDY2_RUN$gates)
  combined_dir <- file.path(STUDY2_RUN$run_dir, "combined")
  print(list.files(combined_dir, full.names = TRUE))
}
```

# Session information

```{r session-information}
sessionInfo()
```
