---
title: "Study II numerical experiments"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  run_mode: "dry_run"
  workers: 8
  data_root: ""
  output_root: ""
---

# Purpose

This is the sole top-to-bottom fitting and reporting document for the frozen
Study II publication design. Generate the frozen datasets first with
`study2_synthetic_data.Rmd`. The canonical model modules in `codes/` also build
`eivGP`; the study constructors and orchestration remain repository code.

The committed default is a read-only dry run. To reproduce the paper, change
`run_mode` above to `publication` and render this document in a clean R
session. Publication execution always verifies frozen data, fits the required
methods, then aggregates results. Published-competitor and paired-design
validators record their status; missing optional competitors are recorded and
diagnostic warnings do not stop a requested fitting budget.

The repository configuration records fixed iteration budgets and retains every
post-warmup draw. Continuation is explicit. Inspect the resolved configuration
and diagnostic flags before interpreting outputs; the package release does not
itself rerun or validate this study's scientific results.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE,
  error = FALSE, cache = FALSE
)
if (!requireNamespace("eivGP", quietly = TRUE)) {
  stop("Install eivGP before rendering this document; see README.md.")
}
library(eivGP)
document_path <- knitr::current_input(dir = TRUE)
repository_dir <- normalizePath(file.path(dirname(document_path), ".."),
                                mustWork = TRUE)
source(file.path(repository_dir, "codes", "simulation_helpers.R"),
       local = knitr::knit_global())
```

# Configuration

Only run mode, worker count, and artifact locations are exposed. The
proxy-dimension cells, calibration curve, controls, seeds, MCMC settings,
estimands, method eligibility, and diagnostic thresholds belong to the
repository constructor in `codes/simulation_helpers.R`.

```{r configure}
DOCUMENT_PARAMS <- list(
  run_mode = "dry_run", workers = 8L,
  data_root = "", output_root = ""
)
if (exists("params", inherits = FALSE) && is.list(params)) {
  DOCUMENT_PARAMS <- utils::modifyList(DOCUMENT_PARAMS, params)
}
RUN_MODE <- as.character(DOCUMENT_PARAMS$run_mode)
WORKERS <- as.integer(DOCUMENT_PARAMS$workers)
ARTIFACT_ROOT <- Sys.getenv("EIVGP_ARTIFACT_ROOT", unset = "")
if (!nzchar(ARTIFACT_ROOT)) {
  ARTIFACT_ROOT <- normalizePath(file.path(getwd(), "reproduction"),
                                 winslash = "/", mustWork = FALSE)
}
ARTIFACT_ROOT <- normalizePath(
  ARTIFACT_ROOT, winslash = "/", mustWork = FALSE
)
DATA_ROOT <- if (nzchar(DOCUMENT_PARAMS$data_root)) {
  normalizePath(DOCUMENT_PARAMS$data_root, winslash = "/", mustWork = FALSE)
} else {
  file.path(
    ARTIFACT_ROOT, "data",
    if (identical(RUN_MODE, "publication")) "synthetic" else "smoke",
    "study2"
  )
}
OUTPUT_ROOT <- if (nzchar(DOCUMENT_PARAMS$output_root)) {
  normalizePath(DOCUMENT_PARAMS$output_root, winslash = "/", mustWork = FALSE)
} else {
  file.path(ARTIFACT_ROOT, "results", "study2")
}
run_study2_publication <- function(mode, workers, data_root, output_root) {
  mode <- match.arg(mode, c("dry_run", "smoke", "publication"))
  config <- study2_simulation_config(
    code_dir = file.path(repository_dir, "codes"),
    mode = mode,
    workers = workers,
    data_root = data_root,
    output_root = output_root,
    # study2_synthetic_data.Rmd owns data creation.  This wrapper verifies
    # the frozen inputs, fits prespecified methods, and aggregates outputs.
    stages = if (identical(mode, "dry_run")) NULL else c("fit", "aggregate")
  )
  run_study2_simulation(config)
}
```

# Run

```{r run-study2}
STUDY2_RUN <- run_study2_publication(
  mode = RUN_MODE,
  workers = WORKERS,
  data_root = DATA_ROOT,
  output_root = OUTPUT_ROOT
)
```

# Auditable outputs

Dry-run output contains the resolved design, dependency preflight, task plan,
and status of any existing frozen data. Executed runs additionally expose the
fingerprinted run directory, verified input manifest, diagnostic gates, raw
results, paired comparisons, prespecified contrasts, result checksums, and
automatic validation records under `config/`.

```{r summarize-study2}
if (identical(RUN_MODE, "dry_run")) {
  print(STUDY2_RUN$preflight)
  print(STUDY2_RUN$runtime_preflight)
  print(STUDY2_RUN$task_plan)
  print(STUDY2_RUN$data_status)
} else {
  print(STUDY2_RUN$run_dir)
  print(STUDY2_RUN$gates)
  combined_dir <- file.path(STUDY2_RUN$run_dir, "combined")
  print(list.files(combined_dir, full.names = TRUE))
}
```

# Session information

```{r session-information}
sessionInfo()
```
