---
title: "Study I frozen synthetic data"
output:
  html_document:
    toc: true
params:
  run_mode: "dry_run"
  workers: 8
  data_root: ""
---

# Purpose

This document generates the exact, deterministic synthetic datasets used by
the frozen Study I publication design. Run it in `publication` mode once before
`study1_numerical_experiment.Rmd`. Existing files are never overwritten; the
package verifies their manifest and checksums before any fit is allowed.
The committed default is a read-only dry run. `smoke` mode creates a separate,
explicitly non-reportable test dataset; `publication` creates the paper data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE,
                      error = FALSE, cache = FALSE)
if (!requireNamespace("eivGP", quietly = TRUE)) {
  stop("Install eivGP before rendering this document; see README.md.")
}
library(eivGP)
```

# Configuration

```{r configure}
DOCUMENT_PARAMS <- list(run_mode = "dry_run", workers = 8L, data_root = "")
if (exists("params", inherits = FALSE) && is.list(params)) {
  DOCUMENT_PARAMS <- utils::modifyList(DOCUMENT_PARAMS, params)
}
WORKERS <- as.integer(DOCUMENT_PARAMS$workers)
if (length(WORKERS) != 1L || is.na(WORKERS) || WORKERS < 1L) {
  stop("workers must be one positive integer.")
}
RUN_MODE <- match.arg(as.character(DOCUMENT_PARAMS$run_mode),
                      c("dry_run", "smoke", "publication"))
ARTIFACT_ROOT <- Sys.getenv("EIVGP_ARTIFACT_ROOT", unset = "")
if (!nzchar(ARTIFACT_ROOT)) {
  ARTIFACT_ROOT <- file.path(getwd(), "reproduction")
}
DATA_ROOT <- if (nzchar(DOCUMENT_PARAMS$data_root)) {
  normalizePath(DOCUMENT_PARAMS$data_root, winslash = "/", mustWork = FALSE)
} else {
  file.path(
    ARTIFACT_ROOT, "data",
    if (identical(RUN_MODE, "publication")) "synthetic" else "smoke",
    "study1"
  )
}

generate_study1_publication_data <- function(mode, workers, data_root,
                                             output_root) {
  config <- eivGP::study1_simulation_config(
    mode = mode,
    workers = workers,
    data_root = data_root,
    output_root = output_root,
    # The data-only stage does not create simulation results.  In dry-run mode
    # the package retains its read-only default stages.
    stages = if (mode %in% c("smoke", "publication")) "data" else NULL
  )
  eivGP::run_study1_simulation(config)
}
```

# Generate and verify data

```{r generate-data}
STUDY1_DATA_RUN <- generate_study1_publication_data(
  mode = RUN_MODE,
  workers = WORKERS,
  data_root = DATA_ROOT,
  output_root = file.path(ARTIFACT_ROOT, "results", "study1")
)
print(STUDY1_DATA_RUN$data_manifest)
```

# Session information

```{r session-information}
sessionInfo()
```
