---
title: "Study I numerical experiments"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  run_mode: "dry_run"
  workers: 8
  data_root: ""
  output_root: ""
---

# Purpose

This is the sole top-to-bottom fitting and reporting document for the frozen
Study I publication design. Generate the frozen datasets first with
`study1_synthetic_data.Rmd`. The model implementations, including MCMC, live
in `eivGP`; the wrapper below makes the executed stages explicit.

The committed default is a read-only dry run. To reproduce the paper, change
`run_mode` above to `publication` and render this document in a clean R
session. Publication execution always verifies frozen data, fits the required
methods, then aggregates the results. Smoke and publication fits automatically
run the exact published-competitor and paired-design validators; publication
mode requires every competitor validation to succeed.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE,
  error = FALSE, cache = FALSE
)
if (!requireNamespace("eivGP", quietly = TRUE)) {
  stop("Install eivGP before rendering this document; see README.md.")
}
library(eivGP)
```

# Configuration

Only run mode, worker count, and artifact locations are exposed. The design
cells, seeds, sample sizes, MCMC controls, estimands, method eligibility, and
diagnostic gates are frozen inside the package constructor.

```{r configure}
DOCUMENT_PARAMS <- list(
  run_mode = "dry_run", workers = 8L,
  data_root = "", output_root = ""
)
if (exists("params", inherits = FALSE) && is.list(params)) {
  DOCUMENT_PARAMS <- utils::modifyList(DOCUMENT_PARAMS, params)
}
RUN_MODE <- as.character(DOCUMENT_PARAMS$run_mode)
WORKERS <- as.integer(DOCUMENT_PARAMS$workers)
ARTIFACT_ROOT <- Sys.getenv("EIVGP_ARTIFACT_ROOT", unset = "")
if (!nzchar(ARTIFACT_ROOT)) {
  ARTIFACT_ROOT <- normalizePath(file.path(getwd(), "reproduction"),
                                 winslash = "/", mustWork = FALSE)
}
ARTIFACT_ROOT <- normalizePath(
  ARTIFACT_ROOT, winslash = "/", mustWork = FALSE
)
DATA_ROOT <- if (nzchar(DOCUMENT_PARAMS$data_root)) {
  normalizePath(DOCUMENT_PARAMS$data_root, winslash = "/", mustWork = FALSE)
} else {
  file.path(
    ARTIFACT_ROOT, "data",
    if (identical(RUN_MODE, "publication")) "synthetic" else "smoke",
    "study1"
  )
}
OUTPUT_ROOT <- if (nzchar(DOCUMENT_PARAMS$output_root)) {
  normalizePath(DOCUMENT_PARAMS$output_root, winslash = "/", mustWork = FALSE)
} else {
  file.path(ARTIFACT_ROOT, "results", "study1")
}
run_study1_publication <- function(mode, workers, data_root, output_root) {
  mode <- match.arg(mode, c("dry_run", "smoke", "publication"))
  config <- eivGP::study1_simulation_config(
    mode = mode,
    workers = workers,
    data_root = data_root,
    output_root = output_root,
    # Data are created and frozen by study1_synthetic_data.Rmd.  This
    # document only verifies them, fits all prespecified methods, and
    # aggregates the resulting tables and figures.
    stages = if (identical(mode, "dry_run")) NULL else c("fit", "aggregate")
  )
  eivGP::run_study1_simulation(config)
}
```

# Run

```{r run-study1}
STUDY1_RUN <- run_study1_publication(
  mode = RUN_MODE,
  workers = WORKERS,
  data_root = DATA_ROOT,
  output_root = OUTPUT_ROOT
)
```

# Auditable outputs

Dry-run output contains the resolved design, dependency preflight, task plan,
and status of any existing frozen data. Executed runs additionally expose the
fingerprinted run directory, verified input manifest, diagnostic gates, raw
results, paired comparisons, result checksums, and automatic validation records
under `config/`.

```{r summarize-study1}
if (identical(RUN_MODE, "dry_run")) {
  print(STUDY1_RUN$preflight)
  print(STUDY1_RUN$runtime_preflight)
  print(STUDY1_RUN$task_plan)
  print(STUDY1_RUN$data_status)
} else {
  print(STUDY1_RUN$run_dir)
  print(STUDY1_RUN$gates)
  combined_dir <- file.path(STUDY1_RUN$run_dir, "combined")
  print(list.files(combined_dir, full.names = TRUE))
}
```

# Session information

```{r session-information}
sessionInfo()
```
