---
title: "Study II numerical experiments"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  run_mode: "dry_run"
  workers: 8
  data_root: ""
  output_root: ""
---

# Purpose

This is the sole top-to-bottom fitting and reporting document for the frozen
Study II publication design. Generate the frozen datasets first with
`study2_synthetic_data.Rmd`. The model implementations, including MCMC, live
in `eivGP`; the wrapper below makes the executed stages explicit.

The committed default is a read-only dry run. To reproduce the paper, change
`run_mode` above to `publication` and render this document in a clean R
session. Publication execution always verifies frozen data, fits the required
methods, then aggregates results. Smoke and publication fits automatically run
the exact published-competitor and paired-design validators; publication mode
requires every competitor validation to succeed.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE,
  error = FALSE, cache = FALSE
)
if (!requireNamespace("eivGP", quietly = TRUE)) {
  stop("Install eivGP before rendering this document; see README.md.")
}
library(eivGP)
```

# Configuration

Only run mode, worker count, and artifact locations are exposed. The
proxy-dimension cells, calibration curve, controls, seeds, MCMC settings,
estimands, method eligibility, and diagnostic gates are frozen inside the
package constructor.

```{r configure}
DOCUMENT_PARAMS <- list(
  run_mode = "dry_run", workers = 8L,
  data_root = "", output_root = ""
)
if (exists("params", inherits = FALSE) && is.list(params)) {
  DOCUMENT_PARAMS <- utils::modifyList(DOCUMENT_PARAMS, params)
}
RUN_MODE <- as.character(DOCUMENT_PARAMS$run_mode)
WORKERS <- as.integer(DOCUMENT_PARAMS$workers)
ARTIFACT_ROOT <- Sys.getenv("EIVGP_ARTIFACT_ROOT", unset = "")
if (!nzchar(ARTIFACT_ROOT)) {
  ARTIFACT_ROOT <- normalizePath(file.path(getwd(), "reproduction"),
                                 winslash = "/", mustWork = FALSE)
}
ARTIFACT_ROOT <- normalizePath(
  ARTIFACT_ROOT, winslash = "/", mustWork = FALSE
)
DATA_ROOT <- if (nzchar(DOCUMENT_PARAMS$data_root)) {
  normalizePath(DOCUMENT_PARAMS$data_root, winslash = "/", mustWork = FALSE)
} else {
  file.path(
    ARTIFACT_ROOT, "data",
    if (identical(RUN_MODE, "publication")) "synthetic" else "smoke",
    "study2"
  )
}
OUTPUT_ROOT <- if (nzchar(DOCUMENT_PARAMS$output_root)) {
  normalizePath(DOCUMENT_PARAMS$output_root, winslash = "/", mustWork = FALSE)
} else {
  file.path(ARTIFACT_ROOT, "results", "study2")
}
run_study2_publication <- function(mode, workers, data_root, output_root) {
  mode <- match.arg(mode, c("dry_run", "smoke", "publication"))
  config <- eivGP::study2_simulation_config(
    mode = mode,
    workers = workers,
    data_root = data_root,
    output_root = output_root,
    # study2_synthetic_data.Rmd owns data creation.  This wrapper verifies
    # the frozen inputs, fits prespecified methods, and aggregates outputs.
    stages = if (identical(mode, "dry_run")) NULL else c("fit", "aggregate")
  )
  eivGP::run_study2_simulation(config)
}
```

# Run

```{r run-study2}
STUDY2_RUN <- run_study2_publication(
  mode = RUN_MODE,
  workers = WORKERS,
  data_root = DATA_ROOT,
  output_root = OUTPUT_ROOT
)
```

# Auditable outputs

Dry-run output contains the resolved design, dependency preflight, task plan,
and status of any existing frozen data. Executed runs additionally expose the
fingerprinted run directory, verified input manifest, diagnostic gates, raw
results, paired comparisons, prespecified contrasts, result checksums, and
automatic validation records under `config/`.

```{r summarize-study2}
if (identical(RUN_MODE, "dry_run")) {
  print(STUDY2_RUN$preflight)
  print(STUDY2_RUN$runtime_preflight)
  print(STUDY2_RUN$task_plan)
  print(STUDY2_RUN$data_status)
} else {
  print(STUDY2_RUN$run_dir)
  print(STUDY2_RUN$gates)
  combined_dir <- file.path(STUDY2_RUN$run_dir, "combined")
  print(list.files(combined_dir, full.names = TRUE))
}
```

# Session information

```{r session-information}
sessionInfo()
```
