---
title: "Study I frozen synthetic data"
output:
  html_document:
    toc: true
params:
  run_mode: "dry_run"
  workers: 8
  data_root: ""
  visualizations_only: false
  figure_root: ""
---

# Purpose

This document generates the exact, deterministic synthetic datasets used by
the frozen Study I publication design. Run it in `publication` mode once before
`study1_numerical_experiment.Rmd`. Existing files are never overwritten; the
repository helper verifies their manifest and checksums before fitting.
The committed default is a read-only dry run. `smoke` mode creates a separate,
explicitly non-reportable test dataset; `publication` creates the paper data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE,
                      error = FALSE, cache = FALSE)
VISUALIZATIONS_ONLY <- exists("params") && isTRUE(params$visualizations_only)
document_file <- knitr::current_input(dir = TRUE)
REPOSITORY_DIR <- if (is.character(document_file) && length(document_file) == 1L && nzchar(document_file)) {
  normalizePath(file.path(dirname(document_file), ".."), mustWork = TRUE)
} else if (dir.exists("experiments")) normalizePath(".") else normalizePath("..")
if (!VISUALIZATIONS_ONLY && !requireNamespace("eivGP", quietly = TRUE)) {
  stop("Install eivGP before rendering this document; see README.md.")
}
if (!VISUALIZATIONS_ONLY) library(eivGP)
if (!VISUALIZATIONS_ONLY) {
  source(file.path(REPOSITORY_DIR, "codes", "00_diagnostics.R"),
         local = knitr::knit_global())
  source(file.path(REPOSITORY_DIR, "codes", "simulation_helpers.R"),
         local = knitr::knit_global())
}
```

# Configuration

```{r configure}
DOCUMENT_PARAMS <- list(run_mode = "dry_run", workers = 8L, data_root = "")
if (exists("params", inherits = FALSE) && is.list(params)) {
  DOCUMENT_PARAMS <- utils::modifyList(DOCUMENT_PARAMS, params)
}
WORKERS <- as.integer(DOCUMENT_PARAMS$workers)
if (length(WORKERS) != 1L || is.na(WORKERS) || WORKERS < 1L) {
  stop("workers must be one positive integer.")
}
RUN_MODE <- match.arg(as.character(DOCUMENT_PARAMS$run_mode),
                      c("dry_run", "smoke", "publication"))
ARTIFACT_ROOT <- Sys.getenv("EIVGP_ARTIFACT_ROOT", unset = "")
if (!nzchar(ARTIFACT_ROOT)) {
  ARTIFACT_ROOT <- file.path(REPOSITORY_DIR, "reproduction")
}
DATA_ROOT <- if (nzchar(DOCUMENT_PARAMS$data_root)) {
  normalizePath(DOCUMENT_PARAMS$data_root, winslash = "/", mustWork = FALSE)
} else {
  file.path(
    ARTIFACT_ROOT, "data",
    if (identical(RUN_MODE, "smoke")) "smoke" else "synthetic",
    "study1"
  )
}

generate_study1_publication_data <- function(mode, workers, data_root,
                                             output_root) {
  config <- study1_simulation_config(
    code_dir = file.path(REPOSITORY_DIR, "codes"),
    mode = mode,
    workers = workers,
    data_root = data_root,
    output_root = output_root
  )
  if (mode %in% c("smoke", "publication")) config$stages <- "data"
  config <- validate_simulation_config(config)
  run_study1_simulation(config)
}
```

# Generate and verify data

```{r generate-data}
STUDY1_DATA_RUN <- if (VISUALIZATIONS_ONLY) NULL else generate_study1_publication_data(
  mode = RUN_MODE,
  workers = WORKERS,
  data_root = DATA_ROOT,
  output_root = file.path(ARTIFACT_ROOT, "results", "study1")
)
print(STUDY1_DATA_RUN$data_manifest)
```

# Session information

```{r session-information}
sessionInfo()
```

# Visualize the saved synthetic datasets

These are **data-design illustrations, not fitted GP estimates**. We use the
prespecified replication 1 from each available setting, never a visually
favourable replication. Only training observations are plotted. The inventory
reports the actual saved training and test sizes; older frozen files may have
larger test sets than the current defaults. No data are regenerated or truncated.

Set `visualizations_only = TRUE` when rendering to skip generation and the
legacy eivGP dependency. Set `data_root` to the directory holding this study's
setting folders. This section reads files and verifies their manifest checksums.
Set `figure_root` (or `EIVGP_FIGURE_ROOT`) to an output figures directory,
such as the Overleaf project's `figures` folder. Without an override, it uses
`reproduction/figures`. Figures and an inventory go beneath
`frozen-data/study1/` within that directory.

```{r frozen-data-selection, fig.width=10, fig.height=7}
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  stop("Install ggplot2 to draw the frozen-data figures.")
}
library(ggplot2)
VIS_REP <- 1L
VIS_CALIBRATION <- 10L
manifest_paths <- sort(list.files(DATA_ROOT, pattern = "^manifest\\.rds$",
                                 recursive = TRUE, full.names = TRUE))
frozen_examples <- list()
inventory_rows <- list()
for (manifest_path in manifest_paths) {
  manifest <- readRDS(manifest_path)
  if (!is.data.frame(manifest) ||
      !all(c("study", "rep", "file", "md5") %in% names(manifest))) {
    stop("Malformed manifest: ", manifest_path)
  }
  selected <- manifest[manifest$study == "study1" & manifest$rep == VIS_REP, , drop = FALSE]
  if (!nrow(selected)) next
  if (nrow(selected) != 1L) stop("Ambiguous replication in: ", manifest_path)
  data_path <- file.path(dirname(manifest_path), basename(selected$file))
  if (!file.exists(data_path) ||
      !identical(unname(tools::md5sum(data_path)), as.character(selected$md5))) {
    stop("Missing or checksum-mismatched frozen data: ", data_path)
  }
  frozen <- readRDS(data_path)
  stopifnot(identical(frozen$study, "study1"), frozen$rep_id == VIS_REP,
            is.list(frozen$data), is.list(frozen$calibration_sets))
  setting <- basename(dirname(manifest_path))
  if (setting %in% names(frozen_examples)) stop("Duplicate setting folder: ", setting)
  frozen_examples[[setting]] <- frozen
  inventory_rows[[setting]] <- data.frame(
    setting = setting, replication = frozen$rep_id,
    n_train = nrow(frozen$data$train),
    n_test = nrow(frozen$data$test),
    saved_calibration_sizes = paste(names(frozen$calibration_sets), collapse = ", "),
    data_seed = frozen$seeds$data, calibration_seed = frozen$seeds$calibration,
    file = basename(data_path), md5 = as.character(selected$md5)
  )
}
if (!length(frozen_examples)) {
  message("No frozen Study I replication-1 datasets found under ", DATA_ROOT,
          ". No data were generated and no plots were fabricated.")
} else {
  FIGURE_ROOT <- if (exists("params") && is.character(params$figure_root) &&
                    length(params$figure_root) == 1L && nzchar(params$figure_root)) {
    params$figure_root
  } else Sys.getenv("EIVGP_FIGURE_ROOT", unset = file.path(ARTIFACT_ROOT, "figures"))
  figure_dir <- file.path(FIGURE_ROOT, "frozen-data", "study1")
  dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
  inventory <- do.call(rbind, inventory_rows)
  rownames(inventory) <- NULL
  print(knitr::kable(inventory[, 1:7], caption = "Frozen datasets actually visualized"))
  write.csv(inventory, file.path(figure_dir, "visualized_dataset_inventory.csv"), row.names = FALSE)
}
save_frozen_plot <- function(plot, name, width = 10, height = 7) {
  print(plot)
  ggplot2::ggsave(file.path(figure_dir, paste0(name, ".png")), plot,
                 width = width, height = height, dpi = 160, bg = "white")
  ggplot2::ggsave(file.path(figure_dir, paste0(name, ".pdf")), plot,
                 width = width, height = height, bg = "white")
}
frozen_theme <- theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom")
```

## Category frequencies and observed responses

The dashed line represents equal category proportions, not a fitted model.
Finite samples need not have equal counts, even in a balanced population.

```{r study1-frozen-observed, fig.width=10, fig.height=7}
if (length(frozen_examples)) {
  observed <- do.call(rbind, lapply(names(frozen_examples), function(setting) {
    dat <- frozen_examples[[setting]]$data$train
    stopifnot(all(c("x", "u", "c", "y", "f") %in% names(dat)),
              all(is.finite(as.matrix(dat))))
    data.frame(dat, setting = setting)
  }))
  counts <- do.call(rbind, lapply(names(frozen_examples), function(setting) {
    frozen <- frozen_examples[[setting]]
    data.frame(setting = setting, category = factor(seq_len(frozen$design$m)),
               proportion = tabulate(frozen$data$train$c, nbins = frozen$design$m) /
                 nrow(frozen$data$train),
               balanced_reference = 1 / frozen$design$m)
  }))
  p_counts <- ggplot(counts, aes(category, proportion)) +
    geom_col(fill = "#287C8E", width = 0.7) +
    geom_hline(aes(yintercept = balanced_reference), linetype = "dashed",
               colour = "#AA5B32") +
    facet_wrap(~setting, ncol = 2) +
    labs(title = "Study I: ordinal category frequencies",
         subtitle = "Frozen replication 1; training observations",
         x = "Observed ordinal category C", y = "Proportion",
         caption = "Dashed line: equal category probabilities. These are design illustrations.") +
    frozen_theme
  save_frozen_plot(p_counts, "category_frequencies")
  p_response <- ggplot(observed, aes(x, y, colour = factor(c))) +
    geom_point(size = 1.8, alpha = 0.8) +
    facet_wrap(~setting, ncol = 2) +
    scale_colour_viridis_d(option = "D", end = 0.9) +
    labs(title = "Study I: responses versus observed input",
         subtitle = "Frozen replication 1; identical colour scale across settings",
         x = "Observed input x", y = "Observed response y", colour = "Category C") +
    frozen_theme
  save_frozen_plot(p_response, "response_by_category")
}
```

## Latent response and nested calibration

Colour shows the saved noise-free response `f`, not a GP fit. Black rings
identify the saved calibration observations. Only the requested grids
`0, 10, 50` already present in a dataset are shown; no new subsets are sampled.

```{r study1-frozen-calibration, fig.width=12, fig.height=9}
if (length(frozen_examples)) {
  calibration_rows <- list()
  for (setting in names(frozen_examples)) {
    frozen <- frozen_examples[[setting]]
    grids <- intersect(c("0", "10", "50"), names(frozen$calibration_sets))
    if (!length(grids)) {
      message("No requested saved calibration sizes for ", setting)
      next
    }
    for (grid in grids) {
      ids <- frozen$calibration_sets[[grid]]
      stopifnot(length(ids) == as.integer(grid), !anyDuplicated(ids),
                all(ids %in% seq_len(nrow(frozen$data$train))))
      calibration_rows[[paste(setting, grid)]] <- data.frame(
        frozen$data$train, setting = setting, calibration = as.integer(grid),
        calibrated = seq_len(nrow(frozen$data$train)) %in% ids)
    }
    for (i in seq_len(length(grids) - 1L)) {
      stopifnot(all(frozen$calibration_sets[[grids[i]]] %in%
                      frozen$calibration_sets[[grids[i + 1L]]]))
    }
  }
  if (length(calibration_rows)) {
    calibration_data <- do.call(rbind, calibration_rows)
    p_calibration <- ggplot(calibration_data, aes(x, u)) +
      geom_point(aes(colour = f), size = 1.6, alpha = 0.85) +
      geom_point(data = subset(calibration_data, calibrated), shape = 1,
                 colour = "black", size = 2.8, stroke = 0.65) +
      scale_colour_viridis_c(option = "C") +
      facet_grid(setting ~ calibration, labeller = label_both) +
      labs(title = "Study I: latent response and saved calibration sets",
           subtitle = "Colour: saved noise-free response; black rings: calibrated observations",
           x = "Observed input x", y = "Latent input u", colour = "True f(x,u)",
           caption = "Uncalibrated u is visible here only because these are synthetic data.") +
      frozen_theme
    save_frozen_plot(p_calibration, "latent_response_and_calibration", 12,
                    max(5, 2.2 * length(frozen_examples)))
  }
}
```
