---
title: "Prepared real-data checks"
output: html_document
---

These checks begin with the collaborator's cleaned inputs in `real-data/`.
They do not download, rebuild, or alter the cohort. Raw processing lives in
`ADNI_preprocess.Rmd` and `ocean_preprocess.Rmd`; its provenance must be reviewed
separately from these consistency checks. Missing data are reported explicitly.
Run `Rscript codes/run_real_data.R data`; no model is fitted.

```{r setup}
if (!exists("DATA_WORKFLOW_DIR")) {
  input <- knitr::current_input(dir = TRUE)
  if (length(input) != 1L || !nzchar(input)) stop("Use run_real_data.R data or render this notebook.")
  DATA_WORKFLOW_DIR <- dirname(normalizePath(input))
}
source(file.path(DATA_WORKFLOW_DIR, "real_data_paths.R"), local = TRUE)
paths <- application_paths(DATA_WORKFLOW_DIR)
```

## ADNI

```{r check-adni}
source(file.path(DATA_WORKFLOW_DIR, "ADNI_cv.R"), local = TRUE)
if (!file.exists(file.path(paths$adni_data, "toledo_adni_cohort_n495.csv"))) {
  cat("ADNI unavailable:", paths$adni_data, "\n")
} else {
  dat <- adni_read_cohort(paths$adni_data)
  folds <- adni_make_folds(dat)
  stopifnot(ncol(dat) == 29L,
    all(is.finite(as.matrix(dat[, c("y_centiloid", "x_age_years", "x_female", "x_apoe4_dose")]))),
    all(is.finite(dat$u_csf_abeta42[dat$R == 1L])),
    all(dat$c_ab42_ab40_3 %in% 1:3), all(dat$c_gfap_3 %in% 1:3))
  cat("PASS: frozen cohort; 495 participants; 133 observed / 362 missing CSF.\n")
  print(table(fold = folds$fold, observed_CSF = folds$R))
}
```

## Ocean

```{r check-ocean}
OCEAN_REALDATA_DIR <- DATA_WORKFLOW_DIR
source(file.path(DATA_WORKFLOW_DIR, "ocean_data_helpers.R"), local = TRUE)
if (!all(file.exists(file.path(paths$ocean_data, c("study1_data.csv", "meta.json"))))) {
  cat("Ocean unavailable:", paths$ocean_data, "\n")
} else {
  ocean <- load_ocean_prepared()
  allocation <- read_ocean_class_allocation()
  calibration <- make_ocean_nested_calibration(ocean$c_train, allocation)
  stopifnot(nrow(ocean$train) == 100L, nrow(ocean$test) == 300L)
  cat("PASS: ocean schema; 100 training / 300 test rows; nested calibration counts verified.\n")
}
```
